---
title: "02_ALT_status_full"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

Let's load the dataset and have a look at it.

```{r setup}
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(readxl, quietly = TRUE)
library(here)
library(randomForest, quietly = TRUE)
library(caret, quietly = TRUE)
library(pROC, quietly = TRUE)
library(treeshap, quietly = TRUE)
library(shapviz, quietly = TRUE)
library(patchwork, quietly = TRUE)
source(here('scripts', 'useful_plots.R'))
data <- read_xlsx(here('data', 'raw', 'PCAWG_variables_PCA_woK.xlsx'))
summary(data)
```

We have to rename columns, set categories as factors, deal with NA values, and handle typing.

```{r cleaning-data}
source(here('scripts', 'clean_columns.R'))
clean_data <- clean_pcawg_data(data)
summary(clean_data)
```

# Define the full dataset

We must account for the fact that we have many **NA values**, especially in the **TERT_FPKM column**. These rows must be excluded, since further analyses such as PCA or RF require the absence of missing values. Therefore this extended model will use a subset of samples with complete **telomere biology annotations**.

```{r model-data-full}
model_data_full_rows <- complete.cases(clean_data)
model_data_full <- clean_data[model_data_full_rows,-c(1,2)]
model_data_full_meta <- clean_data[model_data_full_rows,c("donor_id", "cancer_type")]
summary(model_data_full)
```

Now we don't have NA values anymore, but we should be aware of the fact that we are dealing with a **much smaller dataset** (1943 before, 817 now) and some cancer types are underrepresented or totally absent.

# Define the larger dataset

We also keep a dataframe with all observation, thus **dropping the TERT_FPKM column**.

```{r model-data-large}
model_data_large <- clean_data %>% select(-TERT_FPKM) %>% drop_na()
```

# Correlation matrix

```{r corr-matrix}
corr_data <- model_data_full %>% 
  select(-alt_status)

cor_mat <- corr_data %>%
  cor(method = "spearman", use = "pairwise.complete.obs")
corrplot::corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.col = "black",
  tl.cex = 0.55,
  tl.srt = 45
)
```

# PCA with all features

```{r PCA_full}
# model_data_full has alt_status as first column, features after
res_full <- pca_plot(model_data_full, coloring = model_data_full$alt_status,
                     title = "PCA of all features")
res_full$plot
summary(res_full$pca)
res_full$pca$rotation
```

## Biplot

```{r biplot}
feats_full <- colnames(model_data_full)[-1]
comp_full  <- ifelse(grepl("singleton_dist", feats_full),
                     sub("_singleton_dist", "", feats_full), feats_full)

bip_full <- pca_biplot(model_data_full, coloring = model_data_full$alt_status,
                       scale_arrows = 10, compnames = comp_full,
                       title = "PCA biplot of all features")
bip_full$plot
ggsave(here('outputs', 'PCA_biplot_all.pdf'), bip_full$plot, width = 7, height = 6)
```

# PCA without TERT (more patients)

```{r pca-no-tert}
# drop the id/cancer_type columns so alt_status is the first (label) column
large_pca <- model_data_large %>% select(-donor_id, -cancer_type)

res_large <- pca_plot(large_pca, coloring = large_pca$alt_status,
                      title = "PCA without TERT - more patients (1943)")
res_large$plot
summary(res_large$pca)
res_large$pca$rotation
```

## Biplot

```{r biplot-no-tert}
feats_large <- colnames(large_pca)[-1]
comp_large  <- ifelse(grepl("singleton_dist", feats_large),
                      sub("_singleton_dist", "", feats_large), feats_large)

bip_large <- pca_biplot(large_pca, coloring = large_pca$alt_status,
                        scale_arrows = 18, compnames = comp_large,
                        title = "PCA without TERT - more patients (1943)")
bip_large$plot
ggsave(here('outputs', 'PCA_biplot_no_TERT.pdf'), bip_large$plot, width = 7, height = 6)
```

# PCA Mesenchymal groups

Here we repeat the PCA on the full set of telomere features, but label the samples by **mesenchymal grouping** (`cancer_group`: Mesenchymal vs Non-Mesenchymal origin) instead of ALT status. We use the processed dataset which carries this annotation.

```{r mesench-data}
mesench_raw <- read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))

# all telomere features (everything numeric except the metadata/label columns)
mesench_features <- c(
  "tf_primary_rate",
  "tf_blood_rate",
  "telomere_insertion_rate",
  "telomere_content_log2",
  "TERT_FPKM",
  "ATAGGG_singleton_dist",
  "CTAGGG_singleton_dist",
  "GTAGGG_singleton_dist",
  "TAAGGG_singleton_dist",
  "TCAGGG_singleton_dist",
  "TGAGGG_singleton_dist",
  "TTCGGG_singleton_dist",
  "TTGGGG_singleton_dist",
  "TTTGGG_singleton_dist"
)

mesench_data <- mesench_raw %>%
  mutate(cancer_group = factor(cancer_group))
```

## With TERT (complete telomere annotation)

Including TERT_FPKM requires dropping samples with missing expression, leaving a smaller subset.

```{r pca-mesench-tert}
# cancer_group as first (label) column, features after
mesench_full <- mesench_data %>%
  select(cancer_group, all_of(mesench_features)) %>%
  drop_na()

res_m <- pca_plot(mesench_full, coloring = mesench_full$cancer_group,
                  title = paste0("PCA of all features (with TERT) - mesenchymal grouping (n = ",
                                 nrow(mesench_full), ")"))
res_m$plot
summary(res_m$pca)
ggsave(here('outputs', 'PCA_mesenchymal_all.pdf'), res_m$plot, width = 7, height = 6)
```

### Biplot

```{r biplot-mesench-tert}
comp_m <- ifelse(grepl("singleton_dist", mesench_features),
                 sub("_singleton_dist", "", mesench_features), mesench_features)

bip_m <- pca_biplot(mesench_full, coloring = mesench_full$cancer_group,
                    scale_arrows = 10, compnames = comp_m,
                    title = "PCA biplot (with TERT) - mesenchymal grouping")
bip_m$plot
ggsave(here('outputs', 'PCA_biplot_mesenchymal_all.pdf'), bip_m$plot, width = 7, height = 6)
```

## Without TERT (more samples)

Dropping TERT_FPKM keeps many more samples.

```{r pca-mesench-no-tert}
mesench_features_noT <- setdiff(mesench_features, "TERT_FPKM")

mesench_large <- mesench_data %>%
  select(cancer_group, all_of(mesench_features_noT)) %>%
  drop_na()

res_m2 <- pca_plot(mesench_large, coloring = mesench_large$cancer_group,
                   title = paste0("PCA without TERT - mesenchymal grouping (n = ",
                                  nrow(mesench_large), ")"))
res_m2$plot
summary(res_m2$pca)
ggsave(here('outputs', 'PCA_mesenchymal_no_TERT.pdf'), res_m2$plot, width = 7, height = 6)
```

### Biplot

```{r biplot-mesench-no-tert}
comp_m2 <- ifelse(grepl("singleton_dist", mesench_features_noT),
                  sub("_singleton_dist", "", mesench_features_noT), mesench_features_noT)

bip_m2 <- pca_biplot(mesench_large, coloring = mesench_large$cancer_group,
                     scale_arrows = 18, compnames = comp_m2,
                     title = "PCA biplot without TERT - mesenchymal grouping")
bip_m2$plot
ggsave(here('outputs', 'PCA_biplot_mesenchymal_no_TERT.pdf'), bip_m2$plot, width = 7, height = 6)
```



# Split the dataset

```{r train-test}
set.seed(21)
train_index <- sample(1:nrow(model_data_full), size = 0.7*nrow(model_data_full))
train <- model_data_full[train_index,]
test <- model_data_full[-train_index,]
```

# First classifier: Logistic Regression (all features)

```{r logreg-full, warning=FALSE}
train$alt_status <- relevel(train$alt_status, ref = "ALT-low")
test$alt_status  <- factor(test$alt_status, levels = levels(train$alt_status))

logreg_model_full <- glm(alt_status~., data=train, family = 'binomial')

prob <- predict(logreg_model_full, newdata = test, type = 'response')
```

## Evaluate performance

```{r ROC-AUC-logreg-full}
library(pROC, quietly = TRUE)

roc_obj_full <- roc(
  response = test$alt_status,
  predictor = prob,
  levels = c("ALT-low", "ALT-high"),
  direction = "<"
)

plot(roc_obj_full, main = "ROC curve - Logistic regression (all features)", xlim = c(1, 0))

auc(roc_obj_full)
```

When the model is trained using the **full set of telomere-related variables** (TVR motifs, TERT expression, telomere fusion rate, telomere insertions, and telomere content), the ROC curve shows almost perfect separation between ALT-high and ALT-low tumors. The **AUC of 0.998** indicates **near-perfect discriminatory ability**, meaning the model can almost always rank ALT-high tumors above ALT-low ones. Compared with the TVR-only model, this represents a **dramatic improvement** in predictive performance.

```{r pred-prob-distribution, warning=FALSE}
prob_df <- data.frame(
  prob = prob,
  alt_status = test$alt_status
)

library(ggplot2)

ggplot(prob_df, aes(x = prob, fill = alt_status)) +
  geom_density(alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Predicted probability distributions",
    x = "Predicted probability of ALT-high",
    y = "Density"
  )
```

## Other metrics

```{r confusion-matrix-logreg-full}
pred_class <- ifelse(prob >= 0.5, "ALT-high", "ALT-low")
pred_class <- factor(pred_class, levels = c("ALT-low", "ALT-high"))

library(caret, quietly = TRUE)
confusionMatrix(pred_class, test$alt_status, positive = 'ALT-high')
```

The confusion matrix confirms the excellent performance of the logistic regression model when all telomere-related features are included. The model achieves an **overall accuracy of 98.4%** and a **very high balanced accuracy of 99.1%**, indicating strong performance for both classes despite the dataset imbalance. Notably, the model reaches **perfect sensitivity (1.00)** for this fold, correctly identifying all ALT-high tumors in the test set, while maintaining a **high specificity of 98.3%**, meaning that most ALT-low tumors are also classified correctly. The positive predictive value of 0.79 reflects the fact that **some ALT-low tumors are still predicted as ALT-high**, but the negative predictive value of 1.00 shows that tumors predicted as ALT-high are always correct in this sample.

# Second classifier: Random Forest

```{r RF-full}
library(randomForest, quietly = TRUE)
rf_model_full <- randomForest(alt_status~., data=train, importance=T)
rf_model_full
```

## Evaluate performance

```{r ROC-AUC-RF-full}
rf_prob_full <- predict(rf_model_full, test, type = "prob")[, "ALT-high"]

library(pROC, quietly = TRUE)

roc_rf_full <- roc(
  response = test$alt_status,
  predictor = rf_prob_full,
  levels = c("ALT-low", "ALT-high")
)

plot(
  roc_rf_full,
  main = "ROC curve - Random Forest (TVR motifs)"
)

auc(roc_rf_full)
```

**Almost perfect** (AUC = 0.9877). With this sample, our Random Forest model achieves almost perfect discriminatory ability.

## Other metrics

```{r confusion-matrix-RF}
rf_pred_full <- ifelse(rf_prob_full >= 0.5, "ALT-high", "ALT-low")
rf_pred_full <- factor(rf_pred_full, levels = c("ALT-low", "ALT-high"))

library(caret)

confusionMatrix(
  rf_pred_full,
  test$alt_status,
  positive = "ALT-high"
)
```

With this fold, the RF model achieved **slightly worse performances** with respect to Log. Reg., but still very good. We should be careful to not interpret this results as a general metric of performance, since we did not do CV yet.

```{r feature-importance}
varImpPlot(rf_model_full, main = 'Variable importance plot - Random Forest (all features)' )
```

# TreeSHAP interpretation: Random Forest

As in the tTF regression notebooks, we estimate out-of-fold TreeSHAP values so that each sample is explained by a model that did not train on that sample. For the binary classifier, the SHAP values explain the RF probability assigned to the **ALT-high** class.

```{r treeshap-rf-cv, message=FALSE, warning=FALSE}
clean_data_shap <- model_data_full %>%
  mutate(
    alt_status_rf = factor(
      if_else(alt_status == "ALT-high", 1, 0),
      levels = c(0, 1)
    )
  ) %>%
  select(-alt_status)

k <- 5
shap_list <- list()

set.seed(21)
fold_list <- createFolds(
  clean_data_shap$alt_status_rf,
  k = k,
  returnTrain = FALSE
)
folds <- integer(nrow(clean_data_shap))
for (i in seq_along(fold_list)) folds[fold_list[[i]]] <- i

for (i in 1:k) {
  fold_train      <- clean_data_shap[folds != i, ]
  fold_test       <- clean_data_shap[folds == i, ]
  fold_train_shap <- fold_train %>% select(-alt_status_rf)
  fold_test_shap  <- fold_test  %>% select(-alt_status_rf)
  
  fit_rf_fold <- randomForest(
    alt_status_rf ~ .,
    data = fold_train,
    ntree = 500,
    importance = TRUE
  )
  unified_fold <- randomForest.unify(fit_rf_fold, fold_train_shap)
  shap_fold    <- treeshap(unified_fold, fold_test_shap, verbose = FALSE)
  
  shap_list[[i]] <- list(
    shap     = shap_fold$shaps,
    X        = fold_test_shap,
    prob     = predict(fit_rf_fold, fold_test_shap, type = "prob")[, "1"],
    baseline = rep(
      mean(predict(fit_rf_fold, fold_train_shap, type = "prob")[, "1"]),
      nrow(fold_test_shap)
    ),
    row_id   = which(folds == i)
  )
}

shap_matrix <- bind_rows(lapply(shap_list, `[[`, "shap"))
X_matrix    <- bind_rows(lapply(shap_list, `[[`, "X"))
rf_oof_prob <- unlist(lapply(shap_list, `[[`, "prob"))
baseline    <- unlist(lapply(shap_list, `[[`, "baseline"))
row_ids     <- unlist(lapply(shap_list, `[[`, "row_id"))
ord         <- order(row_ids)

shap_matrix <- shap_matrix[ord, ]
X_matrix    <- X_matrix[ord, ]
rf_oof_prob <- rf_oof_prob[ord]
baseline    <- baseline[ord]

stopifnot(nrow(shap_matrix) == nrow(model_data_full))
stopifnot(nrow(X_matrix) == nrow(model_data_full))

shap_meta <- model_data_full_meta %>%
  mutate(
    alt_status = model_data_full$alt_status,
    cancer_type = fct_lump_n(factor(cancer_type), n = 8),
    rf_prob_alt_high = rf_oof_prob,
    baseline = baseline
  )

sv_oof_rf <- shapviz(
  object   = as.matrix(shap_matrix),
  X        = X_matrix,
  baseline = mean(baseline)
)

rf_shap_importance <- shap_matrix %>%
  summarise(across(everything(), ~ mean(abs(.x)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "mean_abs_shap") %>%
  arrange(desc(mean_abs_shap))

rf_shap_importance

max(abs(rowSums(shap_matrix) + baseline - rf_oof_prob))
```

```{r treeshap-rf-plots, message=FALSE, warning=FALSE}
dep_by_alt <- function(v) {
  tibble(
    feature_value = X_matrix[[v]],
    shap_value    = shap_matrix[[v]],
    alt_status    = shap_meta$alt_status
  ) %>%
    ggplot(aes(feature_value, shap_value, color = alt_status)) +
    geom_point(alpha = 0.7, size = 1.5) +
    labs(x = v, y = paste("SHAP value for", v), color = "ALT status") +
    theme_bw()
}

dep_by_cancer <- function(v) {
  tibble(
    feature_value = X_matrix[[v]],
    shap_value    = shap_matrix[[v]],
    cancer_type   = shap_meta$cancer_type
  ) %>%
    ggplot(aes(feature_value, shap_value, color = cancer_type)) +
    geom_point(alpha = 0.7, size = 1.5) +
    labs(x = v, y = paste("SHAP value for", v), color = "Cancer type") +
    theme_bw()
}

sv_importance(sv_oof_rf, kind = "beeswarm", bee_adjust = 20, alpha = 0.8) +
  theme_bw()

dep_by_alt("telomere_content_log2")

dep_by_alt("telomere_insertion_rate")

dep_by_alt("tf_rate")

dep_by_alt("TERT_FPKM")

dep_by_cancer("telomere_content_log2")
```

```{r treeshap-rf-singletons, message=FALSE, warning=FALSE}
singleton_vars <- c(
  "ATAGGG_singleton_dist",
  "CTAGGG_singleton_dist",
  "GTAGGG_singleton_dist",
  "TAAGGG_singleton_dist",
  "TCAGGG_singleton_dist",
  "TGAGGG_singleton_dist",
  "TTCGGG_singleton_dist",
  "TTGGGG_singleton_dist",
  "TTTGGG_singleton_dist"
)

singleton_plots_alt <- purrr::map(
  singleton_vars,
  \(v) {
    dep_by_alt(v) +
      theme(
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 7)
      )
  }
)
wrap_plots(singleton_plots_alt, ncol = 3, guides = "collect") +
  plot_annotation(title = "***GGG singleton SHAP dependence")

singleton_plots_cancer <- purrr::map(
  singleton_vars,
  \(v) {
    dep_by_cancer(v) +
      theme(
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 7)
      )
  }
)
wrap_plots(singleton_plots_cancer, ncol = 3, guides = "collect") +
  plot_annotation(title = "***GGG singleton SHAP dependence by cancer type")
```

# Cross Validation of the performances

To assess the robustness of these two predictive models, we conducted **repeated stratified cross-validation**.

```{r metric-helper, message=FALSE, warning=FALSE}
get_fold_metrics <- function(prob, truth, positive_class = "ALT-high", threshold = 0.5) {
  
  negative_class <- setdiff(levels(truth), positive_class)
  
  pred <- ifelse(prob > threshold, positive_class, negative_class) %>%
    factor(levels = levels(truth))
  
  cm <- confusionMatrix(
    data = pred,
    reference = truth,
    positive = positive_class
  )
  
  roc_obj <- roc(
    response = truth,
    predictor = prob,
    levels = c(negative_class, positive_class),
    quiet = TRUE
  )
  
  tibble(
    auc = as.numeric(auc(roc_obj)),
    sensitivity = unname(cm$byClass["Sensitivity"]),
    specificity = unname(cm$byClass["Specificity"]),
    precision = unname(cm$byClass["Pos Pred Value"]),
    recall = unname(cm$byClass["Recall"]),
    f1 = unname(cm$byClass["F1"]),
    balanced_accuracy = unname(cm$byClass["Balanced Accuracy"])
  )
}
```

## Logistic Regression - CV

```{r logreg-CV, message=FALSE, warning=FALSE}
library(pROC)
library(caret)
library(dplyr)

set.seed(21)

cv_data <- model_data_full
k <- 5
n_repeats <- 5

logreg_results <- tibble()

for (r in 1:n_repeats) {
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    logreg_cv_model <- glm(
      alt_status ~ .,
      data = train_cv,
      family = "binomial"
    )
    
    logreg_cv_prob <- predict(logreg_cv_model, test_cv, type = "response")
    
    fold_metrics <- get_fold_metrics(
      prob = logreg_cv_prob,
      truth = test_cv$alt_status,
      positive_class = "ALT-high"
    ) %>%
      mutate(
        rep = r,
        fold = fold,
        model = "Logistic regression"
      )
    
    logreg_results <- bind_rows(logreg_results, fold_metrics)
  }
}

logreg_results
```

```{r logreg-summary, message=FALSE, warning=FALSE}
logreg_summary <- logreg_results %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    )
  )

logreg_summary
```

## Random Forest - CV

```{r rf-CV-metrics, message=FALSE, warning=FALSE}
library(randomForest)
library(pROC)
library(caret)
library(dplyr)

set.seed(21)

cv_data <- model_data_full
k <- 5
n_repeats <- 5

rf_results <- tibble()

for (r in 1:n_repeats) {
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    rf_cv_model <- randomForest(
      alt_status ~ .,
      data = train_cv,
      ntree = 500
    )
    
    rf_cv_prob <- predict(rf_cv_model, test_cv, type = "prob")[, "ALT-high"]
    
    fold_metrics <- get_fold_metrics(
      prob = rf_cv_prob,
      truth = test_cv$alt_status,
      positive_class = "ALT-high"
    ) %>%
      mutate(
        rep = r,
        fold = fold,
        model = "Random forest"
      )
    
    rf_results <- bind_rows(rf_results, fold_metrics)
  }
}

rf_results
```

```{r rf-summary, message=FALSE, warning=FALSE}
rf_summary <- rf_results %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    )
  )

rf_summary
```

```{r compare-models, message=FALSE, warning=FALSE}
all_results <- bind_rows(logreg_results, rf_results)

model_comparison <- all_results %>%
  group_by(model) %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

model_comparison
```
