---
title: "02_ALT_status_full"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

Let's load the dataset and have a look at it.

```{r setup}
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(readxl, quietly = TRUE)
library(here)
library(randomForest, quietly = TRUE)
library(caret, quietly = TRUE)
library(pROC, quietly = TRUE)
library(treeshap, quietly = TRUE)
library(shapviz, quietly = TRUE)
library(patchwork, quietly = TRUE)
source(here('scripts', 'useful_plots.R'))
data <- read_xlsx(here('data', 'raw', 'PCAWG_variables_PCA_woK.xlsx'))
summary(data)
```

We have to rename columns, set categories as factors, deal with NA values, and handle typing.

```{r cleaning-data}
source(here('scripts', 'clean_columns.R'))
clean_data <- clean_pcawg_data(data)
summary(clean_data)
```

# Define the full dataset

We must account for the fact that we have many **NA values**, especially in the **TERT_FPKM column**. These rows must be excluded, since further analyses such as PCA or RF require the absence of missing values. Therefore this extended model will use a subset of samples with complete **telomere biology annotations**.

```{r model-data-full}
model_data_full_rows <- complete.cases(clean_data)
model_data_full <- clean_data[model_data_full_rows,-c(1,2)]
model_data_full_meta <- clean_data[model_data_full_rows,c('donor_id', 'cancer_type')]
summary(model_data_full)
```

Now we don't have NA values anymore, but we should be aware of the fact that we are dealing with a **much smaller dataset** (1943 before, 817 now) and some cancer types are underrepresented or totally absent.

# Define the larger dataset

We also keep a dataframe with all observation, thus **dropping the TERT_FPKM column**.

```{r model-data-large}
model_data_large <- clean_data %>% select(-TERT_FPKM) %>% drop_na()
```

# Correlation matrix

```{r corr-matrix}
corr_data <- model_data_full %>% 
  select(-alt_status)

cor_mat <- corr_data %>%
  cor(method = 'spearman', use = 'pairwise.complete.obs')
corrplot::corrplot(
  cor_mat,
  method = 'color',
  type = 'upper',
  order = 'hclust',
  tl.col = 'black',
  tl.cex = 0.55,
  tl.srt = 45
)
```

# PCA with all features

```{r PCA_full}
# model_data_full has alt_status as first column, features after
res_full <- pca_plot(model_data_full, coloring = model_data_full$alt_status,
                     title = 'PCA of all features')
res_full$plot
summary(res_full$pca)
res_full$pca$rotation
```

## Biplot

```{r biplot}
feats_full <- colnames(model_data_full)[-1]
comp_full  <- ifelse(grepl('singleton_dist', feats_full),
                     sub('_singleton_dist', '', feats_full), feats_full)

bip_full <- pca_biplot(model_data_full, coloring = model_data_full$alt_status,
                       scale_arrows = 10, compnames = comp_full,
                       title = 'PCA biplot of all features')
bip_full$plot
ggsave(here('outputs', 'PCA_biplot_all.pdf'), bip_full$plot, width = 7, height = 6)
```

# PCA without TERT (more patients)

```{r pca-no-tert}
# drop the id/cancer_type columns so alt_status is the first (label) column
large_pca <- model_data_large %>% select(-donor_id, -cancer_type)

res_large <- pca_plot(large_pca, coloring = large_pca$alt_status,
                      title = 'PCA without TERT - more patients (1943)')
res_large$plot
summary(res_large$pca)
res_large$pca$rotation
```

## Biplot

```{r biplot-no-tert}
feats_large <- colnames(large_pca)[-1]
comp_large  <- ifelse(grepl('singleton_dist', feats_large),
                      sub('_singleton_dist', '', feats_large), feats_large)

bip_large <- pca_biplot(large_pca, coloring = large_pca$alt_status,
                        scale_arrows = 18, compnames = comp_large,
                        title = 'PCA without TERT - more patients (1943)')
bip_large$plot
ggsave(here('outputs', 'PCA_biplot_no_TERT.pdf'), bip_large$plot, width = 7, height = 6)
```

# PCA Mesenchymal groups

Here we repeat the PCA on the full set of telomere features, but label the samples by **mesenchymal grouping** (`cancer_group`: Mesenchymal vs Non-Mesenchymal origin) instead of ALT status. We use the processed dataset which carries this annotation.

```{r mesench-data}
mesench_raw <- read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))

# all telomere features (everything numeric except the metadata/label columns)
mesench_features <- c(
  'tf_primary_rate',
  'tf_blood_rate',
  'telomere_insertion_rate',
  'telomere_content_log2',
  'TERT_FPKM',
  'ATAGGG_singleton_dist',
  'CTAGGG_singleton_dist',
  'GTAGGG_singleton_dist',
  'TAAGGG_singleton_dist',
  'TCAGGG_singleton_dist',
  'TGAGGG_singleton_dist',
  'TTCGGG_singleton_dist',
  'TTGGGG_singleton_dist',
  'TTTGGG_singleton_dist'
)

mesench_data <- mesench_raw %>%
  mutate(cancer_group = factor(cancer_group))
```

## With TERT (complete telomere annotation)

Including TERT_FPKM requires dropping samples with missing expression, leaving a smaller subset.

```{r pca-mesench-tert}
# cancer_group as first (label) column, features after
mesench_full <- mesench_data %>%
  select(cancer_group, all_of(mesench_features)) %>%
  drop_na()

res_m <- pca_plot(mesench_full, coloring = mesench_full$cancer_group,
                  title = paste0('PCA of all features (with TERT) - mesenchymal grouping (n = ',
                                 nrow(mesench_full), ')'))
res_m$plot
summary(res_m$pca)
ggsave(here('outputs', 'PCA_mesenchymal_all.pdf'), res_m$plot, width = 7, height = 6)
```

### Biplot

```{r biplot-mesench-tert}
comp_m <- ifelse(grepl('singleton_dist', mesench_features),
                 sub('_singleton_dist', '', mesench_features), mesench_features)

bip_m <- pca_biplot(mesench_full, coloring = mesench_full$cancer_group,
                    scale_arrows = 10, compnames = comp_m,
                    title = 'PCA biplot (with TERT) - mesenchymal grouping')
bip_m$plot
ggsave(here('outputs', 'PCA_biplot_mesenchymal_all.pdf'), bip_m$plot, width = 7, height = 6)
```

## Without TERT (more samples)

Dropping TERT_FPKM keeps many more samples.

```{r pca-mesench-no-tert}
mesench_features_noT <- setdiff(mesench_features, 'TERT_FPKM')

mesench_large <- mesench_data %>%
  select(cancer_group, all_of(mesench_features_noT)) %>%
  drop_na()

res_m2 <- pca_plot(mesench_large, coloring = mesench_large$cancer_group,
                   title = paste0('PCA without TERT - mesenchymal grouping (n = ',
                                  nrow(mesench_large), ')'))
res_m2$plot
summary(res_m2$pca)
ggsave(here('outputs', 'PCA_mesenchymal_no_TERT.pdf'), res_m2$plot, width = 7, height = 6)
```

### Biplot

```{r biplot-mesench-no-tert}
comp_m2 <- ifelse(grepl('singleton_dist', mesench_features_noT),
                  sub('_singleton_dist', '', mesench_features_noT), mesench_features_noT)

bip_m2 <- pca_biplot(mesench_large, coloring = mesench_large$cancer_group,
                     scale_arrows = 18, compnames = comp_m2,
                     title = 'PCA biplot without TERT - mesenchymal grouping')
bip_m2$plot
ggsave(here('outputs', 'PCA_biplot_mesenchymal_no_TERT.pdf'), bip_m2$plot, width = 7, height = 6)
```



# Model evaluation

Logistic regression and random forest are evaluated **only by repeated
cross-validation** (below); the cross-model head-to-head lives in
`20_classification_comparison`. A single 70/30 split is deliberately not used: with
only 49 ALT-high cases it gives unstable, optimistically biased estimates that are not
a fair basis for comparing models.

# TreeSHAP interpretation: Random Forest

As in the tTF regression notebooks, we estimate out-of-fold TreeSHAP values so that each sample is explained by a model that did not train on that sample. For the binary classifier, the SHAP values explain the RF probability assigned to the **ALT-high** class.

```{r treeshap-rf-cv, message=FALSE, warning=FALSE}
clean_data_shap <- model_data_full %>%
  mutate(
    alt_status_rf = factor(
      if_else(alt_status == 'ALT-high', 1, 0),
      levels = c(0, 1)
    )
  ) %>%
  select(-alt_status)

k <- 5
shap_list <- list()

set.seed(21)
fold_list <- createFolds(
  clean_data_shap$alt_status_rf,
  k = k,
  returnTrain = FALSE
)
folds <- integer(nrow(clean_data_shap))
for (i in seq_along(fold_list)) folds[fold_list[[i]]] <- i

for (i in 1:k) {
  fold_train      <- clean_data_shap[folds != i, ]
  fold_test       <- clean_data_shap[folds == i, ]
  fold_train_shap <- fold_train %>% select(-alt_status_rf)
  fold_test_shap  <- fold_test  %>% select(-alt_status_rf)
  
  fit_rf_fold <- randomForest(
    alt_status_rf ~ .,
    data = fold_train,
    ntree = 500,
    importance = TRUE
  )
  unified_fold <- randomForest.unify(fit_rf_fold, fold_train_shap)
  shap_fold    <- treeshap(unified_fold, fold_test_shap, verbose = FALSE)
  
  shap_list[[i]] <- list(
    shap     = shap_fold$shaps,
    X        = fold_test_shap,
    prob     = predict(fit_rf_fold, fold_test_shap, type = 'prob')[, '1'],
    baseline = rep(
      mean(predict(fit_rf_fold, fold_train_shap, type = 'prob')[, '1']),
      nrow(fold_test_shap)
    ),
    row_id   = which(folds == i)
  )
}

shap_matrix <- bind_rows(lapply(shap_list, `[[`, 'shap'))
X_matrix    <- bind_rows(lapply(shap_list, `[[`, 'X'))
rf_oof_prob <- unlist(lapply(shap_list, `[[`, 'prob'))
baseline    <- unlist(lapply(shap_list, `[[`, 'baseline'))
row_ids     <- unlist(lapply(shap_list, `[[`, 'row_id'))
ord         <- order(row_ids)

shap_matrix <- shap_matrix[ord, ]
X_matrix    <- X_matrix[ord, ]
rf_oof_prob <- rf_oof_prob[ord]
baseline    <- baseline[ord]

stopifnot(nrow(shap_matrix) == nrow(model_data_full))
stopifnot(nrow(X_matrix) == nrow(model_data_full))

shap_meta <- model_data_full_meta %>%
  mutate(
    alt_status = model_data_full$alt_status,
    cancer_type = fct_lump_n(factor(cancer_type), n = 8),
    rf_prob_alt_high = rf_oof_prob,
    baseline = baseline
  )

sv_oof_rf <- shapviz(
  object   = as.matrix(shap_matrix),
  X        = X_matrix,
  baseline = mean(baseline)
)

rf_shap_importance <- shap_matrix %>%
  summarise(across(everything(), ~ mean(abs(.x)))) %>%
  pivot_longer(everything(), names_to = 'feature', values_to = 'mean_abs_shap') %>%
  arrange(desc(mean_abs_shap))

rf_shap_importance

max(abs(rowSums(shap_matrix) + baseline - rf_oof_prob))
```

```{r treeshap-rf-plots, message=FALSE, warning=FALSE}
dep_by_alt <- function(v) {
  tibble(
    feature_value = X_matrix[[v]],
    shap_value    = shap_matrix[[v]],
    alt_status    = shap_meta$alt_status
  ) %>%
    ggplot(aes(feature_value, shap_value, color = alt_status)) +
    geom_point(alpha = 0.7, size = 1.5) +
    labs(x = v, y = paste('SHAP value for', v), color = 'ALT status') +
    theme_bw()
}

dep_by_cancer <- function(v) {
  tibble(
    feature_value = X_matrix[[v]],
    shap_value    = shap_matrix[[v]],
    cancer_type   = shap_meta$cancer_type
  ) %>%
    ggplot(aes(feature_value, shap_value, color = cancer_type)) +
    geom_point(alpha = 0.7, size = 1.5) +
    labs(x = v, y = paste('SHAP value for', v), color = 'Cancer type') +
    theme_bw()
}

sv_importance(sv_oof_rf, kind = 'beeswarm', bee_adjust = 20, alpha = 0.8) +
  theme_bw()

dep_by_alt('telomere_content_log2')

dep_by_alt('telomere_insertion_rate')

dep_by_alt('tf_rate')

dep_by_alt('TERT_FPKM')

dep_by_cancer('telomere_content_log2')
```

```{r treeshap-rf-singletons, message=FALSE, warning=FALSE}
singleton_vars <- c(
  'ATAGGG_singleton_dist',
  'CTAGGG_singleton_dist',
  'GTAGGG_singleton_dist',
  'TAAGGG_singleton_dist',
  'TCAGGG_singleton_dist',
  'TGAGGG_singleton_dist',
  'TTCGGG_singleton_dist',
  'TTGGGG_singleton_dist',
  'TTTGGG_singleton_dist'
)

singleton_plots_alt <- purrr::map(
  singleton_vars,
  \(v) {
    dep_by_alt(v) +
      theme(
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 7)
      )
  }
)
wrap_plots(singleton_plots_alt, ncol = 3, guides = 'collect') +
  plot_annotation(title = '***GGG singleton SHAP dependence')

singleton_plots_cancer <- purrr::map(
  singleton_vars,
  \(v) {
    dep_by_cancer(v) +
      theme(
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 7)
      )
  }
)
wrap_plots(singleton_plots_cancer, ncol = 3, guides = 'collect') +
  plot_annotation(title = '***GGG singleton SHAP dependence by cancer type')
```

# Cross Validation of the performances

To assess the robustness of these two predictive models, we conducted **repeated stratified cross-validation**.

```{r metric-helper, message=FALSE, warning=FALSE}
get_fold_metrics <- function(prob, truth, positive_class = 'ALT-high', threshold = 0.5) {
  
  negative_class <- setdiff(levels(truth), positive_class)
  
  pred <- ifelse(prob > threshold, positive_class, negative_class) %>%
    factor(levels = levels(truth))
  
  cm <- confusionMatrix(
    data = pred,
    reference = truth,
    positive = positive_class
  )
  
  roc_obj <- roc(
    response = truth,
    predictor = prob,
    levels = c(negative_class, positive_class),
    quiet = TRUE
  )
  
  tibble(
    auc = as.numeric(auc(roc_obj)),
    sensitivity = unname(cm$byClass['Sensitivity']),
    specificity = unname(cm$byClass['Specificity']),
    precision = unname(cm$byClass['Pos Pred Value']),
    recall = unname(cm$byClass['Recall']),
    f1 = unname(cm$byClass['F1']),
    balanced_accuracy = unname(cm$byClass['Balanced Accuracy'])
  )
}
```

## Logistic Regression - CV

```{r logreg-CV, message=FALSE, warning=FALSE}
library(pROC)
library(caret)
library(dplyr)

set.seed(21)

cv_data <- model_data_full
k <- 5
n_repeats <- 5

logreg_results <- tibble()

for (r in 1:n_repeats) {
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == 'ALT-low')
  idx_high <- which(cv_data$alt_status == 'ALT-high')
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    logreg_cv_model <- glm(
      alt_status ~ .,
      data = train_cv,
      family = 'binomial'
    )
    
    logreg_cv_prob <- predict(logreg_cv_model, test_cv, type = 'response')
    
    fold_metrics <- get_fold_metrics(
      prob = logreg_cv_prob,
      truth = test_cv$alt_status,
      positive_class = 'ALT-high'
    ) %>%
      mutate(
        rep = r,
        fold = fold,
        model = 'Logistic regression'
      )
    
    logreg_results <- bind_rows(logreg_results, fold_metrics)
  }
}

logreg_results
```

```{r logreg-summary, message=FALSE, warning=FALSE}
logreg_summary <- logreg_results %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = '{.col}_{.fn}'
    )
  )

logreg_summary
```

## Random Forest - CV

```{r rf-CV-metrics, message=FALSE, warning=FALSE}
library(randomForest)
library(pROC)
library(caret)
library(dplyr)

set.seed(21)

cv_data <- model_data_full
k <- 5
n_repeats <- 5

# Build all stratified folds up front (before any forest is fit) so the fold
# assignment is independent of randomForest's RNG, matching the logistic-regression CV.
idx_low  <- which(cv_data$alt_status == 'ALT-low')
idx_high <- which(cv_data$alt_status == 'ALT-high')
fold_mat <- matrix(0L, nrow(cv_data), n_repeats)
for (r in 1:n_repeats) {
  fold_id <- integer(nrow(cv_data))
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  fold_mat[, r] <- fold_id
}

rf_results <- tibble()

for (r in 1:n_repeats) {
  fold_id <- fold_mat[, r]

  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    rf_cv_model <- randomForest(
      alt_status ~ .,
      data = train_cv,
      ntree = 500
    )
    
    rf_cv_prob <- predict(rf_cv_model, test_cv, type = 'prob')[, 'ALT-high']
    
    fold_metrics <- get_fold_metrics(
      prob = rf_cv_prob,
      truth = test_cv$alt_status,
      positive_class = 'ALT-high'
    ) %>%
      mutate(
        rep = r,
        fold = fold,
        model = 'Random forest'
      )
    
    rf_results <- bind_rows(rf_results, fold_metrics)
  }
}

rf_results
```

```{r rf-summary, message=FALSE, warning=FALSE}
rf_summary <- rf_results %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = '{.col}_{.fn}'
    )
  )

rf_summary
```

```{r compare-models, message=FALSE, warning=FALSE}
all_results <- bind_rows(logreg_results, rf_results)

model_comparison <- all_results %>%
  group_by(model) %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = '{.col}_{.fn}'
    ),
    .groups = 'drop'
  )

model_comparison
```
