---
title: "05_ALT_status_full"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

Let's load the dataset and have a look at it.

```{r setup}
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(readxl, quietly = TRUE)
library(here)
data <- read_xlsx(here('data', 'raw', 'PCAWG_variables_PCA_woK.xlsx'))
summary(data)
```

We have to rename columns, set categories as factors, deal with NA values, and handle typing.

```{r cleaning-data}
source(here('scripts', 'clean_columns.R'))
clean_data <- clean_pcawg_data(data)
summary(clean_data)
```

# Define the full dataset

We must account for the fact that we have many **NA values**, especially in the **TERT_FPKM column**. These rows must be excluded, since further analyses such as PCA or RF require the absence of missing values. Therefore this extended model will use a subset of samples with complete **telomere biology annotations**.

```{r model-data-full}
model_data_full <- clean_data[complete.cases(clean_data),-c(1,2)]
summary(model_data_full)
```

Now we don't have NA values anymore, but we should be aware of the fact that we are dealing with a **much smaller dataset** (1943 before, 817 now) and some cancer types are underrepresented or totally absent.

# Define the larger dataset

We also keep a dataframe with all observation, thus **dropping the TERT_FPKM column**.

```{r model-data-large}
model_data_large <- clean_data %>% select(-TERT_FPKM) %>% drop_na()
```


# Correlation matrix

```{r corr-matrix}
corr_data <- model_data_full %>% 
  select(-alt_status)

cor_mat <- corr_data %>%
  cor(method = "spearman", use = "pairwise.complete.obs")
corrplot::corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.col = "black",
  tl.cex = 0.55,
  tl.srt = 45
)
```


# PCA with all features

```{r PCA_full}
X <- scale(model_data_full[, -1])
pca <- prcomp(X, center = TRUE, scale. = TRUE)
summary(pca)

library(ggplot2)

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  ALT = model_data_full$alt_status
)

ggplot(pca_df, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "PCA of all features",
    x = "PC1",
    y = "PC2"
  )

pca$rotation
```
## Biplot

```{r biplot}
# sample coordinates
scores <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  ALT = model_data_full$alt_status
)

# motif vectors
loadings <- data.frame(
  PC1 = pca$rotation[,1],
  PC2 = pca$rotation[,2],
  comp_names = c(rownames(pca$rotation)[1:4], sapply(strsplit(rownames(pca$rotation)[5:13], split = '_', fixed = TRUE), `[`, 1))
)

# scale the loadings (smaller than sample coordinates) to make arrows more visible
loadings$PC1 <- loadings$PC1 * 10
loadings$PC2 <- loadings$PC2 * 10

library(ggplot2)
library(ggrepel)

ggplot(scores, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.25,"cm")),
               inherit.aes = FALSE,
               color = "black") +
  geom_text_repel(data = loadings,
            aes(x = PC1, y = PC2, label = comp_names),
            inherit.aes = FALSE,
            size = 3.5,
            ) +
  theme_bw() +
  labs(
    title = "PCA biplot of all features",
    x = "PC1",
    y = "PC2"
  )
ggsave(here('outputs', 'PCA_biplot_all.pdf'), width = 7, height = 6)
```

# PCA without TERT (more patients)

```{r pca-no-tert}
X2 <- scale(model_data_large[, -c(1,2,3)])
pca2 <- prcomp(X2, center = TRUE, scale. = TRUE)
summary(pca2)

library(ggplot2)

pca_df2 <- data.frame(
  PC1 = pca2$x[,1],
  PC2 = pca2$x[,2],
  ALT = model_data_large$alt_status
)

ggplot(pca_df2, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "PCA without TERT - more patients (1943)",
    x = "PC1",
    y = "PC2"
  )

pca2$rotation
```

## Biplot

```{r biplot-no-tert}
# sample coordinates
scores2 <- data.frame(
  PC1 = pca2$x[,1],
  PC2 = pca2$x[,2],
  ALT = model_data_large$alt_status
)

# motif vectors
loadings2 <- data.frame(
  PC1 = pca2$rotation[,1],
  PC2 = pca2$rotation[,2],
  comp_names = c(rownames(pca2$rotation)[1:3], sapply(strsplit(rownames(pca2$rotation)[4:12], split = '_', fixed = TRUE), `[`, 1))
)

# scale the loadings (smaller than sample coordinates) to make arrows more visible
loadings2$PC1 <- loadings2$PC1 * 18
loadings2$PC2 <- loadings2$PC2 * 18

library(ggplot2)
library(ggrepel)

ggplot(scores2, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_segment(data = loadings2,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.25,"cm")),
               inherit.aes = FALSE,
               color = "black") +
  geom_text_repel(data = loadings2,
            aes(x = PC1, y = PC2, label = comp_names),
            inherit.aes = FALSE,
            size = 3.5,
            ) +
  theme_bw() +
  labs(
    title = "PCA without TERT - more patients (1943)",
    x = "PC1",
    y = "PC2"
  )

ggsave(here('outputs', 'PCA_biplot_no_TERT.pdf'), width = 7, height = 6)
```


# UMAP plot

```{r umap}
library(umap)

umap_df <- clean_data %>% 
  select(-donor_id) %>% 
  drop_na() %>% 
  mutate(cancer_type = factor(cancer_type))

top_n_types <- 12
top_types <- umap_df %>%
  count(cancer_type) %>%
  slice_max(n, n = top_n_types) %>%
  pull(cancer_type)

type_colors <- setNames(
  c(scales::hue_pal()(top_n_types), "grey70"),
  c(sort(top_types), "Other")
)

umap_vars <- umap_df %>%
  select(where(is.numeric))

umap_config <- umap.defaults
umap_config$n_neighbors  <- 30
umap_config$min_dist     <- 0.1
umap_config$random_state <- 21

umap_res <- umap(scale(umap_vars), config = umap_config)

top_types_chr <- as.character(top_types)

scores_umap <- as.data.frame(umap_res$layout) %>%
  rename(UMAP1 = V1, UMAP2 = V2) %>%
  mutate(
    cancer_type_plot = ifelse(
      as.character(umap_df$cancer_type) %in% top_types_chr,
      as.character(umap_df$cancer_type),
      "Other"
    ),
    cancer_type_plot = factor(cancer_type_plot,
                              levels = c(sort(top_types_chr), "Other")),
    ALT_status = umap_df$alt_status
  )

scores_umap_all <- scores_umap %>%
  mutate(cancer_type_all = factor(as.character(umap_df$cancer_type)))

library(RColorBrewer)

n_types <- length(levels(scores_umap_all$cancer_type_all))

type_colors_all <- setNames(
  pals::polychrome(n_types),
  levels(scores_umap_all$cancer_type_all)
)

type_colors <- setNames(
  c(pals::polychrome(top_n_types), "grey70"),
  c(sort(top_types_chr), "Other")
)

ggplot(scores_umap_all, aes(UMAP1, UMAP2, color = cancer_type_all)) +
  geom_point(aes(shape = ALT_status,
                 size  = ALT_status,
                 alpha = ALT_status)) +
  scale_shape_manual(values = c("ALT-low" = 16, "ALT-high" = 17)) +
  scale_size_manual(values  = c("ALT-low" = 1.5, "ALT-high" = 3)) +
  scale_alpha_manual(values = c("ALT-low" = 0.7, "ALT-high" = 1)) +
  scale_color_manual(values = type_colors_all) +
  theme_bw() +
  labs(
    title = "UMAP - PCAWG dataset, all cancer types",
    x = "UMAP1", y = "UMAP2",
    color = "Cancer type",
    shape = "ALT status", size = "ALT status", alpha = "ALT status"
  )

ggplot(scores_umap, aes(UMAP1, UMAP2)) +
  geom_point(
    data = filter(scores_umap, cancer_type_plot == "Other"),
    aes(color = cancer_type_plot, size = ALT_status),
    shape = 16, alpha = 0.3
  ) +
  geom_point(
    data = filter(scores_umap, cancer_type_plot != "Other"),
    aes(color = cancer_type_plot, shape = ALT_status, size = ALT_status),
    alpha = 0.7
  ) +
  scale_color_manual(values = type_colors) +
  scale_shape_manual(values = c("ALT-low" = 16, "ALT-high" = 17)) +
  scale_size_manual(values = c("ALT-low" = 1.5, "ALT-high" = 3)) +
  theme_bw() +
  labs(
    title = "UMAP - PCAWG dataset",
    x = "UMAP1", y = "UMAP2",
    color = "Cancer type", shape = "ALT status", size = "ALT status"
  )

ggplot(scores_umap, aes(UMAP1, UMAP2)) +
  geom_point(
    data = filter(scores_umap, ALT_status == "ALT-low"),
    color = "grey80", size = 1.2, alpha = 0.3, shape = 16
  ) +
  geom_point(
    data = filter(scores_umap, ALT_status == "ALT-high"),
    color = "#E41A1C", size = 2, alpha = 0.8, shape = 17
  ) +
  theme_bw() +
  labs(
    title = "UMAP - PCAWG dataset, ALT status highlighted",
    x = "UMAP1", y = "UMAP2"
  )

library(purrr)

params_grid <- expand.grid(
  n_neighbors = c(10, 15, 30, 50),
  min_dist    = c(0.05, 0.1, 0.5)
)

umap_plots <- pmap(params_grid, function(n_neighbors, min_dist) {
  
  cfg <- umap.defaults
  cfg$n_neighbors  <- n_neighbors
  cfg$min_dist     <- min_dist
  cfg$random_state <- 21
  
  res <- umap(scale(umap_vars), config = cfg)
  
  as.data.frame(res$layout) %>%
    rename(UMAP1 = V1, UMAP2 = V2) %>%
    mutate(
      ALT_status = umap_df$alt_status,
      n_neighbors = n_neighbors,
      min_dist    = min_dist
    )
})

all_umaps <- bind_rows(umap_plots)

ggplot(all_umaps, aes(UMAP1, UMAP2)) +
  geom_point(
    data = filter(all_umaps, ALT_status == "ALT-low"),
    color = "grey80", size = 0.8, alpha = 0.3, shape = 16
  ) +
  geom_point(
    data = filter(all_umaps, ALT_status == "ALT-high"),
    color = "#E41A1C", size = 1.5, alpha = 0.9, shape = 17
  ) +
  facet_grid(min_dist ~ n_neighbors,
             labeller = label_both) +
  theme_bw(base_size = 9) +
  labs(
    title = "UMAP parameter sensitivity - ALT status",
    x = "UMAP1", y = "UMAP2"
  )
```

# Split the dataset
```{r train-test}
set.seed(21)
train_index <- sample(1:nrow(model_data_full), size = 0.5*nrow(model_data_full))
train <- model_data_full[train_index,]
test <- model_data_full[-train_index,]
```

# First classifier: Logistic Regression (all features)
```{r logreg-full, warning=FALSE}
logreg_model_full <- glm(alt_status~., data=train, family = 'binomial')

prob <- predict(logreg_model_full, newdata = test, type = 'response')
```

## Evaluate performance

```{r ROC-AUC-logreg-full}
library(pROC, quietly = TRUE)

roc_obj_full <- roc(
  response = test$alt_status,
  predictor = prob,
  levels = c("ALT-low", "ALT-high"),
  direction = "<"
)

plot(roc_obj_full, main = "ROC curve - Logistic regression (all features)", xlim = c(1, 0))

auc(roc_obj_full)
```

When the model is trained using the **full set of telomere-related variables** (TVR motifs, TERT expression, telomere fusion rate, telomere insertions, and telomere content), the ROC curve shows almost perfect separation between ALT-high and ALT-low tumors. The **AUC of 0.998** indicates **near-perfect discriminatory ability**, meaning the model can almost always rank ALT-high tumors above ALT-low ones. Compared with the TVR-only model, this represents a **dramatic improvement** in predictive performance.

```{r pred-prob-distribution, warning=FALSE} 
prob_df <- data.frame(
  prob = prob,
  alt_status = test$alt_status
)

library(ggplot2)

ggplot(prob_df, aes(x = prob, fill = alt_status)) +
  geom_density(alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Predicted probability distributions",
    x = "Predicted probability of ALT-high",
    y = "Density"
  )
```


## Other metrics

```{r confusion-matrix-logreg-full}
pred_class <- ifelse(prob >= 0.5, "ALT-high", "ALT-low")
pred_class <- factor(pred_class, levels = c("ALT-low", "ALT-high"))

library(caret, quietly = TRUE)
confusionMatrix(pred_class, test$alt_status, positive = 'ALT-high')
```

The confusion matrix confirms the excellent performance of the logistic regression model when all telomere-related features are included. The model achieves an **overall accuracy of 98.4%** and a **very high balanced accuracy of 99.1%**, indicating strong performance for both classes despite the dataset imbalance. Notably, the model reaches **perfect sensitivity (1.00)** for this fold, correctly identifying all ALT-high tumors in the test set, while maintaining a **high specificity of 98.3%**, meaning that most ALT-low tumors are also classified correctly. The positive predictive value of 0.79 reflects the fact that **some ALT-low tumors are still predicted as ALT-high**, but the negative predictive value of 1.00 shows that tumors predicted as ALT-high are always correct in this sample.

# Second classifier: Random Forest

```{r RF-full}
library(randomForest, quietly = TRUE)
rf_model_full <- randomForest(alt_status~., data=train, importance=T)
rf_model_full
```

## Evaluate performance

```{r ROC-AUC-RF-full}
rf_prob_full <- predict(rf_model_full, test, type = "prob")[, "ALT-high"]

library(pROC, quietly = TRUE)

roc_rf_full <- roc(
  response = test$alt_status,
  predictor = rf_prob_full,
  levels = c("ALT-low", "ALT-high")
)

plot(
  roc_rf_full,
  main = "ROC curve - Random Forest (TVR motifs)"
)

auc(roc_rf_full)
```

**Almost perfect** (AUC = 0.9877). With this sample, our Random Forest model achieves almost perfect discriminatory ability.

## Other metrics

```{r confusion-matrix-RF}
rf_pred_full <- ifelse(rf_prob_full >= 0.5, "ALT-high", "ALT-low")
rf_pred_full <- factor(rf_pred_full, levels = c("ALT-low", "ALT-high"))

library(caret)

confusionMatrix(
  rf_pred_full,
  test$alt_status,
  positive = "ALT-high"
)
```

With this fold, the RF model achieved **slightly worse performances** with respect to Log. Reg., but still very good. We should be careful to not interpret this results as a general metric of performance, since we did not do CV yet.

```{r feature-importance}
varImpPlot(rf_model_full, main = 'Variable importance plot - Random Forest (all features)' )
```

# Cross Validation of the performances

To assess the robustness of these two predictive models, we conducted **repeated stratified cross-validation**. 

```{r metric-helper, message=FALSE, warning=FALSE}
get_fold_metrics <- function(prob, truth, positive_class = "ALT-high", threshold = 0.5) {
  
  negative_class <- setdiff(levels(truth), positive_class)
  
  pred <- ifelse(prob > threshold, positive_class, negative_class) %>%
    factor(levels = levels(truth))
  
  cm <- confusionMatrix(
    data = pred,
    reference = truth,
    positive = positive_class
  )
  
  roc_obj <- roc(
    response = truth,
    predictor = prob,
    levels = c(negative_class, positive_class),
    quiet = TRUE
  )
  
  tibble(
    auc = as.numeric(auc(roc_obj)),
    sensitivity = unname(cm$byClass["Sensitivity"]),
    specificity = unname(cm$byClass["Specificity"]),
    precision = unname(cm$byClass["Pos Pred Value"]),
    recall = unname(cm$byClass["Recall"]),
    f1 = unname(cm$byClass["F1"]),
    balanced_accuracy = unname(cm$byClass["Balanced Accuracy"])
  )
}
```

## Logistic Regression - CV

```{r logreg-CV, message=FALSE, warning=FALSE}
library(pROC)
library(caret)
library(dplyr)

set.seed(21)

cv_data <- model_data_full
k <- 5
n_repeats <- 5

logreg_results <- tibble()

for (r in 1:n_repeats) {
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    logreg_cv_model <- glm(
      alt_status ~ .,
      data = train_cv,
      family = "binomial"
    )
    
    logreg_cv_prob <- predict(logreg_cv_model, test_cv, type = "response")
    
    fold_metrics <- get_fold_metrics(
      prob = logreg_cv_prob,
      truth = test_cv$alt_status,
      positive_class = "ALT-high",
      threshold = 0.3
    ) %>%
      mutate(
        rep = r,
        fold = fold,
        model = "Logistic regression"
      )
    
    logreg_results <- bind_rows(logreg_results, fold_metrics)
  }
}

logreg_results
```

```{r logreg-summary, message=FALSE, warning=FALSE}
logreg_summary <- logreg_results %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    )
  )

logreg_summary
```

## Random Forest - CV

```{r rf-CV-metrics, message=FALSE, warning=FALSE}
library(randomForest)
library(pROC)
library(caret)
library(dplyr)

set.seed(21)

cv_data <- model_data_full
k <- 5
n_repeats <- 5

rf_results <- tibble()

for (r in 1:n_repeats) {
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    rf_cv_model <- randomForest(
      alt_status ~ .,
      data = train_cv,
      ntree = 500,
      mtry = 2
    )
    
    rf_cv_prob <- predict(rf_cv_model, test_cv, type = "prob")[, "ALT-high"]
    
    fold_metrics <- get_fold_metrics(
      prob = rf_cv_prob,
      truth = test_cv$alt_status,
      positive_class = "ALT-high",
      threshold = 0.3
    ) %>%
      mutate(
        rep = r,
        fold = fold,
        model = "Random forest"
      )
    
    rf_results <- bind_rows(rf_results, fold_metrics)
  }
}

rf_results
```

```{r rf-summary, message=FALSE, warning=FALSE}
rf_summary <- rf_results %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    )
  )

rf_summary
```

```{r compare-models, message=FALSE, warning=FALSE}
all_results <- bind_rows(logreg_results, rf_results)

model_comparison <- all_results %>%
  group_by(model) %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

model_comparison
```


