---
title: "06_ALT_status_tel_ins"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

```{r load&clean}
library(readxl)
source('../scripts/clean_columns.R')
data <- read_xlsx('../data/raw/PCAWG_variables_PCA_woK.xlsx')
clean_data <- clean_pcawg_data(data)
model_data_reduced <- clean_data[complete.cases(clean_data), -c(1,2,5)]  # we remove also telomere_insertion_rate
summary(model_data_reduced)
```

# PCA without tel. insertion

```{r PCA-red}
X <- scale(model_data_reduced[, -1])
pca <- prcomp(X, center = TRUE, scale. = TRUE)
summary(pca)

library(ggplot2)

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  ALT = model_data_reduced$alt_status
)

ggplot(pca_df, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "PCA without tel. insertion",
    x = "PC1",
    y = "PC2"
  )

pca$rotation
```

# Biplot

```{r biplot-red}
# sample coordinates
scores <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  ALT = model_data_reduced$alt_status
)

# motif vectors
loadings <- data.frame(
  PC1 = pca$rotation[,1],
  PC2 = pca$rotation[,2],
  comp_names = c(rownames(pca$rotation)[1:3], sapply(strsplit(rownames(pca$rotation)[4:12], split = '_', fixed = TRUE), `[`, 1))
)

# scale the loadings (smaller than sample coordinates) to make arrows more visible
loadings$PC1 <- loadings$PC1 * 10
loadings$PC2 <- loadings$PC2 * 10

library(ggplot2)
library(ggrepel)

p <- ggplot(scores, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.7, size = 2) +

  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.25,"cm")),
               inherit.aes = FALSE,
               color = "black") +

  geom_text_repel(data = loadings,
                  aes(PC1, PC2, label = comp_names),
                  inherit.aes = FALSE,
                  size = 3.5) +

  theme_classic() +
  coord_equal() +
  labs(
    title = "PCA biplot without tel. insertion",
    x = "PC1",
    y = "PC2"
  )

plot(p)

ggsave("../outputs/PCA_biplot_red_telins.pdf", plot = p, width = 7, height = 6)
```

Comparing the two PCA biplots, the most evident change **after removing telomere insertion rate** is the **reduction of the strong directional signal toward the lower-left quadrant**. In the full-feature PCA, the telomere insertion rate vector is long and points strongly in that direction, and many ALT-high tumors lie along this axis, indicating that this variable contributes substantially to the separation between ALT-high and ALT-low samples. Once this feature is removed, the PCA structure becomes more centered and the remaining variables distribute more evenly around the origin. The ALT-high samples still tend to occupy the lower-left region of the plot, but **the separation from ALT-low tumors becomes less pronounced**. The remaining vectors, particularly **TGAGGG, TTCGGG, GTAGGG, and TF rate—continue to point toward this region**, suggesting that **these variables now capture part of the variance previously dominated by telomere insertion rate**.

# Split the dataset
```{r train-test}
set.seed(21)

train_index <- sample(1:nrow(model_data_reduced), size=0.7*nrow(model_data_reduced))
train_red <- model_data_reduced[train_index,]
test_red <- model_data_reduced[-train_index,]

# Check distribution
table(train_red$alt_status, dnn = 'train')
table(test_red$alt_status, dnn = 'test')
```

# First classifier: Logistic Regression

```{r logreg-reduced}
logreg_model_reduced <- glm(
  alt_status ~ .,
  data = train_red,
  family = "binomial"
)

prob_red <- predict(
  logreg_model_reduced,
  test_red,
  type = "response"
)
```

## Evaluate performance

```{r ROC-AUC-red-logreg}
library(pROC, quietly = T)
roc_red <- roc(
  test_red$alt_status,
  prob_red,
  levels = c("ALT-low","ALT-high")
)

plot(roc_red, main = 'ROC curve - Log. Reg. (all features except tel. insertion)')

auc(roc_red)
```

```{r pred-prob-distribution, warning=FALSE} 
prob_df <- data.frame(
  prob = prob_red,
  alt_status = test_red$alt_status
)

library(ggplot2)

ggplot(prob_df, aes(x = prob, fill = alt_status)) +
  geom_density(alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Predicted probability distributions",
    x = "Predicted probability of ALT-high",
    y = "Density"
  )
```

### Other metrics

```{r confusion-matrix-logreg-red}
pred_class <- ifelse(prob_red >= 0.5, "ALT-high", "ALT-low")
pred_class <- factor(pred_class, levels = c("ALT-low", "ALT-high"))

library(caret, quietly = TRUE)
confusionMatrix(pred_class, test_red$alt_status, positive = 'ALT-high')
```

When telomere insertion rate is removed, the logistic regression model still achieves strong predictive performance. The model reaches an **AUC of 0.914**, indicating excellent discrimination between ALT-high and ALT-low tumors even without the strongest structural telomere feature. The confusion matrix shows a **high overall accuracy of 97.6%** and a **balanced accuracy of 0.83**, reflecting strong performance despite the class imbalance. The model maintains **very high specificity (0.996)**, meaning that ALT-low tumors are almost never misclassified, while **sensitivity is moderate (0.67)**, indicating that roughly two thirds of ALT-high tumors are correctly detected. The **positive predictive value of 0.91** further indicates that most tumors predicted as ALT-high are true positives.

# Second classifier: Random Forest

```{r rf-red}
library(randomForest, quietly = T)
rf_red <- randomForest(
  alt_status ~ .,
  data = train_red,
  ntree = 500,
  importance = TRUE
)

rf_pred_red <- predict(
  rf_red,
  test_red,
  type = "prob"
)[,"ALT-high"]
```

## Evaluate performance

```{r ROC-AUC-red-rf}
roc_rf_red <- roc(
  test_red$alt_status,
  rf_pred_red,
  levels = c("ALT-low","ALT-high")
)

plot(roc_rf_red, main = 'ROC curve - RF (all features except tel. insertion')

auc(roc_rf_red)
```

### Other metrics

```{r confusion-matrix-RF}
rf_pred_red <- ifelse(rf_pred_red >= 0.5, "ALT-high", "ALT-low")
rf_pred_red <- factor(rf_pred_red, levels = c("ALT-low", "ALT-high"))

library(caret)

confusionMatrix(
  rf_pred_red,
  test_red$alt_status,
  positive = "ALT-high"
)
```

When telomere insertion rate is removed, the Random Forest classifier still achieves strong discriminatory performance, with an **AUC of approximately 0.91**. This is very similar to the logistic regression reduced model and indicates that the remaining predictors—TF rate, telomere content, TERT expression, and TVR motifs—retain substantial information about ALT status. The confusion matrix shows a **high overall accuracy (97.6%)** and a **balanced accuracy of 0.80**, confirming that the model remains highly effective despite the removal of the strongest structural telomere feature.

The model exhibits **perfect specificity (1.00)**, meaning that ALT-low tumors are never misclassified as ALT-high in this split. However, **sensitivity is moderate (0.60)**, indicating that some ALT-high tumors remain difficult to detect when telomere insertion rate is excluded. The **positive predictive value of 1.00** shows that when the model predicts ALT-high, the prediction is always correct in this dataset. Overall, the model remains highly reliable but detects slightly fewer ALT-high cases compared with the full model.

```{r feature-importance}
varImpPlot(rf_red, main = 'Variable importance plot - Random Forest (all features except tel. insertion)' )
```

After removing telomere insertion, the PCA shows increased overlap between ALT-high and ALT-low samples, indicating that the **remaining features provide weaker but still meaningful separation**. This is reflected in the classification results, where the **AUC decreases slightly (≈0.91) but remains high**. The Random Forest variable importance plot further supports this interpretation: in the absence of telomere insertion rate, the **most influential predictors** become specific TVR motifs—particularly **TTTGGG**, **TGAGGG**, and **TTCGGG** followed by **TF rate** and **GTAGGG**. These features are also oriented toward the region of the PCA space where many ALT-high tumors cluster, indicating that they capture the remaining signal associated with ALT status. 

Together, the PCA and the classifier results therefore suggest that while telomere insertion rate is a dominant driver of discrimination, a combination of **TVR motifs and other telomere-related features still provides substantial predictive information** about the ALT status.


# Cross Validation of the performances

To assess the robustness of these two predictive models, we conducted **repeated stratified cross-validation**. 

## Logistic Regression - CV

```{r logreg-CV, message=FALSE, warning=FALSE}
library(pROC)

set.seed(21)

cv_data <- model_data_reduced
k <- 5
n_repeats <- 5

auc_results <- c()

for (r in 1:n_repeats) {
  # stratified fold assignment
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    logreg_cv_model <- glm(alt_status~., data = train_cv, family = 'binomial')
    
    logreg_cv_prob <- predict(logreg_cv_model, test_cv, type = 'response')
    
    roc_obj <- roc(
      response = test_cv$alt_status,
      predictor = logreg_cv_prob,
      levels = c("ALT-low", "ALT-high")
    )
    
    auc_results <- c(auc_results, as.numeric(auc(roc_obj)))
  }
}

mean(auc_results)
# sd(auc_results)
# auc_results
```

## Random Forest - CV

```{r RF-CV, message=FALSE}
library(randomForest)
library(pROC)

set.seed(21)

cv_data <- model_data_reduced
k <- 5
n_repeats <- 5

auc_results <- c()

for (r in 1:n_repeats) {
  # stratified fold assignment
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    rf_cv_model <- randomForest(
      alt_status ~ .,
      data = train_cv,
      ntree = 500,
      mtry = 2
    )
    
    rf_cv_prob <- predict(rf_cv_model, test_cv, type = "prob")[, "ALT-high"]
    
    roc_obj <- roc(
      response = test_cv$alt_status,
      predictor = rf_cv_prob,
      levels = c("ALT-low", "ALT-high")
    )
    
    auc_results <- c(auc_results, as.numeric(auc(roc_obj)))
  }
}

mean(auc_results)
# sd(auc_results)
# auc_results
```

These results confirm the **robustness of the AUC values previously recorded** for the two models, and agree with the fact that allowing non-linear relationships among features (i.e. using Random Forest) provide a better classification power.