---
title: "07_ALT_status_tf_rate"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

```{r load&clean}
library(readxl)
source('../scripts/clean_columns.R')
data <- read_xlsx('../data/raw/PCAWG_variables_PCA_woK.xlsx')
clean_data <- clean_pcawg_data(data)
model_data_reduced <- clean_data[complete.cases(clean_data), -c(1,2,4)]  # we remove also tf_rate
summary(model_data_reduced)
```

# PCA without TF rate

```{r PCA-red}
X <- scale(model_data_reduced[, -1])
pca <- prcomp(X, center = TRUE, scale. = TRUE)
summary(pca)

library(ggplot2)

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  ALT = model_data_reduced$alt_status
)

ggplot(pca_df, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "PCA without TF rate",
    x = "PC1",
    y = "PC2"
  )
```

# Biplot

```{r biplot-red}
# sample coordinates
scores <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  ALT = model_data_reduced$alt_status
)

# motif vectors
loadings <- data.frame(
  PC1 = pca$rotation[,1],
  PC2 = pca$rotation[,2],
  comp_names = c(rownames(pca$rotation)[1:3], sapply(strsplit(rownames(pca$rotation)[4:12], split = '_', fixed = TRUE), `[`, 1))
)

# scale the loadings (smaller than sample coordinates) to make arrows more visible
loadings$PC1 <- loadings$PC1 * 10
loadings$PC2 <- loadings$PC2 * 10

library(ggplot2)
library(ggrepel)

p <- ggplot(scores, aes(PC1, PC2, color = ALT)) +
  geom_point(alpha = 0.7, size = 2) +

  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.25,"cm")),
               inherit.aes = FALSE,
               color = "black") +

  geom_text_repel(data = loadings,
                  aes(PC1, PC2, label = comp_names),
                  inherit.aes = FALSE,
                  size = 3.5) +

  theme_bw() +
  coord_equal() +
  labs(
    title = "PCA biplot without TF rate",
    x = "PC1",
    y = "PC2"
  )

plot(p)

ggsave("../outputs/PCA_biplot_red_tf.pdf", plot = p, width = 7, height = 6)
```


# Split the dataset
```{r train-test}
set.seed(21)

train_index <- sample(1:nrow(model_data_reduced), size=0.7*nrow(model_data_reduced))
train_red <- model_data_reduced[train_index,]
test_red <- model_data_reduced[-train_index,]

# Check distribution
table(train_red$alt_status, dnn = 'train')
table(test_red$alt_status, dnn = 'test')
```

# First classifier: Logistic Regression

```{r logreg-reduced}
logreg_model_reduced <- glm(
  alt_status ~ .,
  data = train_red,
  family = "binomial"
)

prob_red <- predict(
  logreg_model_reduced,
  test_red,
  type = "response"
)
```

## Evaluate performance

```{r ROC-AUC-red-logreg}
library(pROC, quietly = T)
roc_red <- roc(
  test_red$alt_status,
  prob_red,
  levels = c("ALT-low","ALT-high")
)

plot(roc_red, main = 'ROC curve - Log. Reg. (all features except TF rate)')

auc(roc_red)
```

```{r pred-prob-distribution, warning=FALSE} 
prob_df <- data.frame(
  prob = prob_red,
  alt_status = test_red$alt_status
)

library(ggplot2)

ggplot(prob_df, aes(x = prob, fill = alt_status)) +
  geom_density(alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Predicted probability distributions",
    x = "Predicted probability of ALT-high",
    y = "Density"
  )
```

### Other metrics

```{r confusion-matrix-logreg-red}
pred_class <- ifelse(prob_red >= 0.5, "ALT-high", "ALT-low")
pred_class <- factor(pred_class, levels = c("ALT-low", "ALT-high"))

library(caret, quietly = TRUE)
confusionMatrix(pred_class, test_red$alt_status, positive = 'ALT-high')
```

# Second classifier: Random Forest

```{r rf-red}
library(randomForest, quietly = T)
rf_red <- randomForest(
  alt_status ~ .,
  data = train_red,
  ntree = 500,
  importance = TRUE
)

rf_pred_red <- predict(
  rf_red,
  test_red,
  type = "prob"
)[,"ALT-high"]
```

## Evaluate performance

```{r ROC-AUC-red-rf}
roc_rf_red <- roc(
  test_red$alt_status,
  rf_pred_red,
  levels = c("ALT-low","ALT-high")
)

plot(roc_rf_red, main = 'ROC curve - RF (all features except TF rate)')

auc(roc_rf_red)
```

### Other metrics

```{r confusion-matrix-RF}
rf_pred_red <- ifelse(rf_pred_red >= 0.5, "ALT-high", "ALT-low")
rf_pred_red <- factor(rf_pred_red, levels = c("ALT-low", "ALT-high"))

library(caret)

confusionMatrix(
  rf_pred_red,
  test_red$alt_status,
  positive = "ALT-high"
)
```


```{r feature-importance}
varImpPlot(rf_red, main = 'Variable importance plot - Random Forest (all features except TF rate)' )
```


# Cross Validation of the performances

To assess the robustness of these two predictive models, we conducted **repeated stratified cross-validation**. 

## Logistic Regression - CV

```{r logreg-CV, message=FALSE, warning=FALSE}
library(pROC)

set.seed(21)

cv_data <- model_data_reduced
k <- 5
n_repeats <- 5

auc_results <- c()

for (r in 1:n_repeats) {
  # stratified fold assignment
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    logreg_cv_model <- glm(alt_status~., data = train_cv, family = 'binomial')
    
    logreg_cv_prob <- predict(logreg_cv_model, test_cv, type = 'response')
    
    roc_obj <- roc(
      response = test_cv$alt_status,
      predictor = logreg_cv_prob,
      levels = c("ALT-low", "ALT-high")
    )
    
    auc_results <- c(auc_results, as.numeric(auc(roc_obj)))
  }
}

mean(auc_results)
# sd(auc_results)
# auc_results
```

## Random Forest - CV

```{r RF-CV, message=FALSE}
library(randomForest)
library(pROC)

set.seed(21)

cv_data <- model_data_reduced
k <- 5
n_repeats <- 5

auc_results <- c()

for (r in 1:n_repeats) {
  # stratified fold assignment
  fold_id <- integer(nrow(cv_data))
  
  idx_low  <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_cv <- cv_data[fold_id != fold, ]
    test_cv  <- cv_data[fold_id == fold, ]
    
    rf_cv_model <- randomForest(
      alt_status ~ .,
      data = train_cv,
      ntree = 500,
      mtry = 2
    )
    
    rf_cv_prob <- predict(rf_cv_model, test_cv, type = "prob")[, "ALT-high"]
    
    roc_obj <- roc(
      response = test_cv$alt_status,
      predictor = rf_cv_prob,
      levels = c("ALT-low", "ALT-high")
    )
    
    auc_results <- c(auc_results, as.numeric(auc(roc_obj)))
  }
}

mean(auc_results)
# sd(auc_results)
# auc_results
```

These results confirm the **robustness of the AUC values previously recorded** for the two models, and agree with the fact that allowing non-linear relationships among features (i.e. using Random Forest) provide a better classification power.