---
title: "08_feature_ablation"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

## Prepare dataset

```{r setup}
library(readxl)
library(tidyverse)
library(ggplot2)
library(caret)
library(dplyr)
source('../scripts/clean_columns.R')

data <- read_xlsx('../data/raw/PCAWG_variables_PCA_woK.xlsx')
clean_data <- clean_pcawg_data(data)
model_data <- clean_data[complete.cases(clean_data), -c(1,2)]
model_data$alt_status <- factor(model_data$alt_status, levels=c("ALT-low","ALT-high"), labels = c("ALT_low","ALT_high"))
summary(model_data)
```

## Define feature sets

```{r feature-sets}
feature_sets <- list(
  full = model_data,
  no_tel_insert = model_data[, !names(model_data) %in% "telomere_insertion_rate"],
  no_tf = model_data[, !names(model_data) %in% "tf_rate"],
  no_TERT = model_data[, !names(model_data) %in% "TERT_FPKM"],
  no_tel_content = model_data[, !names(model_data) %in% "telomere_content_log2"],
  no_TVRs = model_data[, -c(grep('singleton', names(model_data), ))]
)
```

## CV setup

```{r cv-setup}
set.seed(21)
folds <- createMultiFolds(model_data$alt_status, k = 5, times = 20)

ctrl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 20,
  index = folds,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = 'final'
)
```

# Train models

```{r training, warning=FALSE, message=FALSE}
library(caret)
library(pROC)
library(dplyr)
library(randomForest)

results <- data.frame()
all_resamples <- list()

for (fs in names(feature_sets)) {
  
  df <- feature_sets[[fs]]
  
  log_model <- train(
    alt_status ~ .,
    data = df,
    method = "glm",
    family = "binomial",
    metric = "ROC",
    trControl = ctrl
  )
  
  rf_model <- train(
    alt_status ~ .,
    data = df,
    method = "rf",
    metric = "ROC",
    trControl = ctrl
  )
  
  models <- list(LogReg = log_model, RF = rf_model)
  
  for (m in names(models)) {
    
    model <- models[[m]]
    
    rs <- model$resample
    
    rs$feature_set <- fs
    rs$model <- m
    
    all_resamples[[paste(fs, m, sep = "_")]] <- rs
    
    results <- rbind(
      results,
      data.frame(
        feature_set = fs,
        model = m,
        mean_AUC = mean(rs$ROC, na.rm = TRUE),
        sd_AUC = sd(rs$ROC, na.rm = TRUE),
        mean_sensitivity = mean(rs$Sens, na.rm = TRUE),
        mean_specificity = mean(rs$Spec, na.rm = TRUE)
      )
    )
  }
}

results <- results %>%
  group_by(model) %>%
  mutate(
    delta_AUC = mean_AUC - mean_AUC[feature_set == "full"]
  ) %>%
  ungroup() 

print(results, n = 12)
```

# Plots

```{r plotting}
ggplot(results, aes(x = feature_set, y = mean_AUC, fill = model)) +
  geom_col(position = position_dodge2()) +
  geom_errorbar(
    aes(ymin = mean_AUC - sd_AUC, ymax = mean_AUC + sd_AUC),
    position = position_dodge(width = 0.9),
    width = 0.2
  ) +
  theme_bw()+
  labs(x='Feature subset',
       y='Average AUC',
       title='Performance of models across feature subsets')

ggplot(results, aes(x = feature_set, y = delta_AUC, fill = model)) +
  geom_col(position = position_dodge2()) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Feature subset",
    y = "Change in ROC AUC vs full model",
    title = "Impact of feature removal on model performance"
  ) +
  theme_bw()

results_long <- results %>%
  select(feature_set, model, mean_sensitivity, mean_specificity) %>%
  pivot_longer(
    cols = c(mean_sensitivity, mean_specificity),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      mean_sensitivity = "Sensitivity",
      mean_specificity = "Specificity"
    )
  )

ggplot(results_long, aes(x = feature_set, y = value, fill = feature_set)) +
  geom_col(show.legend = FALSE) +
  labs(
    x = "Feature subset",
    y = "Average value",
    title = "Sensitivity and specificity across feature subsets"
  ) +
  theme_bw() +
  facet_grid(metric ~ model)+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The **small increases** in AUC that we see we have after removing certain features such as telomeric content, TERT, or TF rate **are normal** and are due to random variation in the data and the training process. This can happen when the **removed feature is not very informative** for the model, or when **its contribution is overshadowed** by other features.

**Removing TVR motif** features causes the **largest performance decrease** in both models, particularly for Random Forest, suggesting that telomere variant repeat composition contains strong predictive information for ALT status. **Removing telomere insertion rate** also **noticeably reduces performance**.

Of course, **when removing TVR motifs we are removing 9 individual features instead of one**, and thus an abrupt decrease in performance can be expected. Although informative on TVR motifs contribution to classification, **the comparison is not fair**. What could be interesting is to perform a **LOFO on each individual TVR motif**, looking for the ones having more discriminatory power.