---
title: "09_cell_lines"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

```{r setup}
library(tidyverse)
library(readxl)
library(here)
source(here('scripts', 'useful_plots.R'))
data <- read_xlsx(here('data', 'raw', 'tab1.xlsx'))

```

# Motifs subset

We keep only a subset of variables, corresponding to the ones we have available for primary tumors.

```{r select-columns}
pca_data <- data %>% 
  select(`ALT assay strength`, TelFusDetector_rate, TelFuse_newtelomere, TelSeq, TERT_exp, ATAGGG, CTAGGG, GTAGGG, TAAGGG, TCAGGG, TGAGGG, TTCGGG, TTTGGG) %>%
  drop_na() %>% 
  rename(alt_status=`ALT assay strength`, tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status))
```

We retain 83 observations (63 experimentally confirmed as non-ALT, 20 as ALT) but **we don't have the TTGGGG column**.

```{r pca-cells}
pcaplt <- pca_plot(pca_data, title = 'PCA (motif subset)')
pcaplt$plot
```

```{r biplot}
biplot <- pca_biplot(pca_data, title = 'PCA biplot (motif subset)', scale_arrows = 5)
biplot$plot
```

# Repeat using TVR column instead of individual motifs
```{r tvr}
pca_data <- data %>% 
  select(`ALT assay strength`, TelFusDetector_rate, TelFuse_newtelomere, TelSeq, TVR, TERT_exp) %>%
  drop_na() %>% 
  rename(alt_status=`ALT assay strength`, tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status))
```

```{r pca-cells-tvr}
pcaplt <- pca_plot(pca_data, title = 'PCA plot (TVR feature)')
pcaplt$plot
```

```{r biplot-tvr}
biplot <- pca_biplot(pca_data, title = 'PCA biplot (TVR feature)', scale_arrows = 5)
biplot$plot

ggsave(here('outputs', 'cell_lines_PCA_biplot.pdf'))
```

This is an **almost exact replica of the paper's Fig 1B**. 

# Decompose TVR feature in individual motifs

```{r tvr-decomp}
pca_data <- data %>% 
  select(`ALT assay strength`, TelFusDetector_rate, TelFuse_newtelomere, TelSeq, 6:54, TERT_exp) %>%
  drop_na() %>% 
  rename(alt_status=`ALT assay strength`, tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status))
```

```{r pca-cells-tvr-decomp}
pcaplt <- pca_plot(pca_data, title = 'PCA (all motifs)')
pcaplt$plot
```

```{r biplot-tvr-decomp}
biplot <- pca_biplot(pca_data, title = 'PCA biplot (all motifs)', scale_arrows = 30)
biplot$plot
```

# Interpretation and discussion

In the **PCAWG primary tumor dataset**, the biplot revealed a clear separation of ALT-high tumors along PC1 (53.1%), driven
primarily by tel_insertion_rate, tf_rate, and tel_content. Notably, specific TVR motifs, in particular TTCGGG and TGAGGG, loaded
in the direction of ALT-high samples, while TTTGGG loaded in the opposite direction. These results are **consistent with published literature** and **confirmed by supervised classification results**.

In the **cell line dataset (Bennett et al., 2026)**, replicating the authors' Figure 1B using the pre-computed batch-corrected
TVR feature confirmed that TVR content, together with neo-telomere structures and telomere fusion rate, are the strongest
contributors to ALT-positive separation along PC1 (67.5%), while TERT expression loads in the opposite direction. This is fully
**consistent with the original paper and with the PCAWG results**.

Two additional exploratory biplots were generated using raw individual motif frequencies from the cell line dataset. In both
cases, the motif vectors pointed toward ALT-negative samples, **opposite to what was observed in the PCAWG dataset**. 
**In PCAWG, motif values are per-sample normalized z-scores** (singleton distances), whereas **in the cell line dataset they are raw frequencies** subject to sequencing depth variation and batch effects. Without applying the batch-correction pipeline used
by Bennett et al. (log-transformation followed by regression on depth-derived principal components) individual motif frequencies
are not directly interpretable in the context of ALT biology, and direct comparison with the PCAWG results is not valid.


# Restrict to primary tumors only

Importantly, I noticed that **the training set they used**, and on which I replicated their PCA, **contained both primary AND metastatic tumors**. We could see if, retaining only primary tumors, the same feature contributions to PCA emerge.

```{r setup_primary}
pca_data_primary <- data %>% 
  filter(PrimaryOrMetastasis == 'Primary') %>% 
  select(`ALT assay strength`, TelFusDetector_rate, TelFuse_newtelomere, TelSeq, TERT_exp, ATAGGG, CTAGGG, GTAGGG, TAAGGG, TCAGGG, TGAGGG, TTCGGG, TTTGGG) %>%
  drop_na() %>% 
  rename(alt_status=`ALT assay strength`, tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status))
```

We have **only 50 observations** (38 non-ALT, 12 ALT).

```{r pca-biplot-cells-primary}
biplot <- pca_biplot(pca_data_primary, title = 'PCA biplot on primary tumors (motif subset)')
biplot$plot
```

```{r biplot-tvr-primary}
pca_data_primary <- data %>% 
  filter(PrimaryOrMetastasis=='Primary') %>% 
  select(`ALT assay strength`, TelFusDetector_rate, TelFuse_newtelomere, TelSeq, TVR, TERT_exp) %>%
  drop_na() %>% 
  rename(alt_status=`ALT assay strength`, tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status))

biplot <- pca_biplot(pca_data_primary, title = 'PCA biplot on primary tumors (TVR feature)', scale_arrows = 5)
biplot$plot
```

```{r tvr-decomp-primary}
pca_data_primary <- data %>% 
  filter(PrimaryOrMetastasis=='Primary') %>% 
  select(`ALT assay strength`, TelFusDetector_rate, TelFuse_newtelomere, TelSeq, 6:54, TERT_exp) %>%
  drop_na() %>% 
  rename(alt_status=`ALT assay strength`, tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status))
```
We have **even fewer observations** (20 non-ALT, 12 ALT), since many rows have NA values in individual motifs columns.

```{r pca-biplot-tvr-decomp-primary}
biplot <- pca_biplot(pca_data_primary, title = 'PCA biplot on primary tumors (all motifs)', scale_arrows = 30)
biplot$plot
```

By looking at the resulting PCA biplots, we can see how the **contributions of features in the PCA is very much the same as before**, even when restricting the analysis only for primary tumor cell lines.


# Using a larger dataset

We will now re-do the analysis, **enlarging the dataset to more observations**, coming from outside the training set. In particular, we will use observations for which no `ALT assay strength` value is present, but having a predicted value in `ALT_pred_class` with a strong enough `ALT_pred_prob`.

```{r pca-biplot-cells-larger, echo=FALSE}
pca_data_larger <- data %>% 
  select(`ALT assay strength`, ALT_pred_class, TelFusDetector_rate, TelFuse_newtelomere, 
         TelSeq, ATAGGG, CTAGGG, GTAGGG, TAAGGG, TCAGGG, TGAGGG, TTCGGG, TTTGGG, TERT_exp) %>%
  mutate(alt_status = case_when(
    !is.na(`ALT assay strength`) ~ `ALT assay strength`,
    !is.na(ALT_pred_class) ~ ALT_pred_class,
    TRUE ~ NA_character_
  )) %>% 
  mutate(alt_status = case_when(
    alt_status=='Negative' ~ 'ALT-',
    alt_status=='Positive (Strong evidence)' ~ 'ALT+',
    alt_status=='Positive (Weak evidence)' ~ NA_character_,
    TRUE ~ alt_status
  )) %>% 
  rename(tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status),
         `ALT assay strength`=NULL,
         ALT_pred_class=NULL) %>%
  select(alt_status, everything()) %>% 
  drop_na()

biplot <- pca_biplot(pca_data_larger, title = 'PCA biplot (larger dataset, motif subset)', scale_arrows = 15)
biplot$plot
```


```{r pca-biplot-tvr-larger, echo=FALSE}
pca_data_larger <- data %>% 
  select(`ALT assay strength`, ALT_pred_class, TelFusDetector_rate, TelFuse_newtelomere, 
         TelSeq, TVR, TERT_exp) %>%
  mutate(alt_status = case_when(
    !is.na(`ALT assay strength`) ~ `ALT assay strength`,
    !is.na(ALT_pred_class) ~ ALT_pred_class,
    TRUE ~ NA_character_
  )) %>% 
  mutate(alt_status = case_when(
    alt_status=='Negative' ~ 'ALT-',
    alt_status=='Positive (Strong evidence)' ~ 'ALT+',
    alt_status=='Positive (Weak evidence)' ~ NA_character_,
    TRUE ~ alt_status
  )) %>% 
  rename(tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status),
         `ALT assay strength`=NULL,
         ALT_pred_class=NULL) %>%
  select(alt_status, everything()) %>% 
  drop_na()

biplot <- pca_biplot(pca_data_larger, title = 'PCA biplot (larger dataset, TVR feature)')
biplot$plot
```

```{r pca-biplot-decomp-larger, echo=FALSE}
pca_data_larger <- data %>% 
  select(`ALT assay strength`, ALT_pred_class, TelFusDetector_rate, TelFuse_newtelomere, 
         TelSeq, 6:54, TERT_exp) %>%
  mutate(alt_status = case_when(
    !is.na(`ALT assay strength`) ~ `ALT assay strength`,
    !is.na(ALT_pred_class) ~ ALT_pred_class,
    TRUE ~ NA_character_
  )) %>% 
  mutate(alt_status = case_when(
    alt_status=='Negative' ~ 'ALT-',
    alt_status=='Positive (Strong evidence)' ~ 'ALT+',
    alt_status=='Positive (Weak evidence)' ~ NA_character_,
    TRUE ~ alt_status
  )) %>% 
  rename(tf_rate = TelFusDetector_rate, tel_insertion_rate = TelFuse_newtelomere, tel_content = TelSeq) %>% 
  mutate(alt_status = as.factor(alt_status),
         `ALT assay strength`=NULL,
         ALT_pred_class=NULL) %>%
  select(alt_status, everything()) %>% 
  drop_na()

biplot <- pca_biplot(pca_data_larger, title = 'PCA biplot (larger dataset, all motifs)', scale_arrows = 30)
biplot$plot
```
