---
title: "10_TabPFN"
author: "Andrea Polazzi"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

# Setup

```{r setup}
library(reticulate, quietly = TRUE)
tabpfn_venv <- here::here('analysis', '.venv-tabpfn')
use_virtualenv(tabpfn_venv, required = TRUE)
py_config()

library(readxl, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(pROC, quietly = TRUE)
library(caret, quietly = TRUE)
library(here)

source(here('scripts', 'clean_columns.R'))
source(here('scripts', 'clf_metrics.R'))

data <- read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))

features14 <- c(
  'tf_primary_rate', 'tf_blood_rate', 'telomere_insertion_rate',
  'telomere_content_log2', 'TERT_FPKM',
  'ATAGGG_singleton_dist', 'CTAGGG_singleton_dist', 'GTAGGG_singleton_dist',
  'TAAGGG_singleton_dist', 'TCAGGG_singleton_dist', 'TGAGGG_singleton_dist',
  'TTCGGG_singleton_dist', 'TTGGGG_singleton_dist', 'TTTGGG_singleton_dist'
)

model_full <- data %>%
  select(all_of(features14), alt_status, donor_id, cancer_type, cancer_group) %>%
  filter(complete.cases(across(all_of(features14)))) %>%
  mutate(alt_status = factor(alt_status, levels = c('ALT-low', 'ALT-high')))

model_data <- model_full %>% select(all_of(features14), alt_status)

summary(model_data)
```

# Export for out-of-fold SHAP (5-fold CV)

```{r export-shap-cv}
# SHAP uses one 5-fold partition = repeat 1 of the shared CV folds (scripts/make_folds.R)
source(here('scripts', 'make_folds.R'))
folds_vec <- make_folds(model_data$alt_status, k = 5, n_repeats = 5, seed = 21)[, 1]

shap_cv_dir <- here('outputs', 'shap_input_cv')
dir.create(shap_cv_dir, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(model_data %>% select(all_of(features14)),
                 file.path(shap_cv_dir, 'X_full.csv'))
readr::write_csv(data.frame(y = as.integer(model_data$alt_status == 'ALT-high')),
                 file.path(shap_cv_dir, 'y_full.csv'))
readr::write_csv(data.frame(fold = folds_vec),
                 file.path(shap_cv_dir, 'folds.csv'))
readr::write_csv(model_full %>%
                   transmute(donor_id, cancer_type, cancer_group, alt_status),
                 file.path(shap_cv_dir, 'meta.csv'))

cat(sprintf("Exported %d samples x %d features to %s\n",
            nrow(model_data), length(features14), shap_cv_dir))
table(folds_vec, model_data$alt_status)
```

# Stratified CV

We evaluate TabPFN under the same repeated **5-fold × 5-repeat stratified CV** used for Log. Reg., RF and LASSO (notebooks 02, 09, 20). TabPFN is **not** called through the cloud API here: the heavy compute runs on the GPU box (carmela) with the local `tabpfn` package via `scripts/tabpfn_cv_oof.py`, which writes per-sample out-of-fold probabilities. This chunk only builds and exports the cohort + folds; the next chunk reads the probabilities back and scores them with the shared `fold_metrics()` helper.

```{r cv-export}
source(here('scripts', 'make_folds.R'))
fold_mat <- make_folds(model_data$alt_status, k = 5, n_repeats = 5, seed = 21)

cmp_dir <- here('outputs', 'shap_input_cmp_cv')
dir.create(cmp_dir, recursive = TRUE, showWarnings = FALSE)
readr::write_csv(model_data %>% select(all_of(features14)),
                 file.path(cmp_dir, 'X_full.csv'))
readr::write_csv(data.frame(y = as.integer(model_data$alt_status == 'ALT-high')),
                 file.path(cmp_dir, 'y_full.csv'))
readr::write_csv(as.data.frame(fold_mat), file.path(cmp_dir, 'folds_5x5.csv'))
cat('Exported cohort + folds to', cmp_dir, "\n",
    'On carmela:  python tabpfn_cv_oof.py --input-dir shap_input_cmp_cv',
    "--output outputs/tabpfn_cv_oof.csv\n")
```

```{r cv-results}
oof_csv <- here('outputs', 'tabpfn_cv_oof.csv')
if (!file.exists(oof_csv)) {
  cat("TabPFN out-of-fold probabilities not found.\n",
      'Run scripts/tabpfn_cv_oof.py on the GPU box (carmela), then sync',
      "outputs/tabpfn_cv_oof.csv back.\n")
} else {
  tabpfn_perfold <- readr::read_csv(oof_csv, show_col_types = FALSE) %>%
    mutate(truth = factor(ifelse(y == 1, 'ALT-high', 'ALT-low'),
                          levels = c('ALT-low', 'ALT-high'))) %>%
    group_by(rep, fold) %>%
    group_modify(~ fold_metrics(.x$yhat, .x$truth)) %>%
    ungroup()

  cat("=== TabPFN per-fold metrics (mean ± SD across 25 folds) ===\n")
  tabpfn_perfold %>%
    summarise(across(c(auc, pr_auc, sensitivity, specificity, precision, f1, balanced_accuracy),
                     list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)),
                     .names = '{.col}_{.fn}')) %>%
    pivot_longer(everything(), names_to = c('metric', '.value'),
                 names_sep = '_(?=[^_]+$)') %>%
    mutate(summary = paste0(round(mean, 4), ' ± ', round(sd, 4))) %>%
    select(metric, summary) %>%
    print()
}
```

# Interpretability (ALT status — out-of-fold SHAP, HPC)

## SHAP plots

```{r tabpfn-shap-hpc-test}
source(here('scripts', 'reconstruct_shapviz.R'))
# rank01() and recolor_by_rank() are defined in reconstruct_shapviz.R (sourced above)

sv_imp_rank <- recolor_by_rank(sv_imp)
sv_rec_rank <- recolor_by_rank(sv_rec)
sv_imp2_rank <- recolor_by_rank(sv_imp2)
sv_rec2_rank <- recolor_by_rank(sv_rec2)

# Imputation, single split train/test
# sv_importance(sv_imp_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8)
# sv_dependence(sv_imp_rank, v = 'telomere_insertion_rate', color_var = 'tf_rate')
# sv_dependence(sv_imp_rank, v = 'telomere_content_log2', color_var = 'TERT_FPKM') + theme_bw()
# sv_dependence(sv_imp_rank, v = 'tf_rate',         color_var = 'telomere_content_log2') + theme_bw()
# sv_dependence(sv_imp_rank, v = 'TERT_FPKM',             color_var = 'TTTGGG_singleton_dist') + theme_bw()

# Imputation, 5-fold CV
# beeswarm has no per-point shape (it stacks all 14 features); shape lives on the
# dependence plots, where alt_status (the target) is mapped via sv_dependence_shape().
sv_importance(sv_imp2_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8) + theme_bw()
sv_dependence_shape(sv_imp2_rank, v = 'telomere_insertion_rate', color_var = 'tf_primary_rate',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_imp2_rank, v = 'telomere_content_log2', color_var = 'TERT_FPKM',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_imp2_rank, v = 'tf_primary_rate', color_var = 'telomere_content_log2',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_imp2_rank, v = 'TERT_FPKM', color_var = 'TTTGGG_singleton_dist',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_imp2_rank, v = 'TTTGGG_singleton_dist', color_var = 'TERT_FPKM',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_imp2_rank, v = 'TGAGGG_singleton_dist', color_var = 'telomere_insertion_rate',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_imp2_rank, v = 'tf_blood_rate', color_var = 'tf_primary_rate',
                    group = meta2$alt_status) + theme_bw()
# Recontextualization, single split train/test
# sv_importance(sv_rec_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8)
# sv_dependence(sv_rec_rank, v = 'telomere_insertion_rate', color_var = 'tf_rate')
# sv_dependence(sv_rec_rank, v = 'telomere_content_log2', color_var = 'TERT_FPKM') + theme_bw()
# sv_dependence(sv_rec_rank, v = 'tf_rate',         color_var = 'telomere_content_log2') + theme_bw()
# sv_dependence(sv_rec_rank, v = 'TERT_FPKM',             color_var = 'TTTGGG_singleton_dist') + theme_bw()

# Recontextualization, 5-fold CV
sv_importance(sv_rec2_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8) + theme_bw() + title("SHAP beeswarm plot of TabPFN ALT classifier", sub = 'Rank-average normalized features. 5-fold CV')



sv_dependence_shape(sv_rec2_rank, v = 'telomere_insertion_rate', color_var = 'tf_primary_rate', group = meta2$alt_status) + theme_bw()
sv_dependence_top(sv_rec2_rank, v = 'telomere_insertion_rate', group = meta2$cancer_type, top_n = 20, shape_group = meta2$alt_status) + theme_bw()
sv_dependence_mes(sv_rec2_rank, v = 'telomere_insertion_rate', group = meta2$cancer_group, shape_group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_rec2_rank, v = 'telomere_content_log2', color_var = 'tf_primary_rate',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_rec2_rank, v = 'tf_primary_rate', color_var = 'telomere_insertion_rate', group = meta2$alt_status) + theme_bw()

sv_dependence_top(sv_rec2_rank, v = 'tf_primary_rate', group = meta2$cancer_type, top_n = 20,
                  shape_group = meta2$alt_status) + theme_bw()
sv_dependence_mes(sv_rec2_rank, v = 'tf_primary_rate', group = meta2$cancer_group, shape_group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_rec2_rank, v = 'TERT_FPKM', color_var = 'TTTGGG_singleton_dist',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_rec2_rank, v = 'TTTGGG_singleton_dist', color_var = 'TERT_FPKM',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_rec2_rank, v = 'TGAGGG_singleton_dist', color_var = 'telomere_insertion_rate',
                    group = meta2$alt_status) + theme_bw()
sv_dependence_shape(sv_rec2_rank, v = 'tf_blood_rate', color_var = 'tf_primary_rate',
                    group = meta2$alt_status) + theme_bw()
```

------------------------------------------------------------------------

# TabPFN Regression

We apply TabPFN to two regression tasks that mirror notebooks 13_1 (`tf_primary_rate`) and 15_1 (`tf_blood_rate`). The same 5-fold × 5-repeat stratified CV and the same metrics (RMSE, MAE, Spearman ρ) are used to allow direct comparison with LM, LASSO, and RF.

## Data setup

```{r reg-setup}
reg_data_raw <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
```

------------------------------------------------------------------------

# tTF regression

Predicting `tf_primary_rate` (log1p scale) from all features including `tf_blood_rate`, mirroring notebook 13_1. The strong outlier (`tf_primary_rate >= 10`) is removed as in the reference analysis.

```{r ttf-data}
reg_data_ttf <- reg_data_raw %>%
  drop_na() %>%
  filter(tf_primary_rate < 10) %>%
  mutate(
    tf_primary_rate         = log1p(tf_primary_rate),
    tf_blood_rate           = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM               = log1p(TERT_FPKM),
    alt_status              = as.numeric(alt_status == 'ALT-high')
  ) %>%
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)

x_ttf_r <- reg_data_ttf %>% select(-tf_primary_rate)
y_ttf_r  <- reg_data_ttf$tf_primary_rate
```

## CV results (tTF)

TabPFN regression runs on the GPU box (carmela) with the local `tabpfn` package via `scripts/tabpfn_cv_oof.py --task regression` (no cloud API). The cohort + shared folds are exported by the regression comparison notebook 21; here we read the out-of-fold predictions back and score them with the shared `reg_fold_metrics()`.

```{r ttf-cv-results}
source(here('scripts', 'reg_metrics.R'))
oof_ttf <- here('outputs', 'tabpfn_reg_ttf_oof.csv')
if (!file.exists(oof_ttf)) {
  cat('TabPFN tTF out-of-fold predictions not found.\n',
      'Run (on carmela): python tabpfn_cv_oof.py --task regression',
      '--input-dir cmp_reg_ttf_cv --output outputs/tabpfn_reg_ttf_oof.csv\n')
} else {
  ttf_metrics <- readr::read_csv(oof_ttf, show_col_types = FALSE) %>%
    group_by(rep, fold) %>%
    group_modify(~ reg_fold_metrics(.x$yhat, .x$y)) %>%
    ungroup()
  ttf_metrics %>%
    summarise(across(c(rmse, mae, spearman),
                     list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)),
                     .names = '{.col}_{.fn}')) %>%
    pivot_longer(everything(), names_to = c('metric', '.value'),
                 names_sep = '_(?=[^_]+$)') %>%
    mutate(summary = paste0(round(mean, 4), ' ± ', round(sd, 4))) %>%
    select(metric, summary)
}
```

## Interpretability (tTF — out-of-fold SHAP, HPC)

Out-of-fold SHAP over a single stratified 5-fold split (stratified on `alt_status`, `set.seed(42)`), mirroring the classification `full2` flow. Each sample is explained by the `TabPFNRegressor` that did not train on it; values are in the target (log1p) space.

```{r ttf-export-shap-cv}
source(here('scripts', 'make_folds.R'))

# meta + stratification source, row-aligned with reg_data_ttf (same drop_na + filter)
reg_ttf_filt <- reg_data_raw %>% drop_na() %>% filter(tf_primary_rate < 10)
stopifnot(nrow(reg_ttf_filt) == nrow(reg_data_ttf))

# SHAP uses one 5-fold partition = repeat 1 of the shared CV folds
folds_ttf <- make_folds(factor(reg_data_ttf$alt_status), k = 5, n_repeats = 5, seed = 21)[, 1]

shap_ttf_in <- here('outputs', 'shap_input_ttf_cv')
dir.create(shap_ttf_in, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(as.data.frame(x_ttf_r),          file.path(shap_ttf_in, 'X_full.csv'))
readr::write_csv(data.frame(y = y_ttf_r),         file.path(shap_ttf_in, 'y_full.csv'))
readr::write_csv(data.frame(fold = folds_ttf),    file.path(shap_ttf_in, 'folds.csv'))
readr::write_csv(reg_ttf_filt %>%
                   transmute(donor_id, cancer_type, cancer_group, alt_status),
                 file.path(shap_ttf_in, 'meta.csv'))

cat(sprintf("Exported %d samples x %d features to %s\n",
            nrow(x_ttf_r), ncol(x_ttf_r), shap_ttf_in))
table(folds_ttf, reg_ttf_filt$alt_status)
```

```{r ttf-shap-load}
source(here('scripts', 'reconstruct_shapviz.R'))

shap_ttf_imp_dir <- here('outputs', 'shap_output_ttf_imp')
shap_ttf_ready   <- all(file.exists(
  file.path(shap_ttf_imp_dir, c('shap_values.csv', 'X_test_explained.csv', 'shap_baseline.txt'))
))

if (!shap_ttf_ready) {
  cat("OOF SHAP results for tTF regression not yet available.\n")
  cat("Run on the GPU box (carmela), then sync outputs/ back:\n")
  cat("  bash scripts/run_shap_reg.sh\n")
} else {
  sv_ttf_imp <- load_shap(shap_ttf_imp_dir)                          # target (log1p) space
  sv_ttf_rec <- load_shap(here('outputs', 'shap_output_ttf_rec'))    # recontext target space
  meta_ttf   <- load_meta(shap_ttf_imp_dir)

  # rank-average normalization (same as the ALT-status SHAP plots): beeswarm
  # colors and dependence axes use per-feature rank, robust to outliers/scale.
  sv_ttf_imp_rank <- recolor_by_rank(sv_ttf_imp)
  sv_ttf_rec_rank <- recolor_by_rank(sv_ttf_rec)

  # --- imputation (additive, target log1p space) ---
  # explicit print(): inside this if/else block only the last expression would
  # auto-print, unlike a top-level chunk.
  print(sv_importance(sv_ttf_imp_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8) + theme_bw())
  
  print(sv_dependence_shape(sv_ttf_imp_rank, v = 'tf_blood_rate', color_var = 'telomere_insertion_rate', group = meta_ttf$alt_status) + theme_bw())
  
  print(sv_dependence_shape(sv_ttf_imp_rank, v = 'telomere_insertion_rate', color_var = 'TTCGGG_singleton_dist', group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_ttf_imp_rank, v = 'telomere_insertion_rate', group = meta_ttf$cancer_type, top_n = 20, shape_group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_ttf_imp_rank, v = 'telomere_insertion_rate', group = meta_ttf$cancer_group, shape_group = meta_ttf$alt_status) + theme_bw())

  print(sv_dependence_shape(sv_ttf_imp_rank, v = 'TERT_FPKM', color_var = 'telomere_content_log2', group = meta_ttf$alt_status) + theme_bw())

  print(sv_dependence_shape(sv_ttf_imp_rank, v = 'telomere_content_log2', color_var = 'telomere_insertion_rate', group = meta_ttf$alt_status) + theme_bw())
  
  print(sv_dependence_shape(sv_ttf_imp_rank, v = 'TTCGGG_singleton_dist', color_var = 'telomere_insertion_rate', group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_ttf_imp_rank, v = 'TTCGGG_singleton_dist', group = meta_ttf$cancer_type, top_n = 20, shape_group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_ttf_imp_rank, v = 'TTCGGG_singleton_dist', group = meta_ttf$cancer_group, shape_group = meta_ttf$alt_status) + theme_bw())

  # --- recontextualization (native target space cross-check) ---
  print(sv_importance(sv_ttf_rec_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8) + theme_bw())
  
  print(sv_dependence_shape(sv_ttf_rec_rank, v = 'tf_blood_rate', color_var = 'telomere_insertion_rate', group = meta_ttf$alt_status) + theme_bw())
  
  print(sv_dependence_shape(sv_ttf_rec_rank, v = 'telomere_insertion_rate', color_var = 'TTCGGG_singleton_dist', group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_ttf_rec_rank, v = 'telomere_insertion_rate', group = meta_ttf$cancer_type, top_n = 20, shape_group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_ttf_rec_rank, v = 'telomere_insertion_rate', group = meta_ttf$cancer_group, shape_group = meta_ttf$alt_status) + theme_bw())

  print(sv_dependence_shape(sv_ttf_rec_rank, v = 'TERT_FPKM', color_var = 'telomere_content_log2', group = meta_ttf$alt_status) + theme_bw())
  
  print(sv_dependence_shape(sv_ttf_rec_rank, v = 'telomere_content_log2', color_var = 'telomere_insertion_rate', group = meta_ttf$alt_status) + theme_bw())
  
  print(sv_dependence_shape(sv_ttf_rec_rank, v = 'TTCGGG_singleton_dist', color_var = 'telomere_insertion_rate', group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_ttf_rec_rank, v = 'TTCGGG_singleton_dist', group = meta_ttf$cancer_type, top_n = 20, shape_group = meta_ttf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_ttf_rec_rank, v = 'TTCGGG_singleton_dist', group = meta_ttf$cancer_group, shape_group = meta_ttf$alt_status) + theme_bw())

}
```

------------------------------------------------------------------------

# bTF regression

Predicting `tf_blood_rate` (log1p scale) from all features including `tf_primary_rate`, mirroring notebook 15_1. No outlier filter is applied (consistent with the reference analysis).

```{r btf-data}
reg_data_btf <- reg_data_raw %>%
  drop_na() %>%
  mutate(
    tf_primary_rate         = log1p(tf_primary_rate),
    tf_blood_rate           = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM               = log1p(TERT_FPKM),
    alt_status              = as.numeric(alt_status == 'ALT-high')
  ) %>%
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)

x_btf_r <- reg_data_btf %>% select(-tf_blood_rate)
y_btf_r  <- reg_data_btf$tf_blood_rate
```

## CV results (bTF)

As for tTF, TabPFN regression runs on carmela via `scripts/tabpfn_cv_oof.py --task regression`; the cohort + shared folds are exported by notebook 21 and the out-of-fold predictions are scored here with `reg_fold_metrics()`.

```{r btf-cv-results}
oof_btf <- here('outputs', 'tabpfn_reg_btf_oof.csv')
if (!file.exists(oof_btf)) {
  cat('TabPFN bTF out-of-fold predictions not found.\n',
      'Run (on carmela): python tabpfn_cv_oof.py --task regression',
      '--input-dir cmp_reg_btf_cv --output outputs/tabpfn_reg_btf_oof.csv\n')
} else {
  btf_metrics <- readr::read_csv(oof_btf, show_col_types = FALSE) %>%
    group_by(rep, fold) %>%
    group_modify(~ reg_fold_metrics(.x$yhat, .x$y)) %>%
    ungroup()
  btf_metrics %>%
    summarise(across(c(rmse, mae, spearman),
                     list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)),
                     .names = '{.col}_{.fn}')) %>%
    pivot_longer(everything(), names_to = c('metric', '.value'),
                 names_sep = '_(?=[^_]+$)') %>%
    mutate(summary = paste0(round(mean, 4), ' ± ', round(sd, 4))) %>%
    select(metric, summary)
}
```

## Interpretability (bTF — out-of-fold SHAP, HPC)

Out-of-fold SHAP over a single stratified 5-fold split (stratified on `alt_status`, `set.seed(42)`), mirroring the classification `full2` flow. Each sample is explained by the `TabPFNRegressor` that did not train on it; values are in the target (log1p) space.

```{r btf-export-shap-cv}
source(here('scripts', 'make_folds.R'))

# meta + stratification source, row-aligned with reg_data_btf (same drop_na, no filter)
reg_btf_filt <- reg_data_raw %>% drop_na()
stopifnot(nrow(reg_btf_filt) == nrow(reg_data_btf))

# SHAP uses one 5-fold partition = repeat 1 of the shared CV folds
folds_btf <- make_folds(factor(reg_data_btf$alt_status), k = 5, n_repeats = 5, seed = 21)[, 1]

shap_btf_in <- here('outputs', 'shap_input_btf_cv')
dir.create(shap_btf_in, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(as.data.frame(x_btf_r),          file.path(shap_btf_in, 'X_full.csv'))
readr::write_csv(data.frame(y = y_btf_r),         file.path(shap_btf_in, 'y_full.csv'))
readr::write_csv(data.frame(fold = folds_btf),    file.path(shap_btf_in, 'folds.csv'))
readr::write_csv(reg_btf_filt %>%
                   transmute(donor_id, cancer_type, cancer_group, alt_status),
                 file.path(shap_btf_in, 'meta.csv'))

cat(sprintf("Exported %d samples x %d features to %s\n",
            nrow(x_btf_r), ncol(x_btf_r), shap_btf_in))
table(folds_btf, reg_btf_filt$alt_status)
```

```{r btf-shap-load}
source(here('scripts', 'reconstruct_shapviz.R'))

shap_btf_imp_dir <- here('outputs', 'shap_output_btf_imp')
shap_btf_ready   <- all(file.exists(
  file.path(shap_btf_imp_dir, c('shap_values.csv', 'X_test_explained.csv', 'shap_baseline.txt'))
))

if (!shap_btf_ready) {
  cat("OOF SHAP results for bTF regression not yet available.\n")
  cat("Run on the GPU box (carmela), then sync outputs/ back:\n")
  cat("  bash scripts/run_shap_reg.sh\n")
} else {
  sv_btf_imp <- load_shap(shap_btf_imp_dir)                          # target (log1p) space
  sv_btf_rec <- load_shap(here('outputs', 'shap_output_btf_rec'))    # recontext target space
  meta_btf   <- load_meta(shap_btf_imp_dir)

  # rank-average normalization (same as the ALT-status SHAP plots): beeswarm
  # colors and dependence axes use per-feature rank, robust to outliers/scale.
  sv_btf_imp_rank <- recolor_by_rank(sv_btf_imp)
  sv_btf_rec_rank <- recolor_by_rank(sv_btf_rec)

  # --- imputation (additive, target log1p space) ---
  print(sv_importance(sv_btf_imp_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8)+ theme_bw())
  
  print(sv_dependence_shape(sv_btf_imp_rank, v = 'tf_primary_rate', color_var = 'telomere_content_log2', group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_btf_imp_rank, v = 'tf_primary_rate', group = meta_btf$cancer_type, top_n = 20, shape_group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_btf_imp_rank, v = 'tf_primary_rate', group = meta_btf$cancer_group, shape_group = meta_btf$alt_status) + theme_bw())

  print(sv_dependence_shape(sv_btf_imp_rank, v = 'telomere_insertion_rate', color_var = 'tf_primary_rate', group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_btf_imp_rank, v = 'telomere_insertion_rate', group = meta_btf$cancer_type, top_n = 20, shape_group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_btf_imp_rank, v = 'telomere_insertion_rate', group = meta_btf$cancer_group, shape_group = meta_btf$alt_status) + theme_bw())

  print(sv_dependence_shape(sv_btf_imp_rank, v = 'TERT_FPKM', color_var = 'telomere_content_log2', group = meta_btf$alt_status) + theme_bw())
  
  print(sv_dependence_shape(sv_btf_rec_rank, v = 'telomere_content_log2', color_var = 'tf_primary_rate', group = meta_btf$alt_status) + theme_bw())

  # --- recontextualization (native target space cross-check) ---
  print(sv_importance(sv_btf_rec_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8) + theme_bw())
  
  print(sv_dependence_shape(sv_btf_rec_rank, v = 'tf_primary_rate', color_var = 'telomere_content_log2', group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_btf_rec_rank, v = 'tf_primary_rate', group = meta_btf$cancer_type, top_n = 20, shape_group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_btf_rec_rank, v = 'tf_primary_rate', group = meta_btf$cancer_group, shape_group = meta_btf$alt_status) + theme_bw())

  print(sv_dependence_shape(sv_btf_rec_rank, v = 'telomere_insertion_rate', color_var = 'tf_primary_rate', group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_top(sv_btf_rec_rank, v = 'telomere_insertion_rate', group = meta_btf$cancer_type, top_n = 20, shape_group = meta_btf$alt_status) + theme_bw())
  print(sv_dependence_mes(sv_btf_rec_rank, v = 'telomere_insertion_rate', group = meta_btf$cancer_group, shape_group = meta_btf$alt_status) + theme_bw())

  print(sv_dependence_shape(sv_btf_rec_rank, v = 'TERT_FPKM', color_var = 'telomere_content_log2', group = meta_btf$alt_status) + theme_bw())
  
  print(sv_dependence_shape(sv_btf_rec_rank, v = 'telomere_content_log2', color_var = 'tf_primary_rate', group = meta_btf$alt_status) + theme_bw())
}
```
