---
title: "10_TabPFN"
author: "Andrea Polazzi"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

# Setup

```{r setup}
library(reticulate, quietly = TRUE)
tabpfn_venv <- here::here("analysis", ".venv-tabpfn")
use_virtualenv(tabpfn_venv, required = TRUE)
py_config()

api_token <- Sys.getenv("TABPFN_TOKEN")

library(readxl, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(pROC, quietly = TRUE)
library(caret, quietly = TRUE)
library(here)

source(here('scripts', 'clean_columns.R'))

cache_file <- here('outputs', 'tabpfn_cv_results.rds')
run_tabpfn <- !file.exists(cache_file)

data <- read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))

features14 <- c(
  "tf_primary_rate", "tf_blood_rate", "telomere_insertion_rate",
  "telomere_content_log2", "TERT_FPKM",
  "ATAGGG_singleton_dist", "CTAGGG_singleton_dist", "GTAGGG_singleton_dist",
  "TAAGGG_singleton_dist", "TCAGGG_singleton_dist", "TGAGGG_singleton_dist",
  "TTCGGG_singleton_dist", "TTGGGG_singleton_dist", "TTTGGG_singleton_dist"
)

model_full <- data %>%
  select(all_of(features14), alt_status, donor_id, cancer_type, cancer_group) %>%
  filter(complete.cases(across(all_of(features14)))) %>%
  mutate(alt_status = factor(alt_status, levels = c("ALT-low", "ALT-high")))

model_data <- model_full %>% select(all_of(features14), alt_status)

summary(model_data)
```

# Data preparation for TabPFN (single split)

```{r split}
set.seed(21)
train_index <- sample(seq_len(nrow(model_data)), size = 0.7 * nrow(model_data))
train <- model_data[train_index, ]
test  <- model_data[-train_index, ]

y_train_r <- as.integer(train$alt_status == "ALT-high")
y_test_r  <- as.integer(test$alt_status  == "ALT-high")

x_train_r <- train %>% select(-alt_status)
x_test_r  <- test  %>% select(-alt_status)

x_all_r <- model_data %>% select(-alt_status)
y_all_r  <- as.integer(model_data$alt_status == "ALT-high")
```

Here we have split the data in training (70%) and test (30%), encoding `alt_status` as 0 (ALT-low) and 1 (ALT-high).

# Export for out-of-fold SHAP (5-fold CV)

```{r export-shap-cv}
library(caret)

# one stratified 5-fold split on alt_status (mirrors notebook 02)
set.seed(42)
fold_list <- createFolds(model_data$alt_status, k = 5)
folds_vec <- integer(nrow(model_data))
for (i in seq_along(fold_list)) folds_vec[fold_list[[i]]] <- i

shap_cv_dir <- here('outputs', 'shap_input_cv')
dir.create(shap_cv_dir, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(model_data %>% select(all_of(features14)),
                 file.path(shap_cv_dir, 'X_full.csv'))
readr::write_csv(data.frame(y = as.integer(model_data$alt_status == "ALT-high")),
                 file.path(shap_cv_dir, 'y_full.csv'))
readr::write_csv(data.frame(fold = folds_vec),
                 file.path(shap_cv_dir, 'folds.csv'))
readr::write_csv(model_full %>%
                   transmute(donor_id, cancer_type, cancer_group, alt_status),
                 file.path(shap_cv_dir, 'meta.csv'))

cat(sprintf("Exported %d samples x %d features to %s\n",
            nrow(model_data), length(features14), shap_cv_dir))
table(folds_vec, model_data$alt_status)
```

# Training

```{python tabpfn}
import numpy as np
import tabpfn_client
from tabpfn_client import TabPFNClassifier

tabpfn_client.set_access_token(r.api_token)

X_train = r.x_train_r.to_numpy().astype(float)
X_test  = r.x_test_r.to_numpy().astype(float)
y_train = np.array(r.y_train_r, dtype=int)

clf = TabPFNClassifier()
clf.fit(X_train, y_train)

tabpfn_probs = clf.predict_proba(X_test)[:, 1]
```

# Performance evaluation

We evaluate the performance of the TabPFN classifier with the same metrics used for the other models.

```{r evaluate}
tabpfn_prob <- py$tabpfn_probs
y_test <- y_test_r

roc_tabpfn <- roc(
  response = factor(y_test, levels = c(0, 1), labels = c("ALT-low", "ALT-high")),
  predictor = tabpfn_prob,
  levels = c("ALT-low", "ALT-high"),
  direction = "<"
)

plot(roc_tabpfn, main = "ROC curve - TabPFN (all features)")
auc(roc_tabpfn)
```

```{r confusion-matrix}
pred_class <- factor(
  ifelse(tabpfn_prob >= 0.5, "ALT-high", "ALT-low"),
  levels = c("ALT-low", "ALT-high")
)
true_class <- factor(
  ifelse(y_test == 1, "ALT-high", "ALT-low"),
  levels = c("ALT-low", "ALT-high")
)

confusionMatrix(pred_class, true_class, positive = "ALT-high")
```

These **preliminary results** obtained on a first split of the data are already **promising**: the model achieved an **AUC = 0.9942.**

The confusion matrix, although being calculated on a single split, shows a good behavior, concordant with previous results: a **good, significant accuracy (0.988)**, **perfect specificity (1.000)** and **high sensitivity (0.800)**, that together yield a **good balanced accuracy (0.900)**.

These preliminary results on a single random split are very strong, but they should be interpreted cautiously until confirmed by repeated cross-validation.

# Stratified CV

To have a fair comparison between Log. Reg. and RF models, we have used the same **5-fold stratified CV with 5 repeats**.

Here are reported the results and performance evaluation metrics:

```{r cv-folds}
set.seed(21)

cv_data <- model_data
k <- 5
n_repeats <- 5

cv_splits <- list()
split_id <- 1

for (r in 1:n_repeats) {
  fold_id <- integer(nrow(cv_data))
  
  idx_low <- which(cv_data$alt_status == "ALT-low")
  idx_high <- which(cv_data$alt_status == "ALT-high")
  
  fold_id[idx_low] <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))
  
  for (fold in 1:k) {
    train_idx <- which(fold_id != fold)
    test_idx <- which(fold_id == fold)
    
    cv_splits[[split_id]] <- list(
      train = train_idx,
      test = test_idx,
      rep = r,
      fold = fold
    )
    
    split_id <- split_id + 1
  }
}
```

```{python tabpfn-cv, eval=run_tabpfn}
import numpy as np
from tabpfn_client import TabPFNClassifier
from sklearn.metrics import roc_auc_score

X_all = r.x_all_r.to_numpy().astype(float)
y_all = np.array(r.y_all_r, dtype=int)

cv_splits = r.cv_splits

auc_scores = []

per_fold_metrics = []

for split in cv_splits:
    train_idx = np.array(split["train"], dtype=int) - 1
    test_idx  = np.array(split["test"],  dtype=int) - 1

    clf = TabPFNClassifier()
    clf.fit(X_all[train_idx], y_all[train_idx])

    probs     = clf.predict_proba(X_all[test_idx])[:, 1]
    preds     = (probs >= 0.5).astype(int)
    true      = y_all[test_idx]

    auc  = roc_auc_score(true, probs)
    tp   = int(np.sum((preds == 1) & (true == 1)))
    fp   = int(np.sum((preds == 1) & (true == 0)))
    tn   = int(np.sum((preds == 0) & (true == 0)))
    fn   = int(np.sum((preds == 0) & (true == 1)))

    sens     = tp / (tp + fn)         if (tp + fn) > 0         else float("nan")
    spec     = tn / (tn + fp)         if (tn + fp) > 0         else float("nan")
    prec     = tp / (tp + fp)         if (tp + fp) > 0         else float("nan")
    recall   = sens
    f1       = 2*tp / (2*tp + fp + fn) if (2*tp + fp + fn) > 0 else float("nan")
    bal_acc  = 0.5 * (sens + spec)    if not (np.isnan(sens) or np.isnan(spec)) else float("nan")

    per_fold_metrics.append({
        "auc": auc, "tp": tp, "fp": fp, "tn": tn, "fn": fn,
        "sensitivity": sens, "specificity": spec,
        "precision": prec, "recall": recall,
        "f1": f1, "balanced_accuracy": bal_acc
    })
```

```{r load-or-save-tabpfn}
if (file.exists(cache_file)) {
  results <- readRDS(cache_file)
  message("Loaded cached TabPFN results")
} else {
  results <- list(
    per_fold_metrics = py$per_fold_metrics %>%
      bind_rows() %>%
      mutate(across(everything(), as.numeric)),
    seed = 21,
    k = k,
    n_repeats = n_repeats
  )
  saveRDS(results, cache_file)
  message("Computed TabPFN results and saved")
}

metrics_df <- results$per_fold_metrics %>%
  mutate(
    precision         = if_else((tp + fp) > 0, tp / (tp + fp), NA_real_),
    recall            = if_else((tp + fn) > 0, tp / (tp + fn), NA_real_),
    f1                = if_else((2*tp + fp + fn) > 0, 2*tp / (2*tp + fp + fn), NA_real_),
    balanced_accuracy = 0.5 * (sensitivity + specificity)
  )
```

```{r cv-results}
metrics_df <- metrics_df %>% mutate(model = "TabPFN")

# aggregated confusion matrix (sum over all folds, for reference)
agg_tp <- sum(metrics_df$tp)
agg_fp <- sum(metrics_df$fp)
agg_tn <- sum(metrics_df$tn)
agg_fn <- sum(metrics_df$fn)

cat("=== Aggregated confusion matrix (sum over all CV folds) ===\n")
cm_agg <- matrix(c(agg_tn, agg_fn,
                   agg_fp, agg_tp),
                 nrow = 2, byrow = TRUE,
                 dimnames = list(
                   Predicted = c("ALT-low", "ALT-high"),
                   Reference = c("ALT-low", "ALT-high")
                 ))
print(cm_agg)

cat("\n=== Per-fold metrics (mean ± SD across 25 folds) ===\n")
tabpfn_summary <- metrics_df %>%
  summarise(
    across(
      c(auc, sensitivity, specificity, precision, recall, f1, balanced_accuracy),
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)),
      .names = "{.col}_{.fn}"
    )
  )

tabpfn_summary %>%
  pivot_longer(
    everything(),
    names_to  = c("metric", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  mutate(summary = paste0(round(mean, 4), " ± ", round(sd, 4))) %>%
  select(metric, summary) %>%
  print()
```

# Interpretability

```{r export-data, eval=FALSE}
dir.create(here('outputs', 'shap_input'), recursive = TRUE, showWarnings = FALSE)

readr::write_csv(as.data.frame(x_train_r), here('outputs', 'shap_input', 'X_train.csv'))
n_shap <- min(20, nrow(x_test_r))
readr::write_csv(as.data.frame(x_test_r[1:n_shap, , drop = FALSE]),
                 here('outputs', 'shap_input', 'X_test_small.csv'))    # we use a smaller test set otherwise it takes too much time
readr::write_csv(data.frame(y_train = y_train_r), here('outputs', 'shap_input', 'y_train.csv'))
```

```{r run-python-script, eval=FALSE}
cmd <- paste(
  shQuote(reticulate::py_config()$python),
  shQuote(here("scripts", "run_tabpfn_shap.py"))
)

Sys.setenv(TABPFN_TOKEN = api_token)

system(cmd, intern = FALSE, ignore.stdout = FALSE, ignore.stderr = FALSE)
```

## SHAP plots

```{r shap-images, out.width='85%'}
shap_files <- here(
  "analysis",
  "figures",
  c("tabpfn_shap1.png", "tabpfn_shap2.png", "tabpfn_shap3.png")
)

existing_files <- shap_files[file.exists(shap_files)]

if (length(existing_files) == 0) {
  cat("No SHAP images found in analysis/figures/.")
} else {
  knitr::include_graphics(existing_files)
}
```

```{r tabpfn-shap-hpc-test}
source(here('scripts', 'reconstruct_shapviz.R'))

# per-feature rank, scaled to [0,1], NA-safe
rank01 <- function(x) {
  n <- sum(!is.na(x))
  (rank(x, na.last = "keep", ties.method = "average") - 1) / (n - 1)
}

recolor_by_rank <- function(sv) {
  X_rank <- as.data.frame(lapply(get_feature_values(sv), rank01))
  shapviz(object   = get_shap_values(sv),
          X        = X_rank,
          baseline = get_baseline(sv))
}

sv_imp2_rank <- recolor_by_rank(sv_imp2)
sv_rec2_rank <- recolor_by_rank(sv_rec2)

# Imputation, single split train/test
sv_importance(sv_imp, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8)
sv_dependence(sv_imp, v = 'telomere_insertion_rate', color_var = 'tf_rate')
sv_dependence(sv_imp, v = "telomere_content_log2", color_var = "TERT_FPKM") + theme_bw()
sv_dependence(sv_imp, v = "tf_rate",         color_var = "telomere_content_log2") + theme_bw()
sv_dependence(sv_imp, v = "TERT_FPKM",             color_var = "TTTGGG_singleton_dist") + theme_bw()

# Imputation, 5-fold CV
sv_importance(sv_imp2_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8)
sv_dependence(sv_imp2_rank, v = 'telomere_insertion_rate', color_var = 'tf_primary_rate')
sv_dependence(sv_imp2_rank, v = "telomere_content_log2", color_var = "TERT_FPKM") + theme_bw()
sv_dependence(sv_imp2_rank, v = "tf_primary_rate",         color_var = "telomere_content_log2") + theme_bw()
sv_dependence(sv_imp2_rank, v = "TERT_FPKM",             color_var = "TTTGGG_singleton_dist") + theme_bw()

# Recontextualization, single split train/test
sv_importance(sv_rec, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8)
sv_dependence(sv_rec, v = 'telomere_insertion_rate', color_var = 'tf_rate')
sv_dependence(sv_rec, v = "telomere_content_log2", color_var = "TERT_FPKM") + theme_bw()
sv_dependence(sv_rec, v = "tf_rate",         color_var = "telomere_content_log2") + theme_bw()
sv_dependence(sv_rec, v = "TERT_FPKM",             color_var = "TTTGGG_singleton_dist") + theme_bw()

# Imputation, 5-fold CV
sv_importance(sv_rec2_rank, kind = 'beeswarm', bee_adjust = 10, alpha = 0.8)
sv_dependence(sv_rec2_rank, v = 'telomere_insertion_rate', color_var = 'tf_primary_rate')
sv_dependence(sv_rec2_rank, v = "telomere_content_log2", color_var = "TERT_FPKM") + theme_bw()
sv_dependence(sv_rec2_rank, v = "tf_primary_rate",         color_var = "telomere_content_log2") + theme_bw()
sv_dependence(sv_rec2_rank, v = "TERT_FPKM",             color_var = "TTTGGG_singleton_dist") + theme_bw()
```

------------------------------------------------------------------------

# TabPFN Regression

We apply TabPFN to two regression tasks that mirror notebooks 13_1 (`tf_primary_rate`) and 15_1 (`tf_blood_rate`). The same 5-fold × 5-repeat stratified CV and the same metrics (RMSE, MAE, Spearman ρ) are used to allow direct comparison with LM, LASSO, and RF.

## Data setup

```{r reg-setup}
reg_data_raw <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
```

------------------------------------------------------------------------

# tTF regression

Predicting `tf_primary_rate` (log1p scale) from all features including `tf_blood_rate`, mirroring notebook 13_1. The strong outlier (`tf_primary_rate >= 10`) is removed as in the reference analysis.

```{r ttf-data}
reg_data_ttf <- reg_data_raw %>%
  drop_na() %>%
  filter(tf_primary_rate < 10) %>%
  mutate(
    tf_primary_rate         = log1p(tf_primary_rate),
    tf_blood_rate           = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM               = log1p(TERT_FPKM),
    alt_status              = as.numeric(alt_status == "ALT-high")
  ) %>%
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)

cache_ttf      <- here('outputs', 'tabpfn_reg_ttf_cv.rds')
run_tabpfn_ttf <- !file.exists(cache_ttf)

x_ttf_r <- reg_data_ttf %>% select(-tf_primary_rate)
y_ttf_r  <- reg_data_ttf$tf_primary_rate
```

## CV splits (tTF)

```{r ttf-cv-splits}
set.seed(21)
ttf_cv_splits <- list()
split_id <- 1

for (r in 1:n_repeats) {
  fold_id  <- integer(nrow(reg_data_ttf))
  idx_low  <- which(reg_data_ttf$alt_status == 0)
  idx_high <- which(reg_data_ttf$alt_status == 1)
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))

  for (fold in 1:k) {
    ttf_cv_splits[[split_id]] <- list(
      train = which(fold_id != fold),
      test  = which(fold_id == fold),
      rep   = r,
      fold  = fold
    )
    split_id <- split_id + 1
  }
}
```

```{python tabpfn-reg-ttf, eval=run_tabpfn_ttf}
import numpy as np
from tabpfn_client import TabPFNRegressor
from sklearn.metrics import mean_squared_error, mean_absolute_error
from scipy.stats import spearmanr

X = r.x_ttf_r.to_numpy().astype(float)
y = np.array(r.y_ttf_r, dtype=float)

per_fold_ttf = []

for split in r.ttf_cv_splits:
    tr = np.array(split["train"], dtype=int) - 1
    te = np.array(split["test"],  dtype=int) - 1

    reg = TabPFNRegressor()
    reg.fit(X[tr], y[tr])
    preds = reg.predict(X[te])
    truth = y[te]

    per_fold_ttf.append({
        "rmse":     float(np.sqrt(mean_squared_error(truth, preds))),
        "mae":      float(mean_absolute_error(truth, preds)),
        "spearman": float(spearmanr(truth, preds)[0]),
        "rep":      int(split["rep"]),
        "fold":     int(split["fold"])
    })
```

```{r ttf-load-or-save}
if (file.exists(cache_ttf)) {
  ttf_metrics <- readRDS(cache_ttf)
  message("Loaded cached tTF results")
} else {
  ttf_metrics <- py$per_fold_ttf %>%
    bind_rows() %>%
    mutate(across(everything(), as.numeric), model = "TabPFN")
  saveRDS(ttf_metrics, cache_ttf)
  message("Saved tTF results")
}
```

## CV results (tTF)

```{r ttf-cv-results}
ttf_metrics %>%
  summarise(
    across(
      c(rmse, mae, spearman),
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(),
               names_to  = c("metric", ".value"),
               names_sep = "_(?=[^_]+$)") %>%
  mutate(summary = paste0(round(mean, 4), " ± ", round(sd, 4))) %>%
  select(metric, summary)

ttf_metrics %>%
  pivot_longer(c(rmse, mae, spearman), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "TabPFN — tTF CV metrics (25 folds)", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Interpretability (tTF — HPC)

```{r ttf-export-shap, eval=FALSE}
dir.create(here('outputs', 'shap_input_ttf'), recursive = TRUE, showWarnings = FALSE)
readr::write_csv(as.data.frame(x_ttf_r),
                 here('outputs', 'shap_input_ttf', 'X_train.csv'))
readr::write_csv(as.data.frame(x_ttf_r[seq_len(min(50, nrow(x_ttf_r))), ]),
                 here('outputs', 'shap_input_ttf', 'X_test.csv'))
readr::write_csv(data.frame(y_train = y_ttf_r),
                 here('outputs', 'shap_input_ttf', 'y_train.csv'))
```

```{r ttf-shap-load}
shap_ttf_dir   <- here('outputs', 'shap_output_ttf')
shap_ttf_ready <- all(file.exists(
  file.path(shap_ttf_dir, c('shap_values.csv', 'X_test_explained.csv', 'shap_baseline.txt'))
))

if (!shap_ttf_ready) {
  cat("SHAP results for tTF regression not yet available.\n")
  cat("Run on HPC:\n")
  cat("  python scripts/SHAP_tabpfn_local.py \\\n")
  cat("    --input-dir  outputs/shap_input_ttf \\\n")
  cat("    --output-dir outputs/shap_output_ttf\n")
} else {
  sv_ttf <- shapviz(
    object   = as.matrix(read.csv(file.path(shap_ttf_dir, 'shap_values.csv'))),
    X        = read.csv(file.path(shap_ttf_dir, 'X_test_explained.csv')),
    baseline = as.numeric(readLines(file.path(shap_ttf_dir, 'shap_baseline.txt')))
  )
  sv_importance(sv_ttf, kind = 'both')
  sv_waterfall(sv_ttf, row_id = 1)
}
```

------------------------------------------------------------------------

# bTF regression

Predicting `tf_blood_rate` (log1p scale) from all features including `tf_primary_rate`, mirroring notebook 15_1. No outlier filter is applied (consistent with the reference analysis).

```{r btf-data}
reg_data_btf <- reg_data_raw %>%
  drop_na() %>%
  mutate(
    tf_primary_rate         = log1p(tf_primary_rate),
    tf_blood_rate           = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM               = log1p(TERT_FPKM),
    alt_status              = as.numeric(alt_status == "ALT-high")
  ) %>%
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)

cache_btf      <- here('outputs', 'tabpfn_reg_btf_cv.rds')
run_tabpfn_btf <- !file.exists(cache_btf)

x_btf_r <- reg_data_btf %>% select(-tf_blood_rate)
y_btf_r  <- reg_data_btf$tf_blood_rate
```

## CV splits (bTF)

```{r btf-cv-splits}
set.seed(21)
btf_cv_splits <- list()
split_id <- 1

for (r in 1:n_repeats) {
  fold_id  <- integer(nrow(reg_data_btf))
  idx_low  <- which(reg_data_btf$alt_status == 0)
  idx_high <- which(reg_data_btf$alt_status == 1)
  fold_id[idx_low]  <- sample(rep(1:k, length.out = length(idx_low)))
  fold_id[idx_high] <- sample(rep(1:k, length.out = length(idx_high)))

  for (fold in 1:k) {
    btf_cv_splits[[split_id]] <- list(
      train = which(fold_id != fold),
      test  = which(fold_id == fold),
      rep   = r,
      fold  = fold
    )
    split_id <- split_id + 1
  }
}
```

```{python tabpfn-reg-btf, eval=run_tabpfn_btf}
import numpy as np
from tabpfn_client import TabPFNRegressor
from sklearn.metrics import mean_squared_error, mean_absolute_error
from scipy.stats import spearmanr

X = r.x_btf_r.to_numpy().astype(float)
y = np.array(r.y_btf_r, dtype=float)

per_fold_btf = []

for split in r.btf_cv_splits:
    tr = np.array(split["train"], dtype=int) - 1
    te = np.array(split["test"],  dtype=int) - 1

    reg = TabPFNRegressor()
    reg.fit(X[tr], y[tr])
    preds = reg.predict(X[te])
    truth = y[te]

    per_fold_btf.append({
        "rmse":     float(np.sqrt(mean_squared_error(truth, preds))),
        "mae":      float(mean_absolute_error(truth, preds)),
        "spearman": float(spearmanr(truth, preds)[0]),
        "rep":      int(split["rep"]),
        "fold":     int(split["fold"])
    })
```

```{r btf-load-or-save}
if (file.exists(cache_btf)) {
  btf_metrics <- readRDS(cache_btf)
  message("Loaded cached bTF results")
} else {
  btf_metrics <- py$per_fold_btf %>%
    bind_rows() %>%
    mutate(across(everything(), as.numeric), model = "TabPFN")
  saveRDS(btf_metrics, cache_btf)
  message("Saved bTF results")
}
```

## CV results (bTF)

```{r btf-cv-results}
btf_metrics %>%
  summarise(
    across(
      c(rmse, mae, spearman),
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(),
               names_to  = c("metric", ".value"),
               names_sep = "_(?=[^_]+$)") %>%
  mutate(summary = paste0(round(mean, 4), " ± ", round(sd, 4))) %>%
  select(metric, summary)

btf_metrics %>%
  pivot_longer(c(rmse, mae, spearman), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "TabPFN — bTF CV metrics (25 folds)", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Interpretability (bTF — HPC)

```{r btf-export-shap, eval=FALSE}
dir.create(here('outputs', 'shap_input_btf'), recursive = TRUE, showWarnings = FALSE)
readr::write_csv(as.data.frame(x_btf_r),
                 here('outputs', 'shap_input_btf', 'X_train.csv'))
readr::write_csv(as.data.frame(x_btf_r[seq_len(min(50, nrow(x_btf_r))), ]),
                 here('outputs', 'shap_input_btf', 'X_test.csv'))
readr::write_csv(data.frame(y_train = y_btf_r),
                 here('outputs', 'shap_input_btf', 'y_train.csv'))
```

```{r btf-shap-load}
shap_btf_dir   <- here('outputs', 'shap_output_btf')
shap_btf_ready <- all(file.exists(
  file.path(shap_btf_dir, c('shap_values.csv', 'X_test_explained.csv', 'shap_baseline.txt'))
))

if (!shap_btf_ready) {
  cat("SHAP results for bTF regression not yet available.\n")
  cat("Run on HPC:\n")
  cat("  python scripts/SHAP_tabpfn_local.py \\\n")
  cat("    --input-dir  outputs/shap_input_btf \\\n")
  cat("    --output-dir outputs/shap_output_btf\n")
} else {
  sv_btf <- shapviz(
    object   = as.matrix(read.csv(file.path(shap_btf_dir, 'shap_values.csv'))),
    X        = read.csv(file.path(shap_btf_dir, 'X_test_explained.csv')),
    baseline = as.numeric(readLines(file.path(shap_btf_dir, 'shap_baseline.txt')))
  )
  sv_importance(sv_btf, kind = 'both')
  sv_waterfall(sv_btf, row_id = 1)
}
```
