---
title: "13_1_tTF_regression_+blood"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

**Repeating the same pipeline from the previous document, adding blood TF rate.**

# Setup

We remove missing values and a strong outlier (otherwise Gamma-based model won't converge).

```{r setup}
library(tidyverse)
library(glmnet)
library(here)
library(randomForest)
library(caret)
library(pROC)
library(broom)

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
clean_data <- data %>% 
  drop_na() %>% 
  filter(tf_primary_rate<10)
cancer_type_vector <- clean_data$cancer_type
```

## Transformations

Since we are in a regression task, it is better to have a distribution for our response variable as symmetric as possible. We know it is currently strongly skewed, thus we can apply a log1p transformation to deal with its asymmetry and its zero values. We can do the same also for similarly distributed features, such as telomere insertion.

The log1p transformation reduces the influence of the outliers for linear models, but it doesn't alter much the zero-inflated structure of the distribution.

```{r log1p-transformation}
clean_data_orig <- clean_data %>% 
  mutate(
    tf_blood_rate = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM = log1p(TERT_FPKM),
    alt_status = factor(
      alt_status,
      levels = c("ALT-low", "ALT-high")
    )
  ) %>%  
  select(-donor_id, -cancer_type, -Specimen.Type.Summary)

clean_data <- clean_data %>% 
  mutate(
    tf_primary_rate = log1p(tf_primary_rate),
    tf_blood_rate = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM = log1p(TERT_FPKM),
    alt_status = factor(
      alt_status,
      levels = c("ALT-low", "ALT-high")
    )
  ) %>%  
  select(-donor_id, -cancer_type, -Specimen.Type.Summary)
```

# Split the dataset

We split the dataset with a 80/20 split in training set and test set. We won't touch the latter until final performance estimation, to avoid data leakage.

```{r train-test}
set.seed(21)
train_idx <- sample(nrow(clean_data), 0.8*nrow(clean_data))

train <- clean_data[train_idx,]
test <- clean_data[-train_idx,]

train_orig <- clean_data_orig[train_idx,]
test_orig <- clean_data_orig[-train_idx,]
```

# Metrics functions

We define functions to help us in the evaluation of the performances of the models. In particular we will use RMSE and R squared.

```{r metrics}
rmse_fn <- function(actual, predicted) sqrt(mean((actual-predicted)^2))
mae_fn  <- function(actual, predicted) mean(abs(actual - predicted))
```

------------------------------------------------------------------------

# Manual Cross-Validation

We use a 5-fold cross-validation on the training set. We do it manually to have to possibility to further extend to other metrics, methods, or whatever.

```{r cv-setup}
k <- 5
n_repeats <- 5
set.seed(21)

cv_results <- tibble(fold = integer(), model = character(), rmse = double(), mae = double())
lasso_coef_list <- list()
rf_importance_list <- list()
idx <- 1
```

```{r cv-loop}
for (rep in 1:n_repeats){
  set.seed(21 + rep)
  fold_list <- createFolds(factor(train$alt_status), k = k, returnTrain = FALSE)
  folds <- integer(nrow(train))
  for (i in seq_along(fold_list)) folds[fold_list[[i]]] <- i
  for (i in 1:k){
    fold_train <- train[folds!=i,]
    fold_test <- train[folds==i,]
    
    # Standardize the predictors
    cols_to_scale <- setdiff(names(fold_train), c("tf_primary_rate", "alt_status"))
    preproc <- preProcess(fold_train %>% select(all_of(cols_to_scale)),
                           method = c("center", "scale"))
    fold_train_s <- fold_train %>%
      mutate(predict(preproc, fold_train %>% select(all_of(cols_to_scale))))
    fold_test_s  <- fold_test %>%
      mutate(predict(preproc, fold_test %>% select(all_of(cols_to_scale))))
    
    # LM
    fit_lm <- lm(tf_primary_rate ~ ., data = fold_train_s)
    pred_lm <- predict(fit_lm, newdata = fold_test_s)
  
    # LASSO
    x_fold_train <- model.matrix(tf_primary_rate ~ ., fold_train_s)[,-1]
    y_fold_train <- fold_train_s$tf_primary_rate
    x_fold_test <- model.matrix(tf_primary_rate ~ ., fold_test_s)[,-1]
    
    cv_l <- cv.glmnet(x_fold_train, y_fold_train, alpha=1, nfolds=5)
    fit_lasso <- glmnet(x_fold_train, y_fold_train, alpha=1, lambda = cv_l$lambda.min)
    pred_lasso <- as.vector(predict(fit_lasso, newx = x_fold_test))
    
    ## LASSO coefficients for this fold
    lasso_coef_list[[idx]] <- coef(fit_lasso) %>% 
      as.matrix() %>% 
      as.data.frame() %>% 
      rownames_to_column('feature') %>% 
      rename(coef = s0) %>% 
      mutate(rep = rep, fold = i)
    
    
    # Random Forest
    fit_rf <- randomForest(tf_primary_rate ~ ., data = fold_train, importance = TRUE)
    pred_rf <- predict(fit_rf, newdata = fold_test)
    
    ## Importance of this fold
    rf_importance_list[[idx]] <- importance(fit_rf) %>% 
      as.data.frame() %>% 
      rownames_to_column('feature') %>% 
      mutate(rep = rep, fold = i)
    
    idx <- idx + 1
    
    # Metrics
    cv_results <- cv_results %>% bind_rows(
      tibble(rep = rep, fold = i, model = 'lm',
             rmse = rmse_fn(fold_test$tf_primary_rate, pred_lm),
             mae = mae_fn(fold_test$tf_primary_rate, pred_lm)),
      tibble(rep = rep, fold = i, model = 'lasso',
             rmse = rmse_fn(fold_test$tf_primary_rate, pred_lasso),
             mae = mae_fn(fold_test$tf_primary_rate, pred_lasso)),
      tibble(rep = rep, fold = i, model = 'rf',
             rmse = rmse_fn(fold_test$tf_primary_rate, pred_rf),
             mae = mae_fn(fold_test$tf_primary_rate, pred_rf))
    )
  }
}
```

# CV results

Averages and standard deviations on 5 folds.

```{r cv-performance-table}
cv_summary <- cv_results %>% 
  group_by(model) %>% 
  summarise(
    rmse_mean = mean(rmse), rmse_sd = sd(rmse),
    mae_mean = mean(mae), mae_sd = sd(mae),
    .groups = 'drop'
  ) %>% 
  arrange(rmse_mean)

cv_summary
```

```{r cv-performance-plot}
cv_results |>
  pivot_longer(c(rmse, mae), names_to = "metric", values_to = "value") |>
  mutate(model = fct_reorder(model, value, .fun = mean)) |>
  ggplot(aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  facet_wrap(~metric, scales = "free_y",
             labeller = labeller(metric = c(rmse = "RMSE", mae = "MAE"))) +
  labs(title = "Metrics distribution per fold (5-fold)",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Feature importance : LASSO

For each feature we register in how many folds is has been selected from Lasso and the mean/SD of the coefficient. A feature selected in all folds with a stable coefficient should be a strong enough signal. Otherwise, rarely selected features or unstable coefficients must be interpreted carefully.

```{r lasso-stability}
lasso_stability <- bind_rows(lasso_coef_list) %>% 
  filter(feature != '(Intercept)') %>% 
  group_by(feature) %>% 
  summarise(
    n_selected = sum(coef != 0),
    freq = n_selected / (n_repeats*k),
    coef_mean = mean(coef[coef != 0]),
    coef_sd = sd(coef[coef != 0]),
    .groups = 'drop'
  ) %>% 
  arrange(desc(freq), desc(abs(coef_mean)))
lasso_stability
```

```{r lasso-stability-plot}
plt_lasso_coef <- lasso_stability %>% 
                    filter(n_selected > 0) %>% 
                    mutate(feature = fct_reorder(feature, abs(coef_mean))) %>% 
                    ggplot(aes(x = coef_mean, y = feature, fill = freq))+
                    geom_col(alpha = 0.8)+
                    geom_errorbarh(aes(xmin = coef_mean - coef_sd,
                                       xmax = coef_mean + coef_sd),
                                   width = 0.25, color = 'grey30')+
                    scale_fill_gradient(low = 'grey90', high = 'steelblue')+
                    geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey50')+
                    labs(title = 'LASSO coefficients aggregated on 5 folds',
                         subtitle = 'Mean ± SD',
                         x = 'Mean coefficient', y = NULL)+
                    theme_bw()
plt_lasso_coef
```

## Feature importance: Random Forest

```{r rf-importance}
rf_importance_summary <- bind_rows(rf_importance_list) %>% 
  group_by(feature) %>% 
  summarise(
    inc_mse_mean = mean(`%IncMSE`),
    inc_mse_sd = sd(`%IncMSE`),
    .groups = 'drop'
    ) %>% 
  arrange(desc(inc_mse_mean))
rf_importance_summary
```

```{r rf-importance-plot}
plt_rf_importance <- rf_importance_summary %>% 
                        mutate(feature = fct_reorder(feature, inc_mse_mean)) %>% 
                        ggplot(aes(x = inc_mse_mean, y = feature))+
                        geom_col(fill='seagreen', alpha = 0.8) +
                        geom_errorbarh(aes(xmin = inc_mse_mean - inc_mse_sd,
                                           xmax = inc_mse_mean + inc_mse_sd),
                                       width = 0.3, color = 'grey30')+
                        labs(title = 'RF importance aggregated on 5 folds (%IncMSE)',
                             subtitle = 'Mean ± SD',
                             x = 'Mean %IncMSE', y = NULL) +
                        theme_bw()
plt_rf_importance
```

# Final evaluation on test set

Let's fit each model on the whole training set and evaluate on the test set.

```{r final-lm}
final_lm   <- lm(tf_primary_rate ~ ., data = train)
pred_lm_f  <- predict(final_lm, newdata = test)
```

```{r final-lasso}
x_train_f  <- model.matrix(tf_primary_rate ~ ., train)[, -1]
y_train_f  <- train$tf_primary_rate
x_test_f   <- model.matrix(tf_primary_rate ~ ., test)[,  -1]

set.seed(21)
cv_lasso_f    <- cv.glmnet(x_train_f, y_train_f, alpha = 1)
final_lasso   <- glmnet(x_train_f, y_train_f, alpha = 1,
                        lambda = cv_lasso_f$lambda.min)
pred_lasso_f  <- as.vector(predict(final_lasso, newx = x_test_f))
```

```{r final-rf}
set.seed(21)
final_rf  <- randomForest(tf_primary_rate ~ ., data = train, importance = TRUE)
pred_rf_f <- predict(final_rf, newdata = test)
```

```{r final-results}
tibble(
  model = c("LM", "LASSO", "Random Forest"),
  rmse  = c(rmse_fn(test$tf_primary_rate, pred_lm_f),
            rmse_fn(test$tf_primary_rate, pred_lasso_f),
            rmse_fn(test$tf_primary_rate, pred_rf_f)),
  mae   = c(mae_fn(test$tf_primary_rate, pred_lm_f),
            mae_fn(test$tf_primary_rate, pred_lasso_f),
            mae_fn(test$tf_primary_rate, pred_rf_f)),
  Spearman = c(
    cor(test$tf_primary_rate, pred_lm_f,    method = "spearman"),
    cor(test$tf_primary_rate, pred_lasso_f, method = "spearman"),
    cor(test$tf_primary_rate, pred_rf_f,    method = "spearman")
  )
) %>% 
  arrange(rmse)
```

## Quantification of the residuals

Beware that these results are in log1p scale!

```{r quantification-error}
df <- tibble(
  true = test$tf_primary_rate,
  pred = pred_rf_f,
  error = pred - true,
  abs_error = abs(error),
  perc_error = paste0(round(error/true, digits = 3) * 100,'%'),
  cancer_type = cancer_type_vector[-train_idx]
)
print(df, n=50)

df %>%
  summarise(
    mean_error = mean(error),
    sd_error = sd(error),
    mae = mean(abs_error),
    median_ae = median(abs_error),
    max_ae = max(abs_error),
    rmse = sqrt(mean(error^2))
  )

df %>%
  group_by(cancer_type) %>% 
  mutate(n_label = paste0(cancer_type, ' (n=', n(), ')')) %>% 
  ungroup() %>% 
  mutate(n_label = fct_reorder(n_label, abs_error, .fun = median)) %>%
  ggplot(aes(x = n_label, y = abs_error)) +
  geom_boxplot(outlier.shape = 21, fill = "steelblue", alpha = 0.6) +
  coord_flip() +
  labs(title = "Absolute error by cancer type (test set)",
       x = NULL, y = "Absolute error (log1p scale)") +
  theme_bw()
```

# Two-part model

```{r two-model}
k <- 5
n_repeats <- 5
presence_coef_list <- list()
intensity_coef_list <- list()
presence_results <- tibble()
idx <- 1

for(rep in 1:n_repeats){
  set.seed(21 + rep)
  fold_list <- createFolds(
    factor(train$alt_status),
    k = k,
    returnTrain = FALSE
  )
  folds <- integer(nrow(train))

  for(i in seq_along(fold_list)){
    folds[fold_list[[i]]] <- i
  }

  for(i in 1:k){
    fold_train <- train[folds != i, ]
    fold_test  <- train[folds == i, ]
    cols_to_scale <- setdiff(
      names(fold_train),
      c("tf_primary_rate", "alt_status")
    )
    preproc <- preProcess(
      fold_train %>% select(all_of(cols_to_scale)),
      method = c("center", "scale")
    )
    fold_train_s <- fold_train %>%
      mutate(
        predict(
          preproc,
          fold_train %>% select(all_of(cols_to_scale))
        )
      )
    fold_test_s <- fold_test %>%
      mutate(
        predict(
          preproc,
          fold_test %>% select(all_of(cols_to_scale))
        )
      )
    # PART 1: Presence
    m_presence_cv <- glm(
      I(tf_primary_rate > 0) ~ .,
      family = binomial(link = "logit"),
      data = fold_train_s
    )
    presence_coef_list[[idx]] <- tidy(m_presence_cv) %>%
      filter(term != "(Intercept)") %>%
      mutate(
        rep = rep,
        fold = i
      )
    pred_presence_cv <- predict(
      m_presence_cv,
      newdata = fold_test_s,
      type = "response"
    )
    presence_results <- bind_rows(
      presence_results,
      tibble(
        rep = rep,
        fold = i,
        truth = as.numeric(fold_test_s$tf_primary_rate > 0),
        prob = pred_presence_cv
      )
    )
    # PART 2: Intensity
    df_positive_cv <- fold_train_s %>%
      filter(tf_primary_rate > 0)
    m_intensity_cv <- glm(
      tf_primary_rate ~ .,
      family = Gamma(link = "log"),
      data = df_positive_cv
    )
    intensity_coef_list[[idx]] <- tidy(m_intensity_cv) %>%
      filter(term != "(Intercept)") %>%
      mutate(
        rep = rep,
        fold = i
      )
    idx <- idx + 1
  }
}
```

## Importance of the two models

```{r importance-two-models}
importance_cv <- bind_rows(
  bind_rows(presence_coef_list) %>%
    mutate(model_part = "Presence (P(TF>0))"),
  bind_rows(intensity_coef_list) %>%
    mutate(model_part = "Intensity (E[TF|TF>0])")
) %>%
  group_by(model_part, term) %>%
  summarise(
    coef_mean = mean(estimate),
    coef_sd = sd(estimate),
    p_median = median(p.value, na.rm = TRUE),
    freq_p05 = mean(p.value < 0.05, na.rm = TRUE),
    sign_stability =
      mean(sign(estimate) == sign(mean(estimate))),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      freq_p05 >= 0.9 ~ "***",
      freq_p05 >= 0.7 ~ "**",
      freq_p05 >= 0.5 ~ "*",
      TRUE ~ "ns"
    )
  )

importance_cv %>%
  mutate(term = fct_reorder(term, abs(coef_mean))) %>%
  ggplot(aes(x = coef_mean,
             y = term,
             fill = model_part)) +
  geom_col(alpha = 0.8) +
  geom_errorbarh(
    aes(
      xmin = coef_mean - coef_sd,
      xmax = coef_mean + coef_sd
    ),
    height = 0.2,
    color = "grey30"
  ) +
  geom_text(
    aes(label = significance),
    hjust = -0.3,
    vjust = -0.5,
    size = 3
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey50"
  ) +
  facet_wrap(~model_part, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Two-part model coefficient stability across CV folds",
    subtitle =
      "***: significant in ≥90% folds; **: ≥70%; *: ≥50%",
    x = "Mean coefficient across CV folds",
    y = NULL
  ) + 
  theme(legend.position = "none")
```

## Regression on test set

Here we do regression using the two-part model. We preprocess the training and test set with a standardization and, to have fair comparisons with other models, we apply the Duan smearing correction to expm1 re-transformation.

```{r two-part model regression}
# Preprocessing on training set
cols_to_scale_orig <- setdiff(names(train_orig), c("tf_primary_rate", "alt_status"))

preproc_orig <- preProcess(
  train_orig %>% select(all_of(cols_to_scale_orig)),
  method = c("center", "scale")
)

train_orig_s <- train_orig %>%
  mutate(predict(preproc_orig, train_orig %>% select(all_of(cols_to_scale_orig))))

test_orig_s <- test_orig %>%
  mutate(predict(preproc_orig, test_orig %>% select(all_of(cols_to_scale_orig))))

# Fit final models on the whole training set
m_presence_final <- glm(
  I(tf_primary_rate > 0) ~ .,
  family = binomial(link = "logit"),
  data = train_orig_s
)

m_intensity_final <- glm(
  tf_primary_rate ~ .,
  family = Gamma(link = "log"),
  data = train_orig_s %>% filter(tf_primary_rate > 0),
  control = glm.control(maxit = 200)
)

# Prediction on test set
pred_presence  <- predict(m_presence_final, newdata = test_orig_s, type = "response")
pred_intensity <- predict(m_intensity_final, newdata = test_orig_s, type = "response")
pred_combined  <- pred_presence * pred_intensity    # formula of the two-part model

true_orig <- test_orig$tf_primary_rate

# Correction of log1p retransformation of LM/LASSO/RF (smearing)
smear_lm    <- mean(exp(residuals(final_lm)))
smear_lasso <- mean(exp(y_train_f - as.vector(predict(final_lasso, newx = x_train_f))))
smear_rf    <- mean(exp(train$tf_primary_rate - predict(final_rf)))

pred_lm_orig    <- expm1(pred_lm_f)    * smear_lm
pred_lasso_orig <- expm1(pred_lasso_f) * smear_lasso
pred_rf_orig    <- expm1(pred_rf_f)    * smear_rf

tibble(
  model = c("LM", "LASSO", "Random Forest", "Two-part"),
  RMSE = c(
    rmse_fn(true_orig, pred_lm_orig),
    rmse_fn(true_orig, pred_lasso_orig),
    rmse_fn(true_orig, pred_rf_orig),
    rmse_fn(true_orig, pred_combined)
  ),
  MAE = c(
    mean(abs(true_orig - pred_lm_orig)),
    mean(abs(true_orig - pred_lasso_orig)),
    mean(abs(true_orig - pred_rf_orig)),
    mean(abs(true_orig - pred_combined))
  ),
  Spearman = c(
    cor(true_orig, pred_lm_orig,    method = "spearman"),
    cor(true_orig, pred_lasso_orig, method = "spearman"),
    cor(true_orig, pred_rf_orig,    method = "spearman"),
    cor(true_orig, pred_combined,   method = "spearman")
  )
) %>% arrange(RMSE)

```

## How is the PRESENCE model performing?

```{r presence-model-auc}
# AUC of presence model
presence_auc <- presence_results %>%
  group_by(rep, fold) %>%
  summarise(
    auc = as.numeric(pROC::roc(truth, prob, quiet = TRUE)$auc),
    .groups = "drop"
  )
presence_auc %>%
  summarise(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE)
  )
presence_auc %>%
  ggplot(aes(x = "Presence model", y = auc)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, width = 0.4) +
  geom_jitter(width = 0.08, alpha = 0.6) +
  labs(
    title = "Presence model performance across CV folds",
    subtitle = "Logistic part of the two-part model",
    x = NULL,
    y = "ROC AUC"
  ) +
  theme_bw()

# Confusion matrix
pred_class <- if_else(pred_presence > 0.5, 1, 0)
confusionMatrix(
  factor(pred_class),
  factor(as.numeric(test_orig_s$tf_primary_rate > 0)),
  positive = "1"
)
```

## How is the INTENSITY model performing?

```{r intensity-model-performance}
test_pos_idx <- test_orig_s$tf_primary_rate > 0

cat("Intensity component (evaluated on true positives only, n =", 
    sum(test_pos_idx), ")\n")
cat("MAE:     ", round(mae_fn(true_orig[test_pos_idx], 
                               pred_intensity[test_pos_idx]), 4), "\n")
cat("RMSE:    ", round(rmse_fn(true_orig[test_pos_idx], 
                                pred_intensity[test_pos_idx]), 4), "\n")
cat("Spearman:", round(cor(true_orig[test_pos_idx], 
                            pred_intensity[test_pos_idx], 
                            method = "spearman"), 4), "\n")

tibble(
  true = true_orig[test_pos_idx],
  pred = pred_intensity[test_pos_idx]
) %>%
  ggplot(aes(x = true, y = pred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, 
              linetype = "dashed", color = "grey40") +
  coord_equal() +
  labs(
    title = "Intensity component: predicted vs observed",
    subtitle = paste0("True positives only (n = ", sum(test_pos_idx), 
                      ") — original scale"),
    x = "Observed tf_primary_rate",
    y = "Predicted (Gamma GLM)"
  ) +
  theme_bw()
```

## 
