---
title: "14_TERT_missingness"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

# Purpose

TERT expression is missing for a large subset of samples. This analysis checks
whether missingness is associated with ALT status, cancer type, or other telomere
features before complete-case models are interpreted.

# Setup

Create explicit missingness indicators so observed and missing TERT groups can be
compared without imputing expression values.

```{r setup}
library(tidyverse)
library(naniar)
library(here)
source(here('scripts', 'clean_columns.R'))

data <- readxl::read_xlsx(here('data', 'raw', 'PCAWG_variables_PCA_woK.xlsx'))
clean_data <- clean_pcawg_data(data)

data_analysis <- clean_data %>%
  mutate(
    tert_missing_indicator = is.na(TERT_FPKM),
    tert_observed = !is.na(TERT_FPKM)
  )
```

# Missingness overview

The plots assess both overall missingness and whether TERT availability occupies a
distinct region of the telomere-feature space.

```{r visualize}
cat("=== MISSINGNESS OVERVIEW ===\n")
print(naniar::miss_var_summary(clean_data))
cat("\nTERT NA:', round(sum(is.na(clean_data$TERT_FPKM))/nrow(clean_data)*100, 1), '%\n")

cat('=== MISSING DATA HEATMAP ===\n')
vis_miss(clean_data, sort_miss = TRUE)

clean_data %>% 
  mutate(missing_tert = factor(is.na(TERT_FPKM),
                                levels = c('TRUE', 'FALSE'),
                                labels = c('Missing', 'Not Missing'))) %>% 
  ggplot(aes(x = alt_status, fill = missing_tert))+
  geom_bar(position = 'stack')+
  labs(title = 'Distribution of ALT status for Missing vs. Non-Missing',
       x = 'ALT status',
       y = 'Counts',
       fill = 'Missingness')+
  theme_bw()

# Restrict the display range to keep extreme values from obscuring the main cloud.
clean_data %>% 
  filter(telomere_insertion_rate < 4, tf_rate < 15) %>% 
  select(TERT_FPKM, tf_rate, telomere_insertion_rate, telomere_content_log2) %>% 
  ggplot(aes(x = tf_rate,
             y = telomere_insertion_rate,
             size = telomere_content_log2,
             colour = is.na(TERT_FPKM)))+
  geom_point(alpha=0.6)+
  facet_wrap(~is.na(TERT_FPKM))+
  labs(title = 'Missing TERT data by TF rate and Telomeric Insertion',
       x = 'TF rate',
       y = 'Telomeric insertion',
       color = 'Missing TERT data')+
  theme_bw()
```

# Compare observed and missing TERT groups

```{r distributions-comparison}
cat("\n=== Comparison TERT Present vs Missing ===\n")
 
# Compare the principal continuous telomere measurements.
numeric_vars <- c('tf_rate', 'telomere_insertion_rate', 
                  'telomere_content_log2', 'TERT_FPKM',
                  'ATAGGG_singleton_dist', 'CTAGGG_singleton_dist', 
                  'GTAGGG_singleton_dist', 'TAAGGG_singleton_dist',
                  'TCAGGG_singleton_dist', 'TGAGGG_singleton_dist',
                  'TTCGGG_singleton_dist', 'TTGGGG_singleton_dist',
                  'TTTGGG_singleton_dist')
 
# Keep the analysis robust to source versions that omit a listed measurement.
numeric_vars <- numeric_vars[numeric_vars %in% names(clean_data)]
 
# Summarize each measurement by TERT missingness group.
comparison_numeric <- data_analysis %>%
  group_by(tert_missing_indicator) %>%
  summarise(
    n = n(),
    across(all_of(numeric_vars), list(mean = ~mean(., na.rm = TRUE),
                                       sd = ~sd(., na.rm = TRUE),
                                       median = ~median(., na.rm = TRUE)),
           .names = '{.col}_{.fn}')
  )
 
# Print variable-level summaries for direct inspection.
for (var in numeric_vars) {
  cat('---', var, "---\n")
  
  tert_present <- clean_data[[var]][!is.na(clean_data$TERT_FPKM)]
  tert_missing <- clean_data[[var]][is.na(clean_data$TERT_FPKM)]
  
  cat('TERT Present  | Mean:', round(mean(tert_present, na.rm=T), 4),
      '| SD:', round(sd(tert_present, na.rm=T), 4),
      '| Median:', round(median(tert_present, na.rm=T), 4), "\n")
  cat('TERT Missing  | Mean:', round(mean(tert_missing, na.rm=T), 4),
      '| SD:', round(sd(tert_missing, na.rm=T), 4),
      '| Median:', round(median(tert_missing, na.rm=T), 4), "\n\n")
}
 
# Check whether TERT availability is concentrated in particular cancer types.
cat("\n--- cancer_type ---\n")
comparison_cancer <- data_analysis %>%
  group_by(cancer_type, tert_missing_indicator) %>%
  summarise(n = n(), .groups = 'drop') %>%
  pivot_wider(names_from = tert_missing_indicator, values_from = n, values_fill = 0) %>%
  mutate(
    total = `FALSE` + `TRUE`,
    pct_missing = round(100 * `TRUE` / total, 1)
  ) %>%
  arrange(desc(pct_missing))
 
cat("TERT Missing % by Cancer Type:\n")
print(comparison_cancer, n = nrow(comparison_cancer))

# Check whether complete-case modeling could alter ALT-class balance.
cat("\n--- alt_status ---\n")
comparison_alt <- data_analysis %>%
  group_by(alt_status, tert_missing_indicator) %>%
  summarise(n = n(), .groups = 'drop') %>%
  pivot_wider(names_from = tert_missing_indicator, values_from = n, values_fill = 0) %>%
  mutate(
    total = `FALSE` + `TRUE`,
    pct_missing = round(100 * `TRUE` / total, 1)
  ) %>%
  arrange(desc(pct_missing))
 
cat("TERT Missing % by ALT status:\n")
print(comparison_alt, n = nrow(comparison_alt))
```
