---
title: "15_1_bTF_regression_+tumor"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

This analysis extends the blood TF regression pipeline by including matched primary-tumor TF rate as a predictor. It evaluates whether tumor fusion activity adds predictive signal beyond the remaining telomere features.

# Setup

```{r setup}
library(tidyverse)
library(glmnet)
library(here)
library(randomForest)
library(caret)
library(pROC)
library(broom)
library(treeshap)
library(shapviz)
library(patchwork)

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
clean_data <- data %>%
  drop_na()
cancer_type_vector <- clean_data$cancer_type
cancer_group_vector <- clean_data$cancer_group
```

## Transformations

Since we are in a regression task, it is better to have a distribution for our response variable as symmetric as possible. We know it is currently strongly skewed, thus we apply a \`log1p\` transformation to \`tf_blood_rate\`, \`tf_primary_rate\`, \`TERT_FPKM\`, and \`telomere_insertion_rate\` to deal with their asymmetry and zero-inflation.

The log1p transformation reduces the influence of the outliers for linear models, but it doesn't alter much the zero-inflated structure of the distribution.

```{r log1p-transformation}
clean_data_orig <- clean_data %>% 
  mutate(
    tf_primary_rate = log1p(tf_primary_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM = log1p(TERT_FPKM),
    alt_status = factor(
      alt_status,
      levels = c('ALT-low', 'ALT-high')
    )
  ) %>%  
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)

clean_data <- clean_data %>% 
  mutate(
    tf_primary_rate = log1p(tf_primary_rate),
    tf_blood_rate = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM = log1p(TERT_FPKM),
    alt_status = factor(
      alt_status,
      levels = c('ALT-low', 'ALT-high')
    )
  ) %>% 
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)
```

# Split the dataset

We split the dataset with a 80/20 split in training set and test set. We won't touch the latter until final performance estimation, to avoid data leakage.

```{r train-test}
set.seed(21)
train_idx <- sample(nrow(clean_data), 0.8*nrow(clean_data))

train <- clean_data[train_idx,]
test <- clean_data[-train_idx,]

train_orig <- clean_data_orig[train_idx,]
test_orig <- clean_data_orig[-train_idx,]
```

# Metrics functions

We define functions to help us in the evaluation of the performances of the models. In particular we will use RMSE and R squared.

```{r metrics}
rmse_fn <- function(actual, predicted) sqrt(mean((actual-predicted)^2))
mae_fn  <- function(actual, predicted) mean(abs(actual - predicted))
```

------------------------------------------------------------------------

# Manual Cross-Validation

We use a 5-fold cross-validation on the training set. We do it manually to have to possibility to further extend to other metrics, methods, or whatever.

```{r cv-setup}
k <- 5
n_repeats <- 5
set.seed(21)

cv_results <- tibble(fold = integer(), model = character(), rmse = double(), mae = double())
lasso_coef_list <- list()
rf_importance_list <- list()
idx <- 1
```

```{r cv-loop}
for (rep in 1:n_repeats){
  set.seed(21 + rep)
  fold_list <- createFolds(factor(train$alt_status), k = k, returnTrain = FALSE)
  folds <- integer(nrow(train))
  for (i in seq_along(fold_list)) folds[fold_list[[i]]] <- i
  for (i in 1:k){
    fold_train <- train[folds!=i,]
    fold_test <- train[folds==i,]
    
    # Standardize the predictors
    cols_to_scale <- setdiff(names(fold_train), c('tf_blood_rate', 'alt_status'))
    preproc <- preProcess(fold_train %>% select(all_of(cols_to_scale)),
                           method = c('center', 'scale'))
    fold_train_s <- fold_train %>%
      mutate(predict(preproc, fold_train %>% select(all_of(cols_to_scale))))
    fold_test_s  <- fold_test %>%
      mutate(predict(preproc, fold_test %>% select(all_of(cols_to_scale))))
    
    # LM
    fit_lm <- lm(tf_blood_rate ~ ., data = fold_train_s)
    pred_lm <- predict(fit_lm, newdata = fold_test_s)
    
    # LASSO
    x_fold_train <- model.matrix(tf_blood_rate ~ ., fold_train_s)[,-1]
    y_fold_train <- fold_train_s$tf_blood_rate
    x_fold_test <- model.matrix(tf_blood_rate ~ ., fold_test_s)[,-1]
    
    cv_l <- cv.glmnet(x_fold_train, y_fold_train, alpha=1, nfolds=5)
    fit_lasso <- glmnet(x_fold_train, y_fold_train, alpha=1, lambda = cv_l$lambda.min)
    pred_lasso <- as.vector(predict(fit_lasso, newx = x_fold_test))
    
    ## LASSO coefficients for this fold
    lasso_coef_list[[idx]] <- coef(fit_lasso) %>% 
      as.matrix() %>% 
      as.data.frame() %>% 
      rownames_to_column('feature') %>% 
      rename(coef = s0) %>% 
      mutate(rep = rep, fold = i)
    
    
    # Random Forest
    fit_rf <- randomForest(tf_blood_rate ~ ., data = fold_train, importance = TRUE)
    pred_rf <- predict(fit_rf, newdata = fold_test)
    
    ## Importance of this fold
    rf_importance_list[[idx]] <- importance(fit_rf) %>% 
      as.data.frame() %>% 
      rownames_to_column('feature') %>% 
      mutate(rep = rep, fold = i)
    
    idx <- idx + 1
    
    # Metrics
    cv_results <- cv_results %>% bind_rows(
      tibble(rep = rep, fold = i, model = 'lm',
             rmse = rmse_fn(fold_test$tf_blood_rate, pred_lm),
             mae = mae_fn(fold_test$tf_blood_rate, pred_lm)),
      tibble(rep = rep, fold = i, model = 'lasso',
             rmse = rmse_fn(fold_test$tf_blood_rate, pred_lasso),
             mae = mae_fn(fold_test$tf_blood_rate, pred_lasso)),
      tibble(rep = rep, fold = i, model = 'rf',
             rmse = rmse_fn(fold_test$tf_blood_rate, pred_rf),
             mae = mae_fn(fold_test$tf_blood_rate, pred_rf))
    )
  }
}
```

------------------------------------------------------------------------

# CV results

Averages and standard deviations on 5 folds.

```{r cv-performance-table}
cv_summary <- cv_results %>% 
  group_by(model) %>% 
  summarise(
    rmse_mean = mean(rmse), rmse_sd = sd(rmse),
    mae_mean = mean(mae), mae_sd = sd(mae),
    .groups = 'drop'
  ) %>% 
  arrange(rmse_mean)

cv_summary
```

```{r cv-performance-plot}
cv_results %>%
  pivot_longer(c(rmse, mae), names_to = 'metric', values_to = 'value') %>%
  mutate(model = fct_reorder(model, value, .fun = mean)) %>%
  ggplot(aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  facet_wrap(~metric, scales = 'free_y',
             labeller = labeller(metric = c(rmse = 'RMSE', mae = 'MAE'))) +
  labs(title = 'Metrics distribution per fold (5-fold)',
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = 'none')
```

## Feature importance : LASSO

For each feature we register in how many folds is has been selected from Lasso and the mean/SD of the coefficient. A feature selected in all folds with a stable coefficient should be a strong enough signal. Otherwise, rarely selected features or unstable coefficients must be interpreted carefully.

```{r lasso-stability}
lasso_stability <- bind_rows(lasso_coef_list) %>% 
  filter(feature != '(Intercept)') %>% 
  group_by(feature) %>% 
  summarise(
    n_selected = sum(coef != 0),
    freq = n_selected / (n_repeats*k),
    coef_mean = mean(coef[coef != 0]),
    coef_sd = sd(coef[coef != 0]),
    .groups = 'drop'
  ) %>% 
  arrange(desc(freq), desc(abs(coef_mean)))
lasso_stability
```

```{r lasso-stability-plot}
plt_lasso_coef <- lasso_stability %>% 
                    filter(n_selected > 0) %>% 
                    mutate(feature = fct_reorder(feature, abs(coef_mean))) %>% 
                    ggplot(aes(x = coef_mean, y = feature, fill = freq))+
                    geom_col(alpha = 0.8)+
                    geom_errorbarh(aes(xmin = coef_mean - coef_sd,
                                       xmax = coef_mean + coef_sd),
                                   width = 0.25, color = 'grey30')+
                    scale_fill_gradient(low = 'grey90', high = 'steelblue')+
                    geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey50')+
                    labs(title = 'LASSO coefficients aggregated on 5 folds',
                         subtitle = 'Mean ± SD',
                         x = 'Mean coefficient', y = NULL)+
                    theme_bw()
plt_lasso_coef
```

## Feature importance: Random Forest

```{r rf-importance}
rf_importance_summary <- bind_rows(rf_importance_list) %>% 
  group_by(feature) %>% 
  summarise(
    inc_mse_mean = mean(`%IncMSE`),
    inc_mse_sd = sd(`%IncMSE`),
    .groups = 'drop'
    ) %>% 
  arrange(desc(inc_mse_mean))
rf_importance_summary
```

```{r rf-importance-plot}
plt_rf_importance <- rf_importance_summary %>% 
                        mutate(feature = fct_reorder(feature, inc_mse_mean)) %>% 
                        ggplot(aes(x = inc_mse_mean, y = feature))+
                        geom_col(fill='seagreen', alpha = 0.8) +
                        geom_errorbarh(aes(xmin = inc_mse_mean - inc_mse_sd,
                                           xmax = inc_mse_mean + inc_mse_sd),
                                       width = 0.3, color = 'grey30')+
                        labs(title = 'RF importance aggregated on 5 folds (%IncMSE)',
                             subtitle = 'Mean ± SD',
                             x = 'Mean %IncMSE', y = NULL) +
                        theme_bw()
plt_rf_importance
```

# Final evaluation on test set

Let's fit each model on the whole training set and evaluate on the test set.

```{r final-lm}
final_lm   <- lm(tf_blood_rate ~ ., data = train)
pred_lm_f  <- predict(final_lm, newdata = test)
```

```{r final-lasso}
x_train_f  <- model.matrix(tf_blood_rate ~ ., train)[, -1]
y_train_f  <- train$tf_blood_rate
x_test_f   <- model.matrix(tf_blood_rate ~ ., test)[,  -1]

set.seed(21)
cv_lasso_f    <- cv.glmnet(x_train_f, y_train_f, alpha = 1)
final_lasso   <- glmnet(x_train_f, y_train_f, alpha = 1,
                        lambda = cv_lasso_f$lambda.min)
pred_lasso_f  <- as.vector(predict(final_lasso, newx = x_test_f))
```

```{r final-rf}
set.seed(21)
final_rf  <- randomForest(tf_blood_rate ~ ., data = train, importance = TRUE)
pred_rf_f <- predict(final_rf, newdata = test)
```

```{r shapr-rf-cv}
clean_data_shap <- clean_data %>%
  # randomForest.unify() does not support categorical predictors.
  mutate(alt_status = as.numeric(alt_status == 'ALT-high'))

k <- 5
shap_list <- list()

set.seed(21)
fold_list <- createFolds(
  factor(clean_data_shap$alt_status),
  k = k,
  returnTrain = FALSE
)
folds <- integer(nrow(clean_data_shap))
for (i in seq_along(fold_list)) folds[fold_list[[i]]] <- i

for (i in 1:k) {
  fold_train      <- clean_data_shap[folds != i, ]
  fold_test       <- clean_data_shap[folds == i, ]
  fold_train_shap <- fold_train %>% select(-tf_blood_rate)
  fold_test_shap  <- fold_test  %>% select(-tf_blood_rate)
  
  fit_rf_fold  <- randomForest(tf_blood_rate ~ ., data = fold_train)
  unified_fold <- randomForest.unify(fit_rf_fold, fold_train_shap)
  shap_fold    <- treeshap(unified_fold, fold_test_shap, verbose = FALSE)
  
  shap_list[[i]] <- list(
    shap   = shap_fold$shaps,
    X      = fold_test_shap,
    row_id = which(folds == i)
  )
}

shap_matrix <- bind_rows(lapply(shap_list, `[[`, 'shap'))
X_matrix    <- bind_rows(lapply(shap_list, `[[`, 'X'))
row_ids     <- unlist(lapply(shap_list, `[[`, 'row_id'))
ord         <- order(row_ids)
shap_matrix <- shap_matrix[ord, ]
X_matrix    <- X_matrix[ord, ]

stopifnot(nrow(shap_matrix) == length(cancer_group_vector))
stopifnot(nrow(shap_matrix) == length(cancer_type_vector))

# Build a single aligned metadata frame
shap_meta <- tibble(
  cancer_group = factor(cancer_group_vector,
                        levels = c('Non_Mesenchymal_origin', 'Mesenchymal_origin')),
  cancer_type  = fct_lump_n(factor(cancer_type_vector), n = 8),
  alt_status   = clean_data_shap$alt_status
)

# OOF SHAP on the full dataset: exploratory, not predictive validation.
sv_oof <- shapviz(
  object   = as.matrix(shap_matrix),
  X        = X_matrix,
  baseline = mean(clean_data_shap$tf_blood_rate)
)

cancer_lumped <- fct_lump_n(factor(cancer_type_vector), n = 8)

dep_by_group <- function(v) {
  tibble(
    feature_value = X_matrix[[v]],
    shap_value    = shap_matrix[[v]],
    cancer_group  = shap_meta$cancer_group
  ) %>%
    ggplot(aes(feature_value, shap_value)) +
    geom_point(aes(color = cancer_group, size = cancer_group,
                   alpha = cancer_group)) +
    scale_color_manual(values = c('grey70', 'firebrick')) +
    scale_size_manual(values = c(1.2, 2.2)) +
    scale_alpha_manual(values = c(0.4, 0.9)) +
    labs(x = v, y = paste('SHAP value for', v)) +
    theme_bw()
}

dep_by_cancer <- function(v) {
  tibble(
    feature_value = X_matrix[[v]],
    shap_value    = shap_matrix[[v]],
    cancer_type   = shap_meta$cancer_type
  ) %>%
    ggplot(aes(feature_value, shap_value, color = cancer_type)) +
    geom_point(alpha = 0.7, size = 1.5) +
    labs(x = v, y = paste('SHAP value for', v), color = 'Cancer type') +
    theme_bw()
}

shap_by_group_summary <- shap_matrix %>%
  mutate(cancer_group = cancer_group_vector) %>%
  pivot_longer(-cancer_group, names_to = 'feature', values_to = 'shap') %>%
  group_by(feature, cancer_group) %>%
  summarise(mean_abs_shap = mean(abs(shap)),
            n = n(), .groups = 'drop') %>%
  pivot_wider(names_from = cancer_group, values_from = c(mean_abs_shap, n)) %>%
  mutate(diff = mean_abs_shap_Mesenchymal_origin - mean_abs_shap_Non_Mesenchymal_origin) %>%
  arrange(desc(abs(diff)))

# Plots

sv_importance(sv_oof, kind = 'beeswarm', bee_adjust = 5, alpha = 0.8) + theme_bw()
sv_dependence(sv_oof, v = 'telomere_content_log2', color_var = 'alt_status') + theme_bw()
dep_by_cancer('telomere_content_log2')
dep_by_group('telomere_content_log2')

sv_dependence(sv_oof, v = 'telomere_insertion_rate',         color_var = 'alt_status') + theme_bw()
dep_by_cancer('telomere_insertion_rate')
dep_by_group('telomere_insertion_rate')

sv_dependence(sv_oof, v = 'tf_primary_rate',       color_var = 'alt_status') + theme_bw()
dep_by_cancer('tf_primary_rate')
dep_by_group('tf_primary_rate')

sv_dependence(sv_oof, v = 'TERT_FPKM',             color_var = 'alt_status') + theme_bw()
dep_by_cancer('TERT_FPKM')
dep_by_group('TERT_FPKM')

sv_dependence(sv_oof, v = 'alt_status',            color_var = 'tf_primary_rate') + theme_bw()
dep_by_group('alt_status')

singleton_vars <- c(
    'ATAGGG_singleton_dist',
    'CTAGGG_singleton_dist',
    'GTAGGG_singleton_dist',
    'TAAGGG_singleton_dist',
    'TCAGGG_singleton_dist',
    'TGAGGG_singleton_dist',
    'TTCGGG_singleton_dist',
    'TTGGGG_singleton_dist',
    'TTTGGG_singleton_dist'
  )
singleton_plots_alt <- purrr::map(
  singleton_vars,
  \(v) {
    sv_dependence(sv_oof, v = v, color_var = 'alt_status') +
      theme_bw() +
      theme(
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 7)
      )
  }
)
wrap_plots(singleton_plots_alt, ncol = 3, guides = 'collect') +
  plot_annotation(title = '***GGG singleton SHAP dependence')

singleton_plots_cancer <- purrr::map(
  singleton_vars,
  \(v) {
    dep_by_cancer(v) +
      theme_bw() +
      theme(
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 7)
      )
  }
)
wrap_plots(singleton_plots_cancer, ncol = 3, guides = 'collect') +
  plot_annotation(title = '***GGG singleton SHAP dependence')

singleton_plots_cancer <- purrr::map(
  singleton_vars,
  \(v) {
    dep_by_group(v) +
      theme_bw() +
      theme(
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 7)
      )
  }
)
wrap_plots(singleton_plots_cancer, ncol = 3, guides = 'collect') +
  plot_annotation(title = '***GGG singleton SHAP dependence')

alt_high_idx <- which(clean_data_shap$alt_status == 1)
alt_low_idx  <- which(clean_data_shap$alt_status == 0)

sv_waterfall(sv_oof, row_id = alt_high_idx) +
  ggtitle('Average SHAP - ALT-high')

sv_waterfall(sv_oof, row_id = alt_low_idx) +
  ggtitle('Average SHAP - ALT-low')
```

```{r final-results}
tibble(
  model = c('LM', 'LASSO', 'Random Forest'),
  rmse  = c(rmse_fn(test$tf_blood_rate, pred_lm_f),
            rmse_fn(test$tf_blood_rate, pred_lasso_f),
            rmse_fn(test$tf_blood_rate, pred_rf_f)),
  mae   = c(mae_fn(test$tf_blood_rate, pred_lm_f),
            mae_fn(test$tf_blood_rate, pred_lasso_f),
            mae_fn(test$tf_blood_rate, pred_rf_f)),
  Spearman = c(
    cor(test$tf_blood_rate, pred_lm_f,    method = 'spearman'),
    cor(test$tf_blood_rate, pred_lasso_f, method = 'spearman'),
    cor(test$tf_blood_rate, pred_rf_f,    method = 'spearman')
  )
) %>%
  arrange(rmse)
```

## Quantification of the residuals

Results are back-transformed to the original scale.

```{r quantification-error}
df <- tibble(
  true = expm1(test$tf_blood_rate),
  pred = expm1(pred_rf_f),
  error = pred - true,
  abs_error = abs(error),
  perc_error = if_else(
    true == 0,
    NA_real_,
    (pred - true) / true * 100
  ),
  cancer_type = cancer_type_vector[-train_idx]
)
print(df, n=50)

df %>%
  summarise(
    mean_error = mean(error),
    sd_error = sd(error),
    mae = mean(abs_error),
    median_ae = median(abs_error),
    max_ae = max(abs_error),
    rmse = sqrt(mean(error^2))
  )

df %>%
  group_by(cancer_type) %>% 
  mutate(n_label = paste0(cancer_type, ' (n=', n(), ')')) %>% 
  ungroup() %>% 
  mutate(n_label = fct_reorder(n_label, abs_error, .fun = median)) %>%
  ggplot(aes(x = n_label, y = abs_error)) +
  geom_boxplot(outlier.shape = 21, fill = 'steelblue', alpha = 0.6) +
  coord_flip() +
  labs(title = 'Absolute error by cancer type (test set)',
       x = NULL, y = 'Absolute error (original scale)') +
  theme_bw()
```

# Two-part model

In the cross-validation block below, the intensity part is fit on the `log1p`-transformed response used in the main regression pipeline. This keeps the coefficient-stability summaries on comparable transformed-response and standardized-predictor scales. The final two-part regression later is fit on `train_orig_s` to report predictions on the original response scale, so the CV coefficients and final prediction table answer related but different questions.

```{r two-model}
k <- 5
n_repeats <- 5
presence_coef_list <- list()
intensity_coef_list <- list()
presence_results <- tibble()
idx <- 1

for(rep in 1:n_repeats){
  set.seed(21 + rep)
  fold_list <- createFolds(
    factor(train$alt_status),
    k = k,
    returnTrain = FALSE
  )
  folds <- integer(nrow(train))

  for(i in seq_along(fold_list)){
    folds[fold_list[[i]]] <- i
  }

  for(i in 1:k){
    fold_train <- train[folds != i, ]
    fold_test  <- train[folds == i, ]
    cols_to_scale <- setdiff(
      names(fold_train),
      c('tf_blood_rate', 'alt_status')
    )
    preproc <- preProcess(
      fold_train %>% select(all_of(cols_to_scale)),
      method = c('center', 'scale')
    )
    fold_train_s <- fold_train %>%
      mutate(
        predict(
          preproc,
          fold_train %>% select(all_of(cols_to_scale))
        )
      )
    fold_test_s <- fold_test %>%
      mutate(
        predict(
          preproc,
          fold_test %>% select(all_of(cols_to_scale))
        )
      )
    # PART 1: Presence
    m_presence_cv <- glm(
      I(tf_blood_rate > 0) ~ .,
      family = binomial(link = 'logit'),
      data = fold_train_s
    )
    presence_coef_list[[idx]] <- tidy(m_presence_cv) %>%
      filter(term != '(Intercept)') %>%
      mutate(
        rep = rep,
        fold = i
      )
    pred_presence_cv <- predict(
      m_presence_cv,
      newdata = fold_test_s,
      type = 'response'
    )
    presence_results <- bind_rows(
      presence_results,
      tibble(
        rep = rep,
        fold = i,
        truth = as.numeric(fold_test_s$tf_blood_rate > 0),
        prob = pred_presence_cv
      )
    )
    # PART 2: Intensity
    df_positive_cv <- fold_train_s %>%
      filter(tf_blood_rate > 0)
    m_intensity_cv <- glm(
      tf_blood_rate ~ .,
      family = Gamma(link = 'log'),
      data = df_positive_cv
    )
    intensity_coef_list[[idx]] <- tidy(m_intensity_cv) %>%
      filter(term != '(Intercept)') %>%
      mutate(
        rep = rep,
        fold = i
      )
    idx <- idx + 1
  }
}

```

## Importance of the two models

```{r importance-two-models}
importance_cv <- bind_rows(
  bind_rows(presence_coef_list) %>%
    mutate(model_part = 'Presence (P(TF>0))'),
  bind_rows(intensity_coef_list) %>%
    mutate(model_part = 'Intensity (E[TF|TF>0])')
) %>%
  group_by(model_part, term) %>%
  summarise(
    coef_mean = mean(estimate),
    coef_sd = sd(estimate),
    p_median = median(p.value, na.rm = TRUE),
    freq_p05 = mean(p.value < 0.05, na.rm = TRUE),
    sign_stability =
      mean(sign(estimate) == sign(mean(estimate))),
    .groups = 'drop'
  ) %>%
  mutate(
    significance = case_when(
      freq_p05 >= 0.9 ~ '***',
      freq_p05 >= 0.7 ~ '**',
      freq_p05 >= 0.5 ~ '*',
      TRUE ~ 'ns'
    )
  )


importance_cv %>%
  mutate(term = fct_reorder(term, abs(coef_mean))) %>%
  ggplot(aes(x = coef_mean,
             y = term,
             fill = model_part)) +
  geom_col(alpha = 0.8) +
  geom_errorbarh(
    aes(
      xmin = coef_mean - coef_sd,
      xmax = coef_mean + coef_sd
    ),
    height = 0.2,
    color = 'grey30'
  ) +
  geom_text(
    aes(label = significance),
    hjust = -0.3,
    vjust = -0.5,
    size = 3
  ) +
  geom_vline(
    xintercept = 0,
    linetype = 'dashed',
    color = 'grey50'
  ) +
  facet_wrap(~model_part, scales = 'free_y') +
  theme_minimal() +
  labs(
    title = 'Two-part model coefficient stability across CV folds',
    subtitle =
      '***: significant in ≥90% folds; **: ≥70%; *: ≥50%',
    x = 'Mean coefficient across CV folds',
    y = NULL
  ) + 
  theme(legend.position = 'none')
```

## Regression on test set

Here we do regression using the two-part model. We preprocess the training and test set with a standardization and, to have fair comparisons with other models, we apply the Duan smearing correction to expm1 re-transformation.

```{r two-part model regression}
# Preprocessing on training set
cols_to_scale_orig <- setdiff(names(train_orig), c('tf_blood_rate', 'alt_status'))

preproc_orig <- preProcess(
  train_orig %>% select(all_of(cols_to_scale_orig)),
  method = c('center', 'scale')
)

train_orig_s <- train_orig %>%
  mutate(predict(preproc_orig, train_orig %>% select(all_of(cols_to_scale_orig))))

test_orig_s <- test_orig %>%
  mutate(predict(preproc_orig, test_orig %>% select(all_of(cols_to_scale_orig))))

# Fit final models on the whole training set
m_presence_final <- glm(
  I(tf_blood_rate > 0) ~ .,
  family = binomial(link = 'logit'),
  data = train_orig_s
)

m_intensity_final <- glm(
  tf_blood_rate ~ .,
  family = Gamma(link = 'log'),
  data = train_orig_s %>% filter(tf_blood_rate > 0),
  control = glm.control(maxit = 200)
)

# Prediction on test set
pred_presence  <- predict(m_presence_final, newdata = test_orig_s, type = 'response')
pred_intensity <- predict(m_intensity_final, newdata = test_orig_s, type = 'response')
pred_combined  <- pred_presence * pred_intensity    # formula of the two-part model

true_orig <- test_orig$tf_blood_rate

# Correction of log1p retransformation of LM/LASSO/RF (smearing)
smear_lm    <- mean(exp(residuals(final_lm)))
smear_lasso <- mean(exp(y_train_f - as.vector(predict(final_lasso, newx = x_train_f))))
smear_rf    <- mean(exp(train$tf_blood_rate - predict(final_rf)))

pred_lm_orig    <- exp(pred_lm_f)    * smear_lm - 1
pred_lasso_orig <- exp(pred_lasso_f) * smear_lasso - 1
pred_rf_orig    <- exp(pred_rf_f)    * smear_rf - 1

tibble(
  model = c('LM', 'LASSO', 'Random Forest', 'Two-part'),
  RMSE = c(
    rmse_fn(true_orig, pred_lm_orig),
    rmse_fn(true_orig, pred_lasso_orig),
    rmse_fn(true_orig, pred_rf_orig),
    rmse_fn(true_orig, pred_combined)
  ),
  MAE = c(
    mean(abs(true_orig - pred_lm_orig)),
    mean(abs(true_orig - pred_lasso_orig)),
    mean(abs(true_orig - pred_rf_orig)),
    mean(abs(true_orig - pred_combined))
  ),
  Spearman = c(
    cor(true_orig, pred_lm_orig,    method = 'spearman'),
    cor(true_orig, pred_lasso_orig, method = 'spearman'),
    cor(true_orig, pred_rf_orig,    method = 'spearman'),
    cor(true_orig, pred_combined,   method = 'spearman')
  )
) %>% arrange(RMSE)
```

## Presence-model performance

```{r presence-model-auc}
# AUC of presence model
presence_auc <- presence_results %>%
  group_by(rep, fold) %>%
  summarise(
    auc = as.numeric(pROC::roc(truth, prob, quiet = TRUE)$auc),
    .groups = 'drop'
  )
presence_auc %>%
  summarise(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE)
  )
presence_auc %>%
  ggplot(aes(x = 'Presence model', y = auc)) +
  geom_boxplot(fill = 'steelblue', alpha = 0.7, width = 0.4) +
  geom_jitter(width = 0.08, alpha = 0.6) +
  labs(
    title = 'Presence model performance across CV folds',
    subtitle = 'Logistic part of the two-part model',
    x = NULL,
    y = 'ROC AUC'
  ) +
  theme_bw()

# Confusion matrix
pred_class <- if_else(pred_presence > 0.5, 1, 0)
confusionMatrix(
  factor(pred_class),
  factor(as.numeric(test_orig_s$tf_blood_rate > 0)),
  positive = '1'
)
```

## Intensity-model performance

```{r intensity-model-performance}
test_pos_idx <- test_orig_s$tf_blood_rate > 0

cat('Intensity component (evaluated on true positives only, n =',
    sum(test_pos_idx), ")\n")
cat('MAE:     ', round(mae_fn(true_orig[test_pos_idx],
                               pred_intensity[test_pos_idx]), 4), "\n")
cat('RMSE:    ', round(rmse_fn(true_orig[test_pos_idx],
                                pred_intensity[test_pos_idx]), 4), "\n")
cat('Spearman:', round(cor(true_orig[test_pos_idx],
                            pred_intensity[test_pos_idx],
                            method = 'spearman'), 4), "\n")

tibble(
  true = true_orig[test_pos_idx],
  pred = pred_intensity[test_pos_idx]
) %>%
  ggplot(aes(x = true, y = pred)) +
  geom_point(alpha = 0.5, color = 'steelblue') +
  geom_abline(slope = 1, intercept = 0,
              linetype = 'dashed', color = 'grey40') +
  labs(
    title = 'Intensity component: predicted vs observed',
    subtitle = paste0('True positives only (n = ', sum(test_pos_idx),
                      ') — original scale'),
    x = 'Observed tf_blood_rate',
    y = 'Predicted (Gamma GLM)'
  ) +
  theme_bw()
```
