---
title: "19_ALT_status_blood_TF"
author: "Andrea Polazzi"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

# Motivation

All predictors used in the ALT-status classification so far (notebooks 01–06, 09, 10) are **tumor-derived**. The blood telomere fusion rate (`tf_blood_rate`) is instead a **germline / normal-tissue** measurement. Because ALT is a *somatic, tumor-intrinsic* phenotype, a blood-derived feature is not expected to predict tumor ALT status — and if it did, that would itself be an interesting constitutional-predisposition signal.

This makes blood TF a useful **negative control**: a feature that is (a) available without any sample-size cost (`tf_blood_rate` has no missing values, so the complete-case cohort stays at the same 817 patients as the full model in notebook 02), and (b) only weakly correlated with the tumor TF rate already in the model (Spearman ρ ≈ 0.2 on this cohort), so it cannot be dismissed as redundant *a priori*. We therefore test explicitly whether adding it improves ALT prediction on top of the established full feature set.

# Setup

```{r setup}
library(readxl)
library(tidyverse)
library(caret)
library(pROC)
library(glmnet)
library(here)

dir.create(here('outputs', 'blood_tf'), showWarnings = FALSE, recursive = TRUE)

data <- read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))

tvr_cols <- grep('_singleton_dist$', names(data), value = TRUE)

# Full tumor feature set (same as notebook 02) + the candidate blood feature
full_feats <- c('tf_primary_rate', 'telomere_insertion_rate',
                'telomere_content_log2', 'TERT_FPKM', tvr_cols)

model_data <- data %>%
  filter(alt_status %in% c('ALT-low', 'ALT-high')) %>%
  select(alt_status, tf_blood_rate, all_of(full_feats)) %>%
  filter(complete.cases(.)) %>%
  mutate(alt_status = factor(alt_status, levels = c('ALT-low', 'ALT-high')))

cat(sprintf("Cohort: %d patients, %d tumor features (+ blood TF)\n",
            nrow(model_data), length(full_feats)))
table(model_data$alt_status)
```

# Univariate association of blood TF with ALT status

Before any multivariate modeling, we check whether blood TF separates the two classes on its own.

```{r univariate, fig.width=5, fig.height=4}
p_uni <- ggplot(model_data, aes(alt_status, tf_blood_rate, fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  scale_y_log10() +
  theme_bw() +
  labs(x = NULL, y = 'Blood TF rate (log10)',
       title = 'Blood telomere fusion rate by ALT status') +
  theme(legend.position = 'none')

print(p_uni)
ggsave(here('outputs', 'blood_tf', 'blood_tf_by_alt_status.pdf'),
       p_uni, width = 5, height = 4)

w <- wilcox.test(tf_blood_rate ~ alt_status, data = model_data)
model_data %>%
  group_by(alt_status) %>%
  summarise(n = n(),
            median = median(tf_blood_rate),
            mean = mean(tf_blood_rate), .groups = 'drop')
cat('Wilcoxon rank-sum p =', signif(w$p.value, 3), "\n")

cat('Spearman cor(tumor TF, blood TF) =',
    round(cor(model_data$tf_primary_rate, model_data$tf_blood_rate,
              method = 'spearman'), 3), "\n")
```

The two distributions are essentially overlapping — if anything blood TF is marginally *lower* in ALT-high tumors — and the rank-sum test is non-significant. Blood TF is not a univariate discriminator of ALT status.

# Does blood TF add predictive signal?

We compare two models — `full` (tumor features only) and `full + blood TF` — using the same repeated stratified CV scheme as the rest of the project (5-fold × 5-repeat). Because the **same fold assignments** are used for both feature sets, the per-resample AUC difference is a *paired* quantity, and we test whether its distribution is centred at zero with a one-sample Wilcoxon signed-rank test.

```{r cv-setup}
# caret needs syntactically valid factor labels for classProbs
cv_data <- model_data %>%
  mutate(alt_status = factor(alt_status,
                             levels = c('ALT-low', 'ALT-high'),
                             labels = c('ALT_low', 'ALT_high')))

twoClassSummaryPR <- function(data, lev = NULL, model = NULL) {
  out  <- twoClassSummary(data, lev = lev, model = model)
  prob <- data[[lev[2]]]
  yb   <- as.integer(data$obs == lev[2])
  o    <- order(prob, decreasing = TRUE); ybo <- yb[o]
  tpp  <- cumsum(ybo); fpp <- cumsum(1 - ybo)
  P    <- tpp / (tpp + fpp); R <- tpp / sum(yb)
  c(out, PR_AUC = sum((R - c(0, head(R, -1))) * P))
}

set.seed(21)
folds <- createMultiFolds(cv_data$alt_status, k = 5, times = 5)

ctrl <- trainControl(
  method          = 'repeatedcv',
  number          = 5,
  repeats         = 5,
  index           = folds,
  classProbs      = TRUE,
  summaryFunction = twoClassSummaryPR,
  savePredictions = 'final'
)

feature_sets <- list(
  full       = cv_data %>% select(-tf_blood_rate),
  full_blood = cv_data
)
```

```{r training, cache=TRUE}
library(randomForest)

resamples_list <- list()

set.seed(21)
for (fs in names(feature_sets)) {
  df <- feature_sets[[fs]]

  log_model <- train(alt_status ~ ., data = df, method = 'glm',
                     family = 'binomial', metric = 'ROC', trControl = ctrl)
  rf_model  <- train(alt_status ~ ., data = df, method = 'rf',
                     metric = 'ROC', trControl = ctrl)

  for (nm in c('LogReg', 'RF')) {
    m  <- if (nm == 'LogReg') log_model else rf_model
    rs <- m$resample
    rs$feature_set <- fs
    rs$model       <- nm
    resamples_list[[paste(fs, nm, sep = '_')]] <- rs
  }
}

resamples_df <- bind_rows(resamples_list)
```

```{r paired-delta}
# Reference: per-resample AUC and PR-AUC of the full model, by model type
full_ref <- resamples_df %>%
  filter(feature_set == 'full') %>%
  select(Resample, model, ROC_full = ROC, PR_AUC_full = PR_AUC)

# Paired ΔAUC and ΔPR-AUC = metric[full + blood] − metric[full]
delta_df <- resamples_df %>%
  filter(feature_set == 'full_blood') %>%
  left_join(full_ref, by = c('Resample', 'model')) %>%
  mutate(delta_AUC    = ROC    - ROC_full,
         delta_PR_AUC = PR_AUC - PR_AUC_full)

blood_stat <- delta_df %>%
  group_by(model) %>%
  summarise(
    n             = sum(!is.na(delta_AUC)),
    auc_full      = mean(ROC_full,     na.rm = TRUE),
    auc_blood     = mean(ROC,          na.rm = TRUE),
    mean_delta    = mean(delta_AUC,    na.rm = TRUE),
    sd_delta      = sd(delta_AUC,      na.rm = TRUE),
    ci_low        = mean_delta - qt(0.975, df = n - 1) * sd_delta / sqrt(n),
    ci_high       = mean_delta + qt(0.975, df = n - 1) * sd_delta / sqrt(n),
    p_wilcoxon    = wilcox.test(delta_AUC[!is.na(delta_AUC)],
                                mu = 0, exact = FALSE)$p.value,
    pr_auc_full   = mean(PR_AUC_full,  na.rm = TRUE),
    pr_auc_blood  = mean(PR_AUC,       na.rm = TRUE),
    mean_pr_delta = mean(delta_PR_AUC, na.rm = TRUE),
    sd_pr_delta   = sd(delta_PR_AUC,   na.rm = TRUE),
    p_pr_wilcoxon = wilcox.test(delta_PR_AUC[!is.na(delta_PR_AUC)],
                                mu = 0, exact = FALSE)$p.value,
    .groups = 'drop'
  )

blood_stat %>%
  mutate(across(c(auc_full, auc_blood, mean_delta, sd_delta, ci_low, ci_high,
                  pr_auc_full, pr_auc_blood, mean_pr_delta, sd_pr_delta),
                ~ round(.x, 4)),
         p_wilcoxon    = signif(p_wilcoxon, 3),
         p_pr_wilcoxon = signif(p_pr_wilcoxon, 3)) %>%
  print(n = Inf)

write_csv(blood_stat, here('outputs', 'blood_tf', 'blood_tf_delta_auc.csv'))
```

```{r delta-plot, fig.width=7, fig.height=4}
p_delta <- ggplot(delta_df, aes(x = model, y = delta_AUC, fill = model)) +
  geom_hline(yintercept = 0, linetype = 'dashed', colour = 'grey40') +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8, show.legend = FALSE) +
  geom_jitter(width = 0.12, alpha = 0.3, size = 0.8, show.legend = FALSE) +
  coord_flip() +
  labs(
    title    = 'Per-resample ΔAUC from adding blood TF',
    subtitle = 'Positive = blood TF improves AUC; 25 resamples (5-fold × 5-repeat)',
    x = NULL, y = 'ΔAUC (full + blood TF  −  full)'
  ) +
  theme_bw(base_size = 11)

print(p_delta)
ggsave(here('outputs', 'blood_tf', 'blood_tf_delta_auc.pdf'),
       p_delta, width = 7, height = 4)
```

The mean ΔAUC is essentially zero for both models, the 95% CI straddles zero, and the paired Wilcoxon test is non-significant. Adding blood TF does not improve discrimination over the tumor-only full model.

# LASSO check: is blood TF selected?

As an independent confirmation, we add blood TF to the LASSO predictor matrix (same setup as notebook 09) and check whether the penalised model retains it.

```{r lasso}
X <- scale(as.matrix(model_data %>% select(tf_blood_rate, all_of(full_feats))))
y <- model_data$alt_status

set.seed(21)
cv_fit <- cv.glmnet(X, y, alpha = 1, family = 'binomial',
                    standardize = FALSE, nfolds = 10)

coef_1se <- coef(cv_fit, s = 'lambda.1se')
coef_min <- coef(cv_fit, s = 'lambda.min')

tibble(
  feature    = rownames(coef_1se)[-1],
  lambda_min = as.numeric(coef_min)[-1],
  lambda_1se = as.numeric(coef_1se)[-1]
) %>%
  mutate(feature = gsub('_singleton_dist', '', feature)) %>%
  arrange(desc(abs(lambda_1se))) %>%
  print(n = Inf)

cat("\nblood TF coefficient (lambda.1se):",
    signif(coef_1se['tf_blood_rate', 1], 3), "\n")
cat('blood TF coefficient (lambda.min):',
    signif(coef_min['tf_blood_rate', 1], 3), "\n")
```

# Conclusion

Three independent lines of evidence agree that the blood telomere fusion rate carries **no usable predictive signal for tumor ALT status** beyond the existing tumor features:

1. **Univariate:** blood TF distributions overlap across ALT classes (non-significant rank-sum test), despite being only weakly correlated with tumor TF.
2. **Paired CV:** adding blood TF to the full model produces a mean ΔAUC ≈ 0 with a 95% CI spanning zero and a non-significant Wilcoxon test, for both Logistic Regression and Random Forest.
3. **LASSO:** blood TF is shrunk toward / to zero and is not retained as a stable predictor.

This is the expected — and biologically meaningful — result: ALT is a somatic, tumor-intrinsic phenotype, so a germline/normal-tissue telomere-instability marker does not help classify it. The finding justifies the decision **not** to retrofit blood TF into the core ALT-classification series (notebooks 01–06), where it would add a non-informative feature without changing any conclusion. Blood TF is instead more naturally studied as a *response* variable (notebooks 15/17).
