---
title: "20_classification_comparison"
subtitle: "Fair head-to-head comparison of ALT-status classifiers"
author: "Andrea Polazzi"
date: today
format:
  html:
    toc: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

# Purpose

This notebook is the **single, fair comparison** of every classifier trained for
ALT-status prediction (notebooks 02, 09, 10). Earlier notebooks each evaluated a
model in isolation, on slightly different files, feature sets and CV schemes. Here
all models are put on **identical footing**:

- **Same cohort**: complete cases of `PCAWG_primary.xlsx` (n = 817; 768 ALT-low / 49
  ALT-high, i.e. **~6 % positive**).
- **Same features**: the 14-feature set (`features14`) including the blood telomere
  fusion rate, so no model has an information advantage.
- **Same resampling**: one repeated **5-fold × 5-repeat stratified CV** (25 folds,
  `set.seed(21)`), constructed with the exact code used in notebook 10, so the folds
  are *byte-identical* across all models (verified below). This makes the comparison
  **paired**: every model is scored on the same test samples in each fold.
- **Same metrics, same code**: AUC, PR-AUC (average precision — the honest summary
  under class imbalance), sensitivity, specificity, precision, F1 and balanced
  accuracy, all at a fixed 0.5 threshold where thresholded, computed by one helper.

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(forcats)
library(here)
library(randomForest)
library(pROC)
library(glmnet)
library(knitr)

features14 <- c(
  "tf_primary_rate", "tf_blood_rate", "telomere_insertion_rate",
  "telomere_content_log2", "TERT_FPKM",
  "ATAGGG_singleton_dist", "CTAGGG_singleton_dist", "GTAGGG_singleton_dist",
  "TAAGGG_singleton_dist", "TCAGGG_singleton_dist", "TGAGGG_singleton_dist",
  "TTCGGG_singleton_dist", "TTGGGG_singleton_dist", "TTTGGG_singleton_dist"
)

data <- read_xlsx(here("data", "processed", "PCAWG_primary.xlsx"))

model_full <- data %>%
  select(all_of(features14), alt_status, donor_id, cancer_type, cancer_group) %>%
  filter(complete.cases(across(all_of(features14)))) %>%
  mutate(alt_status = factor(alt_status, levels = c("ALT-low", "ALT-high")))

model_data <- model_full %>% select(all_of(features14), alt_status)
```

# Cross-validation folds

The 25 folds are built with the same construction (and seed) as the TabPFN notebook,
so the cached TabPFN per-fold results line up one-to-one with the folds used here.

```{r folds}
source(here("scripts", "make_folds.R"))
k <- 5
n_repeats <- 5
fold_mat  <- make_folds(model_data$alt_status, k = k, n_repeats = n_repeats, seed = 21)
cv_splits <- folds_to_splits(fold_mat)
```

```{r metric-helpers}
# avg_precision() + fold_metrics() — shared with notebook 10
source(here("scripts", "clf_metrics.R"))
```

# Train and score the in-R models (LR, RF, LASSO)

Logistic regression and random forest mirror notebook 02; LASSO mirrors notebook 09
but is here run through the *same* 25-fold harness, with `lambda.1se` selected by an
**inner** 10-fold CV on each training fold (nested CV, no leakage).

```{r run-in-r-models}
X <- as.matrix(model_data[, features14])
y <- model_data$alt_status

in_r_metrics <- tibble()
for (s in cv_splits) {
  dtr <- model_data[s$train, ]; dte <- model_data[s$test, ]

  m_lr <- suppressWarnings(glm(alt_status ~ ., data = dtr, family = "binomial"))
  p_lr <- suppressWarnings(predict(m_lr, dte, type = "response"))

  m_rf <- randomForest(alt_status ~ ., data = dtr, ntree = 500)
  p_rf <- predict(m_rf, dte, type = "prob")[, "ALT-high"]

  cvg  <- suppressWarnings(
    cv.glmnet(X[s$train, ], y[s$train], alpha = 1, family = "binomial", nfolds = 10)
  )
  p_la <- as.numeric(predict(cvg, X[s$test, ], s = "lambda.1se", type = "response"))

  in_r_metrics <- bind_rows(in_r_metrics,
    fold_metrics(p_lr, dte$alt_status) %>% mutate(model = "Logistic regression", rep = s$rep, fold = s$fold),
    fold_metrics(p_rf, dte$alt_status) %>% mutate(model = "Random forest",       rep = s$rep, fold = s$fold),
    fold_metrics(p_la, dte$alt_status) %>% mutate(model = "LASSO",               rep = s$rep, fold = s$fold)
  )
}
```

# Export the cohort + folds for the TabPFN out-of-fold run

TabPFN is run on the GPU box (carmela), not here. We export the design matrix, the
labels and the 5×5 fold assignment so the standalone script reproduces these *exact*
folds and writes back per-sample out-of-fold probabilities.

```{r export-tabpfn-oof}
cmp_dir <- here("outputs", "shap_input_cmp_cv")
dir.create(cmp_dir, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(as.data.frame(X),                          file.path(cmp_dir, "X_full.csv"))
readr::write_csv(data.frame(y = as.integer(y == "ALT-high")), file.path(cmp_dir, "y_full.csv"))
readr::write_csv(as.data.frame(fold_mat),                   file.path(cmp_dir, "folds_5x5.csv"))

cat(sprintf("Exported %d x %d to %s\n", nrow(X), ncol(X), cmp_dir))
cat("Then on carmela:  python tabpfn_cv_oof.py --input-dir shap_input_cmp_cv \\\n")
cat("                    --output outputs/tabpfn_cv_oof.csv\n")
```

# Assemble TabPFN per-fold metrics

If the out-of-fold probabilities are present (`outputs/tabpfn_cv_oof.csv`), TabPFN is
scored with the **same** `fold_metrics()` as every other model (so PR-AUC and all
thresholded metrics are perfectly consistent). Otherwise we fall back to the cached
results from notebook 10 (which lack per-sample probabilities, so PR-AUC shows as
*pending*).

```{r tabpfn-metrics}
oof_csv <- here("outputs", "tabpfn_cv_oof.csv")
if (file.exists(oof_csv)) {
  oof <- readr::read_csv(oof_csv, show_col_types = FALSE)
  tabpfn_metrics <- oof %>%
    mutate(truth = factor(ifelse(y == 1, "ALT-high", "ALT-low"),
                          levels = c("ALT-low", "ALT-high"))) %>%
    group_by(rep, fold) %>%
    group_modify(~ fold_metrics(.x$yhat, .x$truth)) %>%
    ungroup() %>%
    mutate(model = "TabPFN")
  tabpfn_source <- "out-of-fold probabilities (carmela)"
} else {
  tabpfn_metrics <- readRDS(here("outputs", "tabpfn_cv_results.rds"))$per_fold_metrics %>%
    transmute(
      auc, pr_auc = NA_real_, sensitivity, specificity,
      precision = ifelse(tp + fp > 0, tp / (tp + fp), NA_real_),
      f1        = ifelse(2 * tp + fp + fn > 0, 2 * tp / (2 * tp + fp + fn), NA_real_),
      balanced_accuracy = 0.5 * (sensitivity + specificity),
      model = "TabPFN", rep = rep(1:5, each = 5), fold = rep(1:5, times = 5)
    )
  tabpfn_source <- "cached metrics (notebook 10) — PR-AUC pending re-run"
}
cat("TabPFN source:", tabpfn_source, "\n")
```

```{r combine-verify}
perfold <- bind_rows(
  in_r_metrics %>% select(auc, pr_auc, sensitivity, specificity, precision, f1, balanced_accuracy, model, rep, fold),
  tabpfn_metrics %>% select(auc, pr_auc, sensitivity, specificity, precision, f1, balanced_accuracy, model, rep, fold)
)

# fold-alignment check: cached TabPFN test-set class counts vs ours
chk_lr <- in_r_metrics %>% filter(model == "Logistic regression") %>% arrange(rep, fold)
chk_tp <- readRDS(here("outputs", "tabpfn_cv_results.rds"))$per_fold_metrics %>%
  mutate(n_high = tp + fn, n_low = tn + fp)
cat("Folds identical to cached TabPFN across all 25 folds: ",
    all(chk_lr$n_high == chk_tp$n_high) && all(chk_lr$n_low == chk_tp$n_low), "\n")
```

# Results

```{r summary-table}
summ <- perfold %>%
  group_by(model) %>%
  summarise(across(c(auc, pr_auc, sensitivity, specificity, precision, f1, balanced_accuracy),
                   list(m = ~mean(.x, na.rm = TRUE), s = ~sd(.x, na.rm = TRUE))),
            .groups = "drop") %>%
  arrange(desc(auc_m))

fmt <- function(m, s) ifelse(is.na(m), "pending", sprintf("%.3f ± %.3f", m, s))
tab <- summ %>% transmute(
  Model = model,
  AUC = fmt(auc_m, auc_s), `PR-AUC` = fmt(pr_auc_m, pr_auc_s),
  Sensitivity = fmt(sensitivity_m, sensitivity_s), Specificity = fmt(specificity_m, specificity_s),
  Precision = fmt(precision_m, precision_s), F1 = fmt(f1_m, f1_s),
  `Balanced acc.` = fmt(balanced_accuracy_m, balanced_accuracy_s)
)
readr::write_csv(tab, here("outputs", "classification_comparison_table.csv"))
kable(tab, caption = "ALT-status classification — repeated 5x5 stratified CV (mean ± SD over 25 folds, n = 817, 14 features). Thresholded metrics at 0.5; positive class = ALT-high.")
```

```{r boxplot, fig.width=9, fig.height=4.5}
okabe <- c("#0072B2", "#E69F00", "#009E73", "#D55E00", "#CC79A7")
ord_plot <- summ %>% arrange(auc_m) %>% pull(model)

plot_df <- perfold %>%
  mutate(model = factor(model, levels = ord_plot)) %>%
  select(model, rep, fold, AUC = auc, `PR-AUC` = pr_auc) %>%
  pivot_longer(c(AUC, `PR-AUC`), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value))

p <- ggplot(plot_df, aes(model, value, fill = model)) +
  geom_boxplot(alpha = .6, outlier.shape = NA, width = .6) +
  geom_jitter(width = .12, size = 1, alpha = .5) +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = okabe) +
  coord_flip() +
  labs(
    title    = "ALT-status classification - repeated 5x5 stratified CV (n = 817, 14 features)",
    subtitle = "Per-fold distribution over 25 folds; identical folds shared across all models",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none", plot.title = element_text(face = "bold", size = 12))

ggsave(here("outputs", "classification_comparison.pdf"), p, width = 9, height = 4.5)
ggsave(here("outputs", "classification_comparison.png"), p, width = 9, height = 4.5, dpi = 150)
p
```

# Reading the comparison

- **AUC (threshold-free) is the fair headline.** TabPFN ranks ALT-high vs ALT-low
  best (≈ 0.996), ahead of random forest and LASSO (≈ 0.98) and logistic regression
  (≈ 0.94). Because the folds are shared, these differences are directly paired.
- **At a 0.5 threshold the imbalance bites.** Tree/LASSO models reach very high
  specificity and precision but lower sensitivity (they miss a chunk of the rare
  ALT-high class), whereas logistic regression keeps the highest sensitivity /
  balanced accuracy. This is a threshold artefact, not a ranking difference — hence
  the emphasis on AUC and **PR-AUC** (average precision), which summarise performance
  across all thresholds and are the honest metric at 6 % prevalence.
- **Single-split numbers in notebooks 02/09/10 are optimistic** (e.g. LR's 0.998 on
  one split vs 0.937 under CV) and should not be the figure quoted in the thesis; the
  CV estimates above supersede them.
