---
title: "21_regression_comparison"
subtitle: "Fair head-to-head comparison of TF-rate regressors"
author: "Andrea Polazzi"
date: today
format:
  html:
    toc: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

# Purpose

The single fair comparison of every regressor trained for the two
telomere-fusion-rate targets:

-   **tTF** = `tf_primary_rate` (tumour TF rate), mirroring notebook
    13_1 — strong outlier `tf_primary_rate >= 10` removed.
-   **bTF** = `tf_blood_rate` (blood/germline TF rate), mirroring
    notebook 15_1 — no outlier filter.

Earlier regression notebooks each used an **80/20 single split** plus a
5×5 CV built with `createFolds()` **on the training split only**, while
the TabPFN regression (old notebook 10) used a 5×5 CV on the **full**
data — so the numbers were not comparable. Here every model is put on
identical footing:

-   **Same cohort** per target (complete cases of `PCAWG_primary.xlsx`,
    log1p target + skewed features, `alt_status` kept as a predictor).
-   **Same features**: the full feature set (the other TF rate, telomere
    content / insertion, TERT, the 9 TVR singletons, ALT status).
-   **Same resampling**: the shared **5-fold × 5-repeat stratified CV**
    from `scripts/make_folds.R` (`seed = 21`, stratified on
    `alt_status`) — the same folds used by the classification comparison
    and the SHAP runs. No held-out single split.
-   **Same metrics, same code** (`scripts/reg_metrics.R`): RMSE, MAE,
    Spearman ρ, all in the model's log1p working space.

TabPFN is **not** called through the cloud API: it runs on the GPU box
(carmela) with the local `tabpfn` package via
`scripts/tabpfn_cv_oof.py --task regression`, which writes per-sample
out-of-fold predictions that this notebook scores with the same
`reg_fold_metrics()`.

```{r setup}
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(forcats)
library(here)
library(randomForest)
library(glmnet)
library(knitr)

source(here('scripts', 'make_folds.R'))
source(here('scripts', 'reg_metrics.R'))

raw <- read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))

# cohort prep mirroring notebooks 13_1 (tTF) and 15_1 (bTF)
prep_cohort <- function(filter_outlier) {
  d <- raw %>% drop_na()
  if (filter_outlier) d <- d %>% filter(tf_primary_rate < 10)
  d %>%
    mutate(
      tf_primary_rate         = log1p(tf_primary_rate),
      tf_blood_rate           = log1p(tf_blood_rate),
      telomere_insertion_rate = log1p(telomere_insertion_rate),
      TERT_FPKM               = log1p(TERT_FPKM),
      alt_status              = factor(alt_status, levels = c('ALT-low', 'ALT-high'))
    ) %>%
    select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)
}
```

```{r run-target}
# Train LM / LASSO / RF over the shared folds, export the cohort for the TabPFN HPC
# run, and read TabPFN's out-of-fold predictions back if present. Returns per-fold
# metrics (long) for one target.
run_target <- function(target, filter_outlier, oof_name, cmp_subdir) {
  dat <- prep_cohort(filter_outlier)
  X   <- model.matrix(reformulate('.', target), dat)[, -1]   # numeric design (alt_status -> dummy)
  y   <- dat[[target]]

  fold_mat  <- make_folds(dat$alt_status, k = 5, n_repeats = 5, seed = 21)
  cv_splits <- folds_to_splits(fold_mat)

  res <- tibble()
  for (s in cv_splits) {
    dtr <- dat[s$train, ]; dte <- dat[s$test, ]
    p_lm <- predict(lm(reformulate('.', target), dtr), dte)
    cvg  <- cv.glmnet(X[s$train, ], y[s$train], alpha = 1)
    p_la <- as.numeric(predict(cvg, X[s$test, ], s = 'lambda.min'))
    p_rf <- predict(randomForest(reformulate('.', target), data = dtr, ntree = 500), dte)
    res <- bind_rows(res,
      reg_fold_metrics(p_lm, y[s$test]) %>% mutate(model = 'LM',    rep = s$rep, fold = s$fold),
      reg_fold_metrics(p_la, y[s$test]) %>% mutate(model = 'LASSO', rep = s$rep, fold = s$fold),
      reg_fold_metrics(p_rf, y[s$test]) %>% mutate(model = 'RF',    rep = s$rep, fold = s$fold))
  }

  # export cohort + folds for the TabPFN HPC run
  cmp_dir <- here('outputs', cmp_subdir)
  dir.create(cmp_dir, recursive = TRUE, showWarnings = FALSE)
  readr::write_csv(as.data.frame(X),        file.path(cmp_dir, 'X_full.csv'))
  readr::write_csv(data.frame(y = y),       file.path(cmp_dir, 'y_full.csv'))
  readr::write_csv(as.data.frame(fold_mat), file.path(cmp_dir, 'folds_5x5.csv'))

  # TabPFN out-of-fold predictions, if the HPC run is done
  oof_csv <- here('outputs', oof_name)
  if (file.exists(oof_csv)) {
    tp <- readr::read_csv(oof_csv, show_col_types = FALSE) %>%
      group_by(rep, fold) %>%
      group_modify(~ reg_fold_metrics(.x$yhat, .x$y)) %>%
      ungroup() %>% mutate(model = 'TabPFN')
    res <- bind_rows(res, tp)
  }
  res %>% mutate(target = target)
}

ttf <- run_target('tf_primary_rate', TRUE,  'tabpfn_reg_ttf_oof.csv', 'cmp_reg_ttf_cv')
btf <- run_target('tf_blood_rate',   FALSE, 'tabpfn_reg_btf_oof.csv', 'cmp_reg_btf_cv')
perfold <- bind_rows(ttf, btf) %>%
  mutate(target = recode(target, tf_primary_rate = 'tTF (tumour)', tf_blood_rate = 'bTF (blood)'))
```

# Results

```{r summary-table}
summ <- perfold %>%
  group_by(target, model) %>%
  summarise(across(c(rmse, mae, spearman),
                   list(m = ~mean(.x, na.rm = TRUE), s = ~sd(.x, na.rm = TRUE))),
            .groups = 'drop') %>%
  arrange(target, rmse_m)

fmt <- function(m, s) sprintf('%.3f ± %.3f', m, s)
tab <- summ %>% transmute(
  Target = target, Model = model,
  RMSE = fmt(rmse_m, rmse_s), MAE = fmt(mae_m, mae_s),
  `Spearman ρ` = fmt(spearman_m, spearman_s)
)
readr::write_csv(tab, here('outputs', 'regression_comparison_table.csv'))
kable(tab, caption = 'TF-rate regression — shared 5x5 stratified CV (mean ± SD over 25 folds, log1p space). TabPFN rows appear once scripts/tabpfn_cv_oof.py --task regression has been run on carmela.')
```

```{r boxplot, fig.width=9, fig.height=6}
okabe <- c('#0072B2', '#E69F00', '#009E73', '#D55E00')
plot_df <- perfold %>%
  pivot_longer(c(rmse, mae, spearman), names_to = 'metric', values_to = 'value') %>%
  mutate(metric = recode(metric, rmse = 'RMSE', mae = 'MAE', spearman = 'Spearman ρ'),
         model  = factor(model, levels = c('LM', 'LASSO', 'RF', 'TabPFN')))

p <- ggplot(plot_df, aes(model, value, fill = model)) +
  geom_boxplot(alpha = .6, outlier.shape = NA, width = .6) +
  geom_jitter(width = .12, size = .8, alpha = .4) +
  facet_grid(target ~ metric, scales = 'free') +
  scale_fill_manual(values = okabe) +
  coord_flip() +
  labs(title = 'TF-rate regression - shared 5x5 stratified CV (log1p space)',
       subtitle = 'Per-fold distribution over 25 folds; identical folds shared across all models',
       x = NULL, y = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = 'none', plot.title = element_text(face = 'bold', size = 12))

ggsave(here('outputs', 'regression_comparison.pdf'), p, width = 9, height = 6)
ggsave(here('outputs', 'regression_comparison.png'), p, width = 9, height = 6, dpi = 150)
p
```

# Reading the comparison

-   **tTF (tumour TF rate) is moderately predictable** (Spearman ρ ≈
    0.4); tree/linear models land close together on RMSE/MAE in log1p
    space.
-   **bTF (blood TF rate) is only weakly predictable** (Spearman ρ ≈
    0.2). LASSO frequently shrinks to an intercept-only fit on bTF folds
    (constant prediction → undefined ρ), which is itself evidence of
    weak signal.
-   The TabPFN rows fill in after the carmela run; because the folds are
    shared, the comparison is paired and the boxplots are directly
    comparable fold-by-fold.
-   The zero-inflated two-part model and the SHAP interpretability stay
    in notebooks 13_1 / 15_1; this notebook is only the head-to-head
    predictive comparison.
