---
title: "22_tTF_genes"
date: today
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

# Setup

```{r setup}
library(tidyverse)
library(here)
library(broom)
library(lme4)
library(patchwork)
library(scales)

tidy_fixef <- function(model, conf.int = FALSE, conf.level = 0.95) {
  s <- summary(model)$coefficients
  out <- tibble(
    term      = rownames(s),
    estimate  = s[, 'Estimate'],
    std.error = s[, 'Std. Error'],
    statistic = s[, grep('value', colnames(s))[1]],
    p.value   = if ('Pr(>|z|)' %in% colnames(s)) s[, 'Pr(>|z|)'] else NA_real_
  )
  if (conf.int) {
    z <- qnorm(1 - (1 - conf.level) / 2)
    out$conf.low  <- out$estimate - z * out$std.error
    out$conf.high <- out$estimate + z * out$std.error
  }
  out
}

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary+genes2.xlsx'))

# OR4F21 is absent from the PCAWG driver mutations dataset (TableS3) and is excluded.
gene_cols <- c('ATRX', 'EGFR', 'CDKN2A', 'APC', 'KIT', 'MET', 'RB1', 'STK11')

data <- data %>%
  mutate(
    across(all_of(gene_cols), as.logical),
    cancer_group = factor(cancer_group),
    cancer_type  = factor(cancer_type)
  )

cat('Dataset:', nrow(data), 'patients,', ncol(data), 'columns\n')
cat('tTF rate range: [', round(min(data$tf_primary_rate, na.rm = TRUE), 3),
    ',', round(max(data$tf_primary_rate, na.rm = TRUE), 3), ']\n')
cat('Patients with tTF = 0:', sum(data$tf_primary_rate == 0, na.rm = TRUE), '\n')
```

This notebook tests whether higher tumor telomere fusion rates (`tf_primary_rate`) are
associated with mutations in `r paste(gene_cols, collapse = ', ')`.
Gene columns were derived from the PCAWG panorama driver mutation table (TableS3).
OR4F21 is not present in that dataset and is excluded from all analyses.
ATRX reflects ATRX-only mutations; the merged ATRX/DAXX indicator used in earlier
notebooks (`ATRX_DAXX_trunc`) is kept separate and is not used here.

# Descriptive analysis

## Mutation frequencies

```{r freq-table}
na_mask <- is.na(data[[gene_cols[1]]])

freq_tbl <- tibble(
  gene    = gene_cols,
  n_TRUE  = sapply(gene_cols, function(g) sum(data[[g]], na.rm = TRUE)),
  n_FALSE = sapply(gene_cols, function(g) sum(!data[[g]], na.rm = TRUE)),
  n_NA    = sapply(gene_cols, function(g) sum(is.na(data[[g]]))),
  pct_TRUE = round(100 * n_TRUE / (n_TRUE + n_FALSE), 2)
)
knitr::kable(freq_tbl, caption = 'Gene mutation frequencies')
```

## Missingness pattern

```{r missingness}
cat('Patients with no gene annotation:', sum(na_mask), '\n')

na_by_type <- data %>%
  mutate(missing_genes = na_mask) %>%
  group_by(cancer_type) %>%
  summarise(n = n(), n_missing = sum(missing_genes),
            pct_missing = round(100 * n_missing / n, 1)) %>%
  filter(n_missing > 0) %>%
  arrange(desc(pct_missing))
knitr::kable(na_by_type, caption = 'Cancer types with missing gene annotations')
```

## tTF distribution by mutation status

```{r ttf-boxplots, fig.width=12, fig.height=7}
plot_data <- data %>%
  filter(!na_mask) %>%
  select(tf_primary_rate, all_of(gene_cols)) %>%
  pivot_longer(all_of(gene_cols), names_to = 'gene', values_to = 'mutated') %>%
  mutate(
    mutated = ifelse(mutated, 'Mutated', 'Wild-type'),
    ttf_log1p = log1p(tf_primary_rate)
  )

ggplot(plot_data, aes(mutated, ttf_log1p, fill = mutated)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, outlier.size = 0.5, fill = 'white', alpha = 0.8) +
  facet_wrap(~ gene, scales = 'free_x', nrow = 2) +
  scale_fill_manual(values = c('Mutated' = '#e07b54', 'Wild-type' = '#5b8db8')) +
  labs(title = 'log1p(tTF rate) by mutation status',
       x = NULL, y = 'log1p(tf_primary_rate)', fill = NULL) +
  theme_bw() +
  theme(legend.position = 'bottom', strip.text = element_text(face = 'bold'))
```

# Univariate associations with tTF rate

Wilcoxon rank-sum test on the raw tTF rate (zero-inflated, non-parametric approach).

```{r wilcoxon-ttf}
wx_ttf <- map_dfr(gene_cols, function(g) {
  d <- data %>% filter(!is.na(.data[[g]]), !is.na(tf_primary_rate))
  wt <- wilcox.test(d$tf_primary_rate ~ d[[g]], exact = FALSE)
  med_mut <- median(d$tf_primary_rate[d[[g]]], na.rm = TRUE)
  med_wt  <- median(d$tf_primary_rate[!d[[g]]], na.rm = TRUE)
  tibble(
    gene          = g,
    n_mutated     = sum(d[[g]]),
    median_mut    = round(med_mut, 4),
    median_wt     = round(med_wt, 4),
    W             = wt$statistic,
    p_value       = wt$p.value
  )
}) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = 'BH'),
         p_value  = format.pval(p_value,  digits = 2),
         p_adj_BH = format.pval(p_adj_BH, digits = 2))

knitr::kable(wx_ttf, caption = 'Wilcoxon rank-sum: gene mutation vs tTF rate')
```

## Cancer-type confounding check

```{r fisher-group}
fisher_group <- map_dfr(gene_cols, function(g) {
  d <- data %>% filter(!is.na(.data[[g]]))
  tbl <- table(d[[g]], d$cancer_group)
  if (any(dim(tbl) < 2)) return(NULL)
  ft <- fisher.test(tbl)
  tibble(gene = g, p_value = ft$p.value)
}) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = 'BH'),
         p_value  = format.pval(p_value,  digits = 2),
         p_adj_BH = format.pval(p_adj_BH, digits = 2))
knitr::kable(fisher_group,
             caption = 'Fisher exact: gene mutation vs cancer_group (confounding check)')
```

# Adjusted models: gene effect on tTF beyond cancer type

Outcome: `log1p(tf_primary_rate)`. We test each gene one at a time via likelihood ratio
test, adjusting first for `cancer_group` (fixed) and then for `cancer_type` as a random
intercept. Genes with fewer than 20 mutated cases are flagged but excluded from modelling.

```{r select-genes}
genes_used <- freq_tbl %>%
  filter(n_TRUE >= 20) %>%
  pull(gene)

cat('Genes with n_TRUE >= 20 (used in adjusted models):',
    paste(genes_used, collapse = ', '), '\n')
cat('Genes excluded (n_TRUE < 20):',
    paste(setdiff(gene_cols, genes_used), collapse = ', '), '\n')
```

## Adjusted by `cancer_group` (fixed effect)

```{r lrt-cancer-group}
d_cc <- data %>%
  filter(!na_mask) %>%
  mutate(ttf = log1p(tf_primary_rate)) %>%
  drop_na(ttf)

lrt_group <- map_dfr(genes_used, function(g) {
  m0 <- lm(ttf ~ cancer_group, data = d_cc)
  fm <- as.formula(paste0('ttf ~ cancer_group + ', g))
  m1 <- lm(fm, data = d_cc)
  ftest <- anova(m0, m1)
  coef  <- tidy(m1, conf.int = TRUE) %>% filter(term == paste0(g, 'TRUE'))
  tibble(
    gene      = g,
    estimate  = round(coef$estimate, 3),
    conf.low  = round(coef$conf.low, 3),
    conf.high = round(coef$conf.high, 3),
    F_stat    = round(ftest$F[2], 3),
    p_value   = ftest$`Pr(>F)`[2]
  )
}) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = 'BH'),
         p_value  = format.pval(p_value,  digits = 2),
         p_adj_BH = format.pval(p_adj_BH, digits = 2))

knitr::kable(lrt_group,
             caption = 'F-test vs baseline (cancer_group only). Effect on log1p(tTF).')
```

## Adjusted by `cancer_type` (mixed effects)

```{r lrt-mixed}
lrt_mixed <- map_dfr(genes_used, function(g) {
  m0 <- lmer(ttf ~ (1 | cancer_type), data = d_cc, REML = FALSE)
  fm <- as.formula(paste0('ttf ~ ', g, ' + (1 | cancer_type)'))
  m1 <- lmer(fm, data = d_cc, REML = FALSE)
  lr <- anova(m0, m1)
  coef <- tidy_fixef(m1) %>% filter(term == paste0(g, 'TRUE'))
  tibble(
    gene      = g,
    estimate  = round(coef$estimate, 3),
    std.error = round(coef$std.error, 3),
    LRT_chisq = round(lr$Chisq[2], 3),
    LRT_p     = lr$`Pr(>Chisq)`[2]
  )
}) %>%
  mutate(LRT_p_adj_BH = p.adjust(LRT_p, method = 'BH'),
         LRT_p        = format.pval(LRT_p,        digits = 2),
         LRT_p_adj_BH = format.pval(LRT_p_adj_BH, digits = 2))

knitr::kable(lrt_mixed,
             caption = 'LRT vs baseline (random intercept for cancer_type). Effect on log1p(tTF).')
```

# Joint multivariate model

All viable genes simultaneously, with `cancer_type` as random effect.

```{r joint-model}
joint_formula <- as.formula(
  paste0('ttf ~ ', paste(genes_used, collapse = ' + '), ' + (1 | cancer_type)')
)
m_joint <- lmer(joint_formula, data = d_cc, REML = FALSE)

joint_tbl <- tidy_fixef(m_joint, conf.int = TRUE) %>%
  filter(term != '(Intercept)') %>%
  mutate(conf.low  = round(conf.low,  3),
         conf.high = round(conf.high, 3),
         estimate  = round(estimate,  3),
         p.value   = format.pval(p.value, digits = 2)) %>%
  select(term, estimate, conf.low, conf.high, std.error, p.value)

knitr::kable(joint_tbl, digits = 3,
             caption = 'Joint model: all viable genes + cancer_type random effect (outcome: log1p(tTF))')
```

```{r forest-plot, fig.width=7, fig.height=4}
fp <- tidy_fixef(m_joint, conf.int = TRUE) %>%
  filter(term != '(Intercept)') %>%
  mutate(term = str_remove(term, 'TRUE$'))

ggplot(fp, aes(estimate, fct_reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey50') +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(title = 'Gene effects on log1p(tTF rate)',
       subtitle = 'Joint model with cancer_type random effect (95% Wald CI)',
       x = 'Coefficient (log1p scale)', y = NULL) +
  theme_bw()
```

# Stratified analysis within cancer group

```{r stratified}
stratified <- map_dfr(genes_used, function(g) {
  map_dfr(levels(d_cc$cancer_group), function(grp) {
    sub <- d_cc %>% filter(cancer_group == grp)
    n_mut <- sum(sub[[g]], na.rm = TRUE)
    if (n_mut < 5 || sum(!sub[[g]], na.rm = TRUE) < 5) {
      return(tibble(gene = g, group = grp, n_mutated = n_mut,
                    median_mut = NA_real_, median_wt = NA_real_, p_value = NA_real_))
    }
    wt <- wilcox.test(sub$tf_primary_rate ~ sub[[g]], exact = FALSE)
    tibble(
      gene      = g,
      group     = grp,
      n_mutated = n_mut,
      median_mut = round(median(sub$tf_primary_rate[sub[[g]]], na.rm = TRUE), 4),
      median_wt  = round(median(sub$tf_primary_rate[!sub[[g]]], na.rm = TRUE), 4),
      p_value    = wt$p.value
    )
  })
}) %>%
  mutate(p_value = ifelse(is.na(p_value), NA, format.pval(p_value, digits = 2)))

knitr::kable(stratified,
             caption = 'Stratified Wilcoxon tests within each cancer_group')
```

# Sensitivity: impute NA as FALSE

```{r sensitivity}
d_imp <- data %>%
  mutate(across(all_of(gene_cols), ~ replace_na(.x, FALSE)),
         ttf = log1p(tf_primary_rate)) %>%
  drop_na(ttf)

sens <- map_dfr(genes_used, function(g) {
  m0 <- lm(ttf ~ cancer_group, data = d_imp)
  fm <- as.formula(paste0('ttf ~ cancer_group + ', g))
  m1 <- lm(fm, data = d_imp)
  ftest <- anova(m0, m1)
  coef  <- tidy(m1, conf.int = TRUE) %>% filter(term == paste0(g, 'TRUE'))
  tibble(
    gene      = g,
    estimate  = round(coef$estimate, 3),
    p_value   = ftest$`Pr(>F)`[2]
  )
}) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = 'BH'),
         p_value  = format.pval(p_value,  digits = 2),
         p_adj_BH = format.pval(p_adj_BH, digits = 2))

knitr::kable(sens, caption = 'Sensitivity: NA imputed as FALSE')
```

# Summary

-   OR4F21 is absent from the PCAWG driver mutation table and is excluded throughout.
-   The table above shows which of the eight genes (ATRX, EGFR, CDKN2A, APC, KIT, MET, RB1, STK11) have sufficient prevalence (n ≥ 20 mutated cases) for stable regression.
-   The Wilcoxon tests and adjusted linear models (log1p tTF ~ gene + cancer_type random effect) identify which mutations carry a statistically independent signal for higher or lower tTF rates beyond cancer-type composition.
-   The forest plot summarises effect direction and uncertainty for the joint multivariate model.
-   The stratified analysis indicates whether any gene–tTF association is specific to mesenchymal vs non-mesenchymal tumours.
-   If conclusions are consistent between complete-case and NA-imputed analyses, missingness is not driving the results.
