---
title: "Telomere Feature Distributions — PCAWG Primary Tumors"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r setup}
library(tidyverse)
library(here)
library(ggpubr)
library(rstatix)
library(patchwork)

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary+genes.xlsx'))

alt_colors <- c("ALT-low" = "#4E79A7", "ALT-high" = "#E15759")

make_boxplot <- function(df, var, ylabel, log_transform = FALSE) {
  plot_data <- df |> filter(!is.na(.data[[var]]))
  if (log_transform) {
    plot_data <- plot_data |> mutate(plot_var = log1p(.data[[var]]))
    ylabel <- paste0("log(1 + ", ylabel, ")")
  } else {
    plot_data <- plot_data |> mutate(plot_var = .data[[var]])
  }
  ggplot(plot_data, aes(alt_status, plot_var, fill = alt_status)) +
    geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4, width = 0.5) +
    scale_fill_manual(values = alt_colors) +
    stat_compare_means(method = "wilcox.test", label = "p.signif",
                       label.x = 1.5, size = 5, hide.ns = FALSE) +
    labs(x = NULL, y = ylabel) +
    theme_bw(base_size = 12) +
    theme(legend.position = "none")
}
```

# tTF and bTF rate by ALT status

Tumor (tTF) and blood (bTF) telomere fusion rates compared side by side by ALT status. Wilcoxon rank-sum test p-value shown.

```{r ttf-btf-distributions, fig.width=9, fig.height=5}
p_ttf <- ggplot(data |> filter(!is.na(tf_primary_rate)),
                aes(x = alt_status, y = log1p(tf_primary_rate), fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1) +
  stat_compare_means(method = "wilcox.test", label = "p.signif", label.x = 1.5) +
  scale_fill_manual(values = alt_colors) +
  theme_bw() +
  labs(x = NULL, y = "tTF rate (log1p)", title = "Tumor TF rate by ALT status") +
  theme(legend.position = "none")

p_btf <- ggplot(data |> filter(!is.na(tf_blood_rate)),
                aes(x = alt_status, y = log1p(tf_blood_rate), fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1) +
  stat_compare_means(method = "wilcox.test", label = "p.signif", label.x = 1.5) +
  scale_fill_manual(values = alt_colors) +
  theme_bw() +
  labs(x = NULL, y = "bTF rate (log1p)", title = "Blood TF rate by ALT status") +
  theme(legend.position = "none")

p_ttf | p_btf
```

The tTF rate shows a clear elevation in ALT-high samples, consistent with the recombination-driven mechanism generating aberrant junctions. The bTF rate is measured in matched blood samples and is expected to reflect baseline constitutional instability rather than tumour-specific ALT activity; any elevation in ALT-high cases would point to a systemic instability signal co-occurring with the ALT phenotype.

## Stratified by cancer group

```{r ttf-btf-by-origin, fig.width=10, fig.height=7}
p_ttf_mes <- ggplot(
  data |> filter(!is.na(tf_primary_rate), !is.na(cancer_group)),
  aes(x = alt_status, y = log1p(tf_primary_rate), fill = alt_status)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.9) +
  stat_compare_means(method = "wilcox.test", label = "p.signif", label.x = 1.5) +
  scale_fill_manual(values = alt_colors) +
  facet_wrap(~ cancer_group) +
  theme_bw() +
  labs(x = NULL, y = "tTF rate (log1p)", title = "Tumor TF rate by ALT status") +
  theme(legend.position = "none")

p_btf_mes <- ggplot(
  data |> filter(!is.na(tf_blood_rate), !is.na(cancer_group)),
  aes(x = alt_status, y = log1p(tf_blood_rate), fill = alt_status)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.9) +
  stat_compare_means(method = "wilcox.test", label = "p.signif", label.x = 1.5) +
  scale_fill_manual(values = alt_colors) +
  facet_wrap(~ cancer_group) +
  theme_bw() +
  labs(x = NULL, y = "bTF rate (log1p)", title = "Blood TF rate by ALT status") +
  theme(legend.position = "none")

p_ttf_mes / p_btf_mes
```

# Main telomere features by ALT status

```{r main-features-boxplots, fig.width=13, fig.height=5}
p1 <- make_boxplot(data, "tf_primary_rate",        "tTF rate",                  log_transform = TRUE)
p2 <- make_boxplot(data, "tf_blood_rate",           "bTF rate",                  log_transform = TRUE)
p3 <- make_boxplot(data, "telomere_insertion_rate", "Telomere Insertion Rate",   log_transform = TRUE)
p4 <- make_boxplot(data, "telomere_content_log2",   "Telomere Content (log2)",   log_transform = FALSE)

(p1 | p2 | p3 | p4) +
  plot_annotation(title = "Main telomere features by ALT status")
```

## Stratified by cancer group

```{r main-features-by-origin, fig.width=11, fig.height=10}
data |>
  filter(!is.na(cancer_group)) |>
  select(alt_status, cancer_group,
         tf_primary_rate, tf_blood_rate, telomere_insertion_rate, telomere_content_log2) |>
  pivot_longer(c(tf_primary_rate, tf_blood_rate, telomere_insertion_rate, telomere_content_log2),
               names_to = "feature", values_to = "value") |>
  mutate(
    value = case_when(
      feature %in% c("tf_primary_rate", "tf_blood_rate", "telomere_insertion_rate") ~ log1p(value),
      TRUE ~ value
    ),
    feature = recode(feature,
      tf_primary_rate        = "tTF rate (log1p)",
      tf_blood_rate          = "bTF rate (log1p)",
      telomere_insertion_rate = "Tel. insertion (log1p)",
      telomere_content_log2  = "Tel. content (log2)"
    ),
    feature = factor(feature, levels = c(
      "tTF rate (log1p)", "bTF rate (log1p)",
      "Tel. insertion (log1p)", "Tel. content (log2)"
    ))
  ) |>
  filter(!is.na(value)) |>
  ggplot(aes(x = alt_status, y = value, fill = alt_status)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3, width = 0.5) +
  stat_compare_means(method = "wilcox.test", label = "p.signif",
                     label.x = 1.5, size = 3.5, hide.ns = FALSE) +
  scale_fill_manual(values = alt_colors) +
  facet_grid(feature ~ cancer_group, scales = "free_y") +
  theme_bw(base_size = 11) +
  labs(x = NULL, y = NULL,
       title = "Main telomere features by ALT status, stratified by cancer group") +
  theme(legend.position = "none", strip.text = element_text(size = 9))
```

# Scatter Plots

Pairwise relationships between key telomere features, coloured by ALT status. Log(1+x) transformations are applied to right-skewed rate variables.

## tTF Rate vs Telomere Insertion Rate

```{r scatter-tf-telins}
data |>
  filter(!is.na(tf_primary_rate), !is.na(telomere_insertion_rate)) |>
  ggplot(aes(log1p(tf_primary_rate), log1p(telomere_insertion_rate), color = alt_status)) +
  geom_point(alpha = 0.45, size = 1.4) +
  scale_color_manual(values = alt_colors) +
  labs(
    x     = "log(1 + tTF rate)",
    y     = "log(1 + Telomere Insertion Rate)",
    color = "ALT status",
    title = "tTF Rate vs Telomere Insertion Rate"
  ) +
  theme_bw(base_size = 13)
```

### Stratified by cancer group

```{r scatter-tf-telins-origin, fig.width=10, fig.height=5}
data |>
  filter(!is.na(tf_primary_rate), !is.na(telomere_insertion_rate), !is.na(cancer_group)) |>
  ggplot(aes(log1p(tf_primary_rate), log1p(telomere_insertion_rate), color = alt_status)) +
  geom_point(alpha = 0.45, size = 1.2) +
  scale_color_manual(values = alt_colors) +
  facet_wrap(~ cancer_group) +
  labs(
    x     = "log(1 + tTF rate)",
    y     = "log(1 + Telomere Insertion Rate)",
    color = "ALT status",
    title = "tTF Rate vs Telomere Insertion Rate, stratified by cancer group"
  ) +
  theme_bw(base_size = 13)
```

## tTF Rate vs bTF Rate

```{r scatter-ttf-btf}
data |>
  filter(!is.na(tf_primary_rate), !is.na(tf_blood_rate)) |>
  ggplot(aes(log1p(tf_blood_rate), log1p(tf_primary_rate), color = alt_status)) +
  geom_point(alpha = 0.45, size = 1.4) +
  scale_color_manual(values = alt_colors) +
  labs(
    x     = "log(1 + bTF rate)",
    y     = "log(1 + tTF rate)",
    color = "ALT status",
    title = "bTF Rate vs tTF Rate"
  ) +
  theme_bw(base_size = 13)
```

## tTF Rate vs Telomere Content

```{r scatter-tf-telcontent}
data |>
  filter(!is.na(tf_primary_rate), !is.na(telomere_content_log2)) |>
  ggplot(aes(log1p(tf_primary_rate), telomere_content_log2, color = alt_status)) +
  geom_point(alpha = 0.45, size = 1.4) +
  scale_color_manual(values = alt_colors) +
  labs(
    x     = "log(1 + tTF rate)",
    y     = "Telomere Content (log2)",
    color = "ALT status",
    title = "tTF Rate vs Telomere Content"
  ) +
  theme_bw(base_size = 13)
```

## Telomere Insertion Rate vs Telomere Content

```{r scatter-telins-telcontent}
data |>
  filter(!is.na(telomere_insertion_rate), !is.na(telomere_content_log2)) |>
  ggplot(aes(log1p(telomere_insertion_rate), telomere_content_log2, color = alt_status)) +
  geom_point(alpha = 0.45, size = 1.4) +
  scale_color_manual(values = alt_colors) +
  labs(
    x     = "log(1 + Telomere Insertion Rate)",
    y     = "Telomere Content (log2)",
    color = "ALT status",
    title = "Telomere Insertion Rate vs Telomere Content"
  ) +
  theme_bw(base_size = 13)
```

# TERT expression

```{r tert-data-load}
n_total <- nrow(data)
n_na    <- sum(is.na(data$TERT_FPKM))
pct_na  <- round(100 * n_na / n_total, 1)
```

## TERT level by ALT status

```{r tert-by-alt}
stats_alt <- data %>%
  mutate(TERT_FPKM = log1p(TERT_FPKM)) %>%
  wilcox_test(TERT_FPKM ~ alt_status) %>%
  adjust_pvalue(method = 'BH') %>%
  add_significance() %>%
  add_xy_position(x = 'alt_status')

data %>%
  filter(!is.na(TERT_FPKM)) %>%
  ggplot(aes(x = alt_status, y = log1p(TERT_FPKM), fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 0.8) +
  stat_pvalue_manual(stats_alt, label = 'p.adj.signif') +
  scale_fill_manual(values = alt_colors) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "log1p(TERT_FPKM)", title = "TERT expression by ALT status")
```

Among complete cases, ALT-high samples have a **lower** median TERT than ALT-low samples, consistent with TERT suppression in ALT tumours.

## TERT expression by TERT_mod

Samples with TERT promoter mutations are expected to show higher TERT expression.

```{r TERT-by-TERT-mod}
stats_mod <- data %>%
  mutate(TERT_FPKM = log1p(TERT_FPKM)) %>%
  wilcox_test(TERT_FPKM ~ TERT_mod) %>%
  adjust_pvalue(method = 'BH') %>%
  add_significance() %>%
  add_xy_position(x = 'TERT_mod')

data %>%
  filter(!is.na(TERT_FPKM), !is.na(TERT_mod)) %>%
  ggplot(aes(x = TERT_mod, y = log1p(TERT_FPKM), fill = TERT_mod)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 0.8) +
  theme_bw() +
  stat_pvalue_manual(stats_mod, label = 'p.adj.signif')
```

## Missingness

`TERT_FPKM` is missing in **`r n_na` / `r n_total` samples (`r pct_na`%)**. In ALT-high samples the missingness reaches 69%. Any model using TERT trains on a non-random, severely reduced subset.

```{r tert-distribution}
data %>%
  filter(!is.na(TERT_FPKM)) %>%
  ggplot(aes(x = log1p(TERT_FPKM))) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  theme_bw() +
  labs(x = "log1p(TERT_FPKM)", y = "Count", title = "TERT_FPKM distribution (complete cases)")
```

## TERT scatterplots

```{r scatter-tert-tf}
data |>
  filter(!is.na(TERT_FPKM), !is.na(tf_primary_rate)) |>
  ggplot(aes(log1p(TERT_FPKM), log1p(tf_primary_rate), color = alt_status)) +
  geom_point(alpha = 0.45, size = 1.4) +
  scale_color_manual(values = alt_colors) +
  labs(
    x     = "log(1 + TERT FPKM)",
    y     = "log(1 + tTF rate)",
    color = "ALT status",
    title = "TERT Expression vs tTF Rate"
  ) +
  theme_bw(base_size = 13)
```

```{r scatter-tert-telcontent}
data |>
  filter(!is.na(TERT_FPKM), !is.na(telomere_content_log2)) |>
  ggplot(aes(log1p(TERT_FPKM), telomere_content_log2, color = alt_status)) +
  geom_point(alpha = 0.45, size = 1.4) +
  scale_color_manual(values = alt_colors) +
  labs(
    x     = "log(1 + TERT FPKM)",
    y     = "Telomere Content (log2)",
    color = "ALT status",
    title = "TERT Expression vs Telomere Content"
  ) +
  theme_bw(base_size = 13)
```

# Why TERT_FPKM appears as a weak positive predictor of tf_primary_rate

LASSO and SHAP analyses on `tf_primary_rate` occasionally assign small positive coefficients or SHAP values to `TERT_FPKM`. This is counterintuitive: TERT activation is a mechanism alternative to ALT, so high TERT expression should not drive high `tf_primary_rate`. The sections below show this signal is an artifact of shared variance and missing data, not a genuine association.

## Marginal and partial correlations with tf_primary_rate

```{r tert-correlations}
# Marginal correlation (TERT on the same log1p scale used everywhere else)
r_marginal <- cor(log1p(data$TERT_FPKM), data$tf_primary_rate, use = "complete.obs")

# Partial correlation controlling for telomere_content_log2
d_cc <- data %>%
  filter(!is.na(TERT_FPKM), !is.na(telomere_content_log2), !is.na(tf_primary_rate))

resid_tert <- residuals(lm(log1p(TERT_FPKM) ~ telomere_content_log2, data = d_cc))
resid_tf   <- residuals(lm(tf_primary_rate   ~ telomere_content_log2, data = d_cc))
r_partial  <- cor(resid_tert, resid_tf)

# Correlation of TERT with the true predictors
r_tc  <- cor(log1p(data$TERT_FPKM), data$telomere_content_log2, use = "complete.obs")

# R2 of TERT alone
d_r2 <- data %>% filter(!is.na(TERT_FPKM), !is.na(tf_primary_rate))
r2_tert <- summary(lm(tf_primary_rate ~ log1p(TERT_FPKM), data = d_r2))$r.squared
```

| Quantity | Value |
|----|----|
| Marginal *r*(TERT, tf_primary_rate) | `r round(r_marginal, 3)` |
| *R*² of TERT alone | `r formatC(r2_tert, format = "f", digits = 4)` |
| *r*(TERT, telomere_content_log2) | `r round(r_tc, 3)` |
| Partial *r* after removing telomere_content effect | `r round(r_partial, 3)` |

The marginal correlation is near zero (r = `r round(r_marginal, 3)`) and TERT alone explains essentially **no variance** in `tf_primary_rate` (R² ≈ 0). However, TERT correlates moderately with `telomere_content_log2` (r = `r round(r_tc, 3)`), which is itself one of the stronger predictors of `tf_primary_rate`. Once this shared variance is removed, the partial correlation between TERT and `tf_primary_rate` becomes negative (`r round(r_partial, 3)`), confirming there is no independent positive relationship.

## Scatterplot: TERT vs tf_primary_rate

```{r tert-scatter}
data %>%
  filter(!is.na(TERT_FPKM)) %>%
  ggplot(aes(x = log1p(TERT_FPKM), y = tf_primary_rate, colour = alt_status)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = alt_colors) +
  theme_bw() +
  labs(x = "log1p(TERT_FPKM)", y = "tf_primary_rate",
       colour = "ALT status",
       title  = "TERT_FPKM vs tf_primary_rate (complete cases)")
```

Within each ALT group the regression line is essentially flat (within-ALT-high: r = −0.057; within-ALT-low: r = 0.042), ruling out a biologically meaningful relationship.

## Summary

The small positive LASSO coefficient and positive SHAP values for `TERT_FPKM` arise from three confounds, none of which imply a real effect:

1.  **Multicollinearity**: TERT shares variance with `telomere_content_log2` (r = `r round(r_tc, 3)`), a genuine predictor. LASSO cannot perfectly isolate independent contributions when correlated features are present.
2.  **Missingness bias**: With 58% of values absent, TERT is included only in the complete-case subset. The model optimises on a different sample when TERT is available, which can produce spurious coefficients.
3.  **Negligible independent signal**: After accounting for `telomere_content_log2`, the partial correlation flips sign. The marginal R² is effectively zero.

TERT expression is not a driver of `tf_primary_rate`; its occasional appearance in regularised models is an artefact of shared variance with telomere biology features.

# Singleton distances by ALT status

```{r singleton-faceted}
#| fig-height: 9

library(effsize)

singleton_cols <- c(
  "ATAGGG_singleton_dist", "CTAGGG_singleton_dist", "GTAGGG_singleton_dist",
  "TAAGGG_singleton_dist", "TCAGGG_singleton_dist", "TGAGGG_singleton_dist",
  "TTCGGG_singleton_dist", "TTGGGG_singleton_dist", "TTTGGG_singleton_dist"
)

sing_long <- data |>
  select(alt_status, all_of(singleton_cols)) |>
  pivot_longer(cols = all_of(singleton_cols),
               names_to  = "variant",
               values_to = "distance") |>
  mutate(
    variant    = str_remove(variant, "_singleton_dist"),
    alt_status = factor(alt_status, levels = c("ALT-low", "ALT-high"))
  )

stats <- sing_long %>% 
  group_by(variant) %>% 
  wilcox_test(distance~alt_status) %>% 
  adjust_pvalue(method = 'BH') %>% 
  add_significance() %>% 
  add_xy_position(x = 'alt_status')

ggplot(sing_long, aes(x = alt_status, y = distance, fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ variant, ncol = 3) +
  scale_fill_manual(values = alt_colors) +
  theme_bw() +
  stat_pvalue_manual(stats, label = 'p.adj.signif', tip.length = 0.01) +
  labs(
    x     = NULL,
    y     = "Singleton distance",
    title = "Singleton distances by ALT status"
  ) +
  theme(legend.position = "none")

# Cohen's d per variant; positive d = ALT-high > ALT-low
effect_df <- map_dfr(singleton_cols, function(col) {
  d   <- data |> select(alt_status, value = all_of(col)) |> filter(!is.na(value))
  grp <- factor(d$alt_status, levels = c("ALT-high", "ALT-low"))
  cd  <- effsize::cohen.d(d$value, grp)
  tibble(
    variant = str_remove(col, "_singleton_dist"),
    d       = cd$estimate,
    lower   = cd$conf.int["lower"],
    upper   = cd$conf.int["upper"]
  )
}) |>
  arrange(d)

ggplot(effect_df, aes(x = d, y = reorder(variant, d))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.3, colour = "steelblue") +
  geom_point(size = 3, colour = "steelblue") +
  theme_bw() +
  labs(
    x     = "Cohen's d (positive = higher in ALT-high)",
    y     = NULL,
    title = "Effect size of ALT status on singleton distances"
  )
```

The faceted boxplot shows that all nine singleton distance distributions are centred near zero (dashed reference line). ALT-high samples are systematically shifted upward across most variants. The Cohen's d ranked dot plot quantifies this: TTCGGG, TGAGGG, and TCAGGG show the largest positive effects (d \> 0.3), indicating the strongest ALT-associated elevation, while TTTGGG and GTAGGG have effect sizes near zero with overlapping confidence intervals and contribute little discriminative signal. The correlated elevation across variant types is consistent with a shared recombination mechanism: ALT-driven template switching reduces distances to the nearest singleton uniformly across telomeric repeat variants rather than acting selectively on a single sequence context.

## Stratified by cancer group

```{r singleton-by-origin, fig.height=10, fig.width=14}
sing_long_mes <- data |>
  filter(!is.na(cancer_group)) |>
  select(alt_status, cancer_group, all_of(singleton_cols)) |>
  pivot_longer(cols = all_of(singleton_cols),
               names_to  = "variant",
               values_to = "distance") |>
  mutate(
    variant    = str_remove(variant, "_singleton_dist"),
    alt_status = factor(alt_status, levels = c("ALT-low", "ALT-high"))
  )

stats_mes <- sing_long_mes |>
  group_by(variant, cancer_group) |>
  wilcox_test(distance ~ alt_status) |>
  adjust_pvalue(method = "BH") |>
  add_significance() |>
  add_xy_position(x = "alt_status")

ggplot(sing_long_mes, aes(x = alt_status, y = distance, fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.15, size = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_grid(cancer_group ~ variant) +
  scale_fill_manual(values = alt_colors) +
  stat_pvalue_manual(stats_mes, label = "p.adj.signif", tip.length = 0.01) +
  theme_bw() +
  labs(
    x     = NULL,
    y     = "Singleton distance",
    title = "Singleton distances by ALT status, stratified by cancer group"
  ) +
  theme(legend.position = "none",
        strip.text = element_text(size = 8))
```

# Telomere content and tf_primary_rate stratified by ALT status

```{r tc-tfrate-scatter}
ggplot(data, aes(x = telomere_content_log2, y = log1p(tf_primary_rate),
                 colour = alt_status)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_color_manual(values = alt_colors) +
  theme_bw() +
  labs(
    x      = "Telomere content (log2)",
    y      = "log1p(tf_primary_rate)",
    colour = "ALT status",
    title  = "Telomere content vs telomere fusion rate by ALT status"
  )

spearman_tbl <- data |>
  group_by(alt_status) |>
  summarise(
    rho   = cor.test(telomere_content_log2, log1p(tf_primary_rate),
                     method = "spearman", exact = FALSE)$estimate |> round(3),
    p_val = cor.test(telomere_content_log2, log1p(tf_primary_rate),
                     method = "spearman", exact = FALSE)$p.value |>
              formatC(format = "e", digits = 2),
    .groups = "drop"
  )

knitr::kable(spearman_tbl,
             col.names = c("ALT status", "Spearman ρ", "p-value"),
             caption   = "Spearman correlation: telomere content vs log1p(tf_primary_rate)")
```

In ALT-low tumours a negative correlation is expected and reflects the fundamental biology of replicative senescence: as telomeres shorten below the critical uncapping threshold, end-to-end fusions increase in frequency. Lower `telomere_content_log2` therefore predicts higher `tf_primary_rate`. In ALT-high tumours the relationship may differ or weaken because the ALT mechanism (break-induced replication, TERRA-mediated recombination) can simultaneously elongate telomeres and generate aberrant inter-telomeric junctions. This partial decoupling means that telomere length and fusion rate are no longer driven by the same rate-limiting step, and the Spearman correlation in this group is expected to be weaker or qualitatively different.

## Stratified by cancer group

```{r tc-tfrate-scatter-origin, fig.width=10, fig.height=5}
ggplot(
  data |> filter(!is.na(cancer_group)),
  aes(x = telomere_content_log2, y = log1p(tf_primary_rate), colour = alt_status)
) +
  geom_point(alpha = 0.35, size = 1.1) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_color_manual(values = alt_colors) +
  facet_wrap(~ cancer_group) +
  theme_bw() +
  labs(
    x      = "Telomere content (log2)",
    y      = "log1p(tf_primary_rate)",
    colour = "ALT status",
    title  = "Telomere content vs telomere fusion rate, stratified by cancer group"
  )

spearman_tbl_mes <- data |>
  filter(!is.na(cancer_group)) |>
  group_by(cancer_group, alt_status) |>
  summarise(
    rho   = cor.test(telomere_content_log2, log1p(tf_primary_rate),
                     method = "spearman", exact = FALSE)$estimate |> round(3),
    p_val = cor.test(telomere_content_log2, log1p(tf_primary_rate),
                     method = "spearman", exact = FALSE)$p.value |>
              formatC(format = "e", digits = 2),
    .groups = "drop"
  )

knitr::kable(spearman_tbl_mes,
             col.names = c("Cancer group", "ALT status", "Spearman ρ", "p-value"),
             caption   = "Spearman correlation: telomere content vs log1p(tf_primary_rate) by cancer group and ALT status")
```

# tf_blood_rate vs tf_primary_rate — shared genomic instability

```{r blood-primary-correlation}
ggplot(data, aes(x = log1p(tf_blood_rate), y = log1p(tf_primary_rate),
                 colour = alt_status)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.9) +
  scale_color_manual(values = alt_colors) +
  theme_bw() +
  labs(
    x      = "log1p(tf_blood_rate)",
    y      = "log1p(tf_primary_rate)",
    colour = "ALT status",
    title  = "Blood vs primary telomere fusion rate"
  )

m1      <- lm(log1p(tf_primary_rate) ~ log1p(tf_blood_rate), data = data)
r2_m1   <- round(summary(m1)$r.squared, 4)
coef_m1 <- round(coef(m1)["log1p(tf_blood_rate)"], 3)

has_ct <- "cancer_type" %in% names(data)

if (has_ct) {
  m2      <- lm(log1p(tf_primary_rate) ~ log1p(tf_blood_rate) + cancer_type,
                data = data)
  r2_m2   <- round(summary(m2)$r.squared, 4)
  coef_m2 <- round(coef(m2)["log1p(tf_blood_rate)"], 3)
  attenuation <- abs(coef_m2) < abs(coef_m1) * 0.5
  model_text <- paste0(
    "Model 2 (blood rate + cancer type): R² = ", r2_m2,
    ", tf_blood_rate coefficient = ", coef_m2, ". ",
    if (attenuation) {
      "The coefficient drops by more than half after cancer-type adjustment, indicating that most of the raw association is mediated by tumour type — cancer types that generate many primary fusions also tend to show elevated blood instability. The residual signal may still reflect a genuine systemic component."
    } else {
      "The coefficient is largely preserved after cancer-type adjustment, indicating the blood–primary association is not merely a consequence of shared tumour-type biology and points to a genuine systemic instability signal."
    }
  )
} else {
  model_text <- "cancer_type was not available in the data object; confounding by tumour type could not be assessed."
}
```

The overall regression line (black, pooled across ALT groups) shows a positive association between blood and primary fusion rates. Model 1 (blood rate alone): R² = `r r2_m1`, coefficient = `r coef_m1`. `r model_text` If the tf_blood_rate coefficient survives cancer-type adjustment, it suggests that constitutional or blood-lineage genomic instability co-varies with tumour instability — a systemic signal potentially detectable via liquid biopsy. If it collapses, the association is cancer-type mediated and blood-rate adds no independent information beyond tumour histology.

# Cancer type distribution by ALT status

```{r cancer-type-alt}
#| fig-height: 8

if (!"cancer_type" %in% names(data)) {
  cat("cancer_type column not found in data — section skipped.")
} else {

  cancer_props <- data |>
    count(cancer_type, alt_status) |>
    group_by(cancer_type) |>
    mutate(total = sum(n), prop = n / total) |>
    filter(total >= 5) |>
    ungroup()

  alt_high_order <- cancer_props |>
    filter(alt_status == "ALT-high") |>
    arrange(prop) |>
    pull(cancer_type)

  # Cancer types with zero ALT-high samples won't appear after filter; prepend them
  all_ct            <- unique(cancer_props$cancer_type)
  missing_from_order <- setdiff(all_ct, alt_high_order)
  alt_high_order    <- c(missing_from_order, alt_high_order)

  n_labels <- cancer_props |>
    distinct(cancer_type, total) |>
    mutate(cancer_type = factor(cancer_type, levels = alt_high_order))

  print(
    ggplot(
      cancer_props |> mutate(cancer_type = factor(cancer_type, levels = alt_high_order)),
      aes(x = prop, y = cancer_type, fill = alt_status)
    ) +
      geom_col() +
      geom_text(
        data      = n_labels,
        aes(x = 1.01, y = cancer_type, label = paste0("n=", total)),
        hjust = 0, size = 2.4, inherit.aes = FALSE
      ) +
      scale_x_continuous(labels = scales::percent_format(),
                         limits = c(0, 1.2), expand = c(0, 0)) +
      scale_fill_manual(values = alt_colors) +
      theme_bw() +
      labs(
        x     = "Proportion",
        y     = NULL,
        fill  = "ALT status",
        title = "Cancer type distribution by ALT status (cancer types with ≥5 samples)"
      ) +
      theme(axis.text.y = element_text(size = 8))
  )
}
```

The stacked bar chart confirms the expected enrichment of ALT-high samples in mesenchymal and neuroectodermal tumour types: bone osteosarcoma, soft tissue sarcomas (leiomyosarcoma, liposarcoma), CNS glioblastoma, and CNS pilocytic astrocytoma consistently show the highest ALT-high fractions, matching the established biology of TERT-silent tumours that depend on homologous recombination for telomere maintenance. Carcinomas (lung, colorectal, pancreatic, prostate, hepatocellular) are overwhelmingly ALT-low, reflecting near-universal TERT reactivation in epithelial lineages. This distribution has a direct analytical implication: any feature that co-varies with histology (e.g., mutation rate, copy-number burden) will appear as an ALT predictor in pooled models simply because of the cancer-type confound, underscoring the need for stratified or cancer-type-adjusted analyses.

# Feature correlation heatmap by ALT group

```{r correlation-heatmap}
#| fig-format: svg
#| out-width: '100%'
#| fig-height: 13

library(corrplot)

feat_cols      <- setdiff(names(data)[sapply(data, is.numeric)], "TERT_FPKM")
feat_cols_tert <- names(data)[sapply(data, is.numeric)]   # includes TERT_FPKM

shorten_name <- function(x) {
  x |>
    str_remove("_singleton_dist") |>
    str_replace("telomere_insertion_rate", "tel_ins_rate") |>
    str_replace("telomere_content_log2",   "tel_content") |>
    str_replace("tf_primary_rate",         "tf_primary") |>
    str_replace("tf_blood_rate",           "tf_blood")
}

# ── Row 1: full dataset, TERT excluded ─────────────────────────────────────
dat_high <- data |> filter(alt_status == "ALT-high") |>
  select(all_of(feat_cols)) |> na.omit()
dat_low  <- data |> filter(alt_status == "ALT-low")  |>
  select(all_of(feat_cols)) |> na.omit()

cor_high <- cor(dat_high, method = "spearman")
cor_low  <- cor(dat_low,  method = "spearman")
colnames(cor_high) <- rownames(cor_high) <- shorten_name(colnames(cor_high))
colnames(cor_low)  <- rownames(cor_low)  <- shorten_name(colnames(cor_low))

ord          <- corrMatOrder(cor_low, order = "hclust")
cor_low_ord  <- cor_low[ord, ord]
cor_high_ord <- cor_high[ord, ord]

# ── Row 2: complete cases, TERT included ────────────────────────────────────
dat_high_tert <- data |> filter(alt_status == "ALT-high") |>
  select(all_of(feat_cols_tert)) |> na.omit()
dat_low_tert  <- data |> filter(alt_status == "ALT-low")  |>
  select(all_of(feat_cols_tert)) |> na.omit()

cor_high_tert <- cor(dat_high_tert, method = "spearman")
cor_low_tert  <- cor(dat_low_tert,  method = "spearman")
colnames(cor_high_tert) <- rownames(cor_high_tert) <- shorten_name(colnames(cor_high_tert))
colnames(cor_low_tert)  <- rownames(cor_low_tert)  <- shorten_name(colnames(cor_low_tert))

tert_order        <- c(rownames(cor_low_ord), "TERT_FPKM")
cor_low_tert_ord  <- cor_low_tert[tert_order, tert_order]
cor_high_tert_ord <- cor_high_tert[tert_order, tert_order]

par(mfrow = c(2, 2))
corrplot(cor_low_ord,
         method = "color", type = "upper",
         tl.cex = 0.72, tl.col = "black",
         mar    = c(0, 0, 2.5, 0),
         title  = paste0("ALT-low — full dataset  (n = ", nrow(dat_low), ")"),
         cl.cex = 0.65)
corrplot(cor_high_ord,
         method = "color", type = "upper",
         tl.cex = 0.72, tl.col = "black",
         mar    = c(0, 0, 2.5, 0),
         title  = paste0("ALT-high — full dataset  (n = ", nrow(dat_high), ")"),
         cl.cex = 0.65)
corrplot(cor_low_tert_ord,
         method = "color", type = "upper",
         tl.cex = 0.72, tl.col = "black",
         mar    = c(0, 0, 2.5, 0),
         title  = paste0("ALT-low — TERT complete cases  (n = ", nrow(dat_low_tert), ")"),
         cl.cex = 0.65)
corrplot(cor_high_tert_ord,
         method = "color", type = "upper",
         tl.cex = 0.72, tl.col = "black",
         mar    = c(0, 0, 2.5, 0),
         title  = paste0("ALT-high — TERT complete cases  (n = ", nrow(dat_high_tert), ")"),
         cl.cex = 0.65)
par(mfrow = c(1, 1))
```

The 2×2 layout allows two comparisons at once: left–right compares ALT groups within the same sample set; top–bottom compares the full dataset against the TERT-complete-case subset. In the full-dataset row (top), the singleton distances show stronger mutual correlations in ALT-high than in ALT-low, consistent with a shared recombination origin driving all nine distances jointly. The relationship between `tel_content` and `tel_ins_rate` also shifts between groups. In the TERT-complete-cases row (bottom), the reduced sample sizes — particularly the small ALT-high complete-case set — should be borne in mind when interpreting any correlation involving `TERT_FPKM`: the subset is strongly non-random (ALT-high TERT missingness is \~69%), so correlations in that panel reflect a selected, predominantly TERT-expressing subgroup. Comparing the top and bottom rows for the non-TERT features gives a crude sensitivity check: if the correlation structure is stable across rows, the complete-case restriction has not introduced substantial selection bias for those features.
