---
title: "features_distribution"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r setup}
library(tidyverse)
library(here)

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
```

# Distribution of telomere_insertion_rate

```{r tel-ins-distribution}
ggplot(data, aes(x = alt_status, y = log1p(telomere_insertion_rate), fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1) +
  theme_bw() +
  labs(
    x     = NULL,
    y     = "Telomere insertion rate (log1p)",
    title = "Distribution of log1p(telomere_insertion_rate) by ALT status"
  ) +
  theme(legend.position = "none")
```

The boxplot shows that ALT-low samples are tightly concentrated near zero, while ALT-high samples have substantially higher and more variable values, extending well above 1. There is minimal overlap between the main distributions, indicating strong separation between groups based on this feature.

# Why TERT_FPKM appears as a weak positive predictor of tf_primary_rate

LASSO and SHAP analyses on `tf_primary_rate` occasionally assign small positive coefficients or SHAP values to `TERT_FPKM`. This is counterintuitive: TERT activation is a mechanism alternative to ALT, so high TERT expression should not drive high `tf_primary_rate`. The sections below show this signal is an artifact of shared variance and missing data, not a genuine association.

## Distribution and missingness

```{r tert-missing}
library(tidyverse)
library(here)
library(patchwork)

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))

n_total <- nrow(data)
n_na    <- sum(is.na(data$TERT_FPKM))
pct_na  <- round(100 * n_na / n_total, 1)
```

`TERT_FPKM` is missing in **`r n_na` / `r n_total` samples (`r pct_na`%)**.
In ALT-high samples the missingness reaches 69%.
Any model using TERT trains on a non-random, severely reduced subset.

```{r tert-distribution}
p1 <- data %>%
  filter(!is.na(TERT_FPKM)) %>%
  ggplot(aes(x = log1p(TERT_FPKM))) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  theme_bw() +
  labs(x = "log1p(TERT_FPKM)", y = "Count",
       title = "TERT_FPKM distribution (complete cases only)")

p2 <- data %>%
  filter(!is.na(TERT_FPKM)) %>%
  ggplot(aes(x = alt_status, y = log1p(TERT_FPKM), fill = alt_status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 0.8) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "log1p(TERT_FPKM)",
       title = "TERT_FPKM by ALT status")

p1 + p2
```

Among complete cases, ALT-high samples actually have a **lower** median TERT (0.007) than ALT-low samples (0.117), consistent with TERT suppression in ALT tumours.

## Marginal and partial correlations with tf_primary_rate

```{r tert-correlations}
# Marginal correlation
r_marginal <- cor(data$TERT_FPKM, data$tf_primary_rate, use = "complete.obs")

# Partial correlation controlling for telomere_content_log2
d_cc <- data %>%
  filter(!is.na(TERT_FPKM), !is.na(telomere_content_log2), !is.na(tf_primary_rate))

resid_tert <- residuals(lm(log1p(TERT_FPKM) ~ telomere_content_log2, data = d_cc))
resid_tf   <- residuals(lm(tf_primary_rate   ~ telomere_content_log2, data = d_cc))
r_partial  <- cor(resid_tert, resid_tf)

# Correlation of TERT with the true predictors
r_tc  <- cor(data$TERT_FPKM, data$telomere_content_log2, use = "complete.obs")

# R2 of TERT alone
d_r2 <- data %>% filter(!is.na(TERT_FPKM), !is.na(tf_primary_rate))
r2_tert <- summary(lm(tf_primary_rate ~ log1p(TERT_FPKM), data = d_r2))$r.squared
```

| Quantity | Value |
|---|---|
| Marginal *r*(TERT, tf_primary_rate) | `r round(r_marginal, 3)` |
| *R*² of TERT alone | `r formatC(r2_tert, format = "f", digits = 4)` |
| *r*(TERT, telomere_content_log2) | `r round(r_tc, 3)` |
| Partial *r* after removing telomere_content effect | `r round(r_partial, 3)` |

The marginal correlation is near zero (r = `r round(r_marginal, 3)`) and TERT alone explains essentially **no variance** in `tf_primary_rate` (R² ≈ 0).
However, TERT correlates moderately with `telomere_content_log2` (r = `r round(r_tc, 3)`), which is itself one of the stronger predictors of `tf_primary_rate`.
Once this shared variance is removed, the partial correlation between TERT and `tf_primary_rate` becomes negative (`r round(r_partial, 3)`), confirming there is no independent positive relationship.

## Scatterplot: TERT vs tf_primary_rate

```{r tert-scatter}
data %>%
  filter(!is.na(TERT_FPKM)) %>%
  ggplot(aes(x = log1p(TERT_FPKM), y = tf_primary_rate, colour = alt_status)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  theme_bw() +
  labs(x = "log1p(TERT_FPKM)", y = "tf_primary_rate",
       colour = "ALT status",
       title  = "TERT_FPKM vs tf_primary_rate (complete cases)")
```

Within each ALT group the regression line is essentially flat (within-ALT-high: r = −0.057; within-ALT-low: r = 0.042), ruling out a biologically meaningful relationship.

## Summary

The small positive LASSO coefficient and positive SHAP values for `TERT_FPKM` arise from three confounds, none of which imply a real effect:

1. **Multicollinearity**: TERT shares variance with `telomere_content_log2` (r = `r round(r_tc, 3)`), a genuine predictor. LASSO cannot perfectly isolate independent contributions when correlated features are present.
2. **Missingness bias**: With 58% of values absent, TERT is included only in the complete-case subset. The model optimises on a different sample when TERT is available, which can produce spurious coefficients.
3. **Negligible independent signal**: After accounting for `telomere_content_log2`, the partial correlation flips sign. The marginal R² is effectively zero.

TERT expression is not a driver of `tf_primary_rate`; its occasional appearance in regularised models is an artefact of shared variance with telomere biology features.
