---
title: "Mesenchymal vs Non-Mesenchymal group comparison"
author: "Andrea Polazzi"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 9
    fig-height: 7
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

# Setup

```{r setup}
library(tidyverse)
library(here)
library(patchwork)
library(effsize)   # Cliff's delta
library(knitr)     # kable tables

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
clean_data <- data %>%
  select(-TERT_FPKM) %>%    # remove this line to include TERT but lose sample size (n=1942 vs n=816)
  drop_na() %>%
  filter(tf_primary_rate < 10)

cancer_group_vector <- clean_data$cancer_group

clean_data <- clean_data %>%
  mutate(
    tf_primary_rate         = log1p(tf_primary_rate),
    tf_blood_rate           = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    alt_status = factor(alt_status, levels = c('ALT-low', 'ALT-high'))
  ) %>%
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)
```

```{r build-frame}
# ---- Build the working frame -----------------------------
df_group <- clean_data %>%
  mutate(cancer_group = factor(cancer_group_vector))

# Extract group names ONCE from the data, don't hardcode
grp_levels <- levels(df_group$cancer_group)
stopifnot(length(grp_levels) == 2)  # this analysis assumes 2 groups

# Identify which is which by name pattern (robust to '-' vs '_')
grp_mes <- grp_levels[grepl('(?i)^mesenchymal', grp_levels)]
grp_non <- setdiff(grp_levels, grp_mes)
stopifnot(length(grp_mes) == 1, length(grp_non) == 1)

# Reorder so non-mesenchymal is the reference
df_group$cancer_group <- factor(df_group$cancer_group, levels = c(grp_non, grp_mes))

grp_colors      <- setNames(c('grey60', 'firebrick'),  c(grp_non, grp_mes))
grp_colors_dark <- setNames(c('grey40', 'firebrick4'), c(grp_non, grp_mes))

numeric_vars    <- df_group %>% select(-cancer_group, -alt_status) %>% names()
singleton_vars  <- grep('_singleton_dist$', numeric_vars, value = TRUE)
continuous_vars <- setdiff(numeric_vars, singleton_vars)

strata_levels <- c('overall', 'ALT-low', 'ALT-high')
```

```{r tert-data, include=FALSE}
df_tert <- data %>%
  drop_na() %>%
  filter(tf_primary_rate < 10) %>%
  mutate(
    tf_primary_rate         = log1p(tf_primary_rate),
    tf_blood_rate           = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    TERT_FPKM               = log1p(TERT_FPKM),
    alt_status   = factor(alt_status, levels = c('ALT-low', 'ALT-high')),
    cancer_group = factor(cancer_group, levels = c(grp_non, grp_mes))
  ) %>%
  select(-donor_id, -cancer_type, -Specimen.Type.Summary)

tert_numeric_vars <- c(numeric_vars, 'TERT_FPKM')
```

```{r grp-tests}
# Master results table — every numeric feature × {overall, ALT-low, ALT-high}.
# This is the single source of truth: plots show only the significance star,
# the exact numbers (Cliff's delta, p, BH-adjusted p) live in the tables below
# so they can be cited verbatim in the text.
#
#  * Test:      Wilcoxon rank-sum (Mann-Whitney), mesenchymal vs non-mesenchymal.
#  * Effect:    Cliff's delta, computed mes vs non — its SIGN is the direction
#               (delta > 0 => higher in mesenchymal) and its magnitude the size.
#  * Multiple testing: Benjamini-Hochberg, applied WITHIN each view (stratum)
#               across all 13 features tested in that view.
one_test <- function(df, v, stratum) {
  x <- df[[v]][df$cancer_group == grp_mes]
  y <- df[[v]][df$cancer_group == grp_non]
  delta <- tryCatch(suppressWarnings(effsize::cliff.delta(x, y)$estimate),
                    error = function(e) NA_real_)
  p <- tryCatch(suppressWarnings(wilcox.test(x, y)$p.value),
                error = function(e) NA_real_)
  tibble(var = v, stratum = stratum,
         n_mes = length(x), n_non = length(y),
         median_mes = median(x, na.rm = TRUE), median_non = median(y, na.rm = TRUE),
         cliff_delta = delta, p = p)
}

grp_tests <- map_dfr(numeric_vars, function(v) {
  bind_rows(
    one_test(df_group, v, 'overall'),
    one_test(filter(df_group, alt_status == 'ALT-low'),  v, 'ALT-low'),
    one_test(filter(df_group, alt_status == 'ALT-high'), v, 'ALT-high')
  )
}) %>%
  group_by(stratum) %>%
  mutate(p.adj = p.adjust(p, method = 'BH')) %>%
  ungroup() %>%
  mutate(
    stratum = factor(stratum, levels = strata_levels),
    signif  = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                  labels = c('***', '**', '*', 'ns'))
  )

# Thesis-ready table: rounded effect size, p, adjusted p, star.
fmt_grp <- function(tbl, drop_suffix = FALSE) {
  if (drop_suffix) tbl <- tbl %>% mutate(var = sub('_singleton_dist$', '', var))
  tbl %>%
    arrange(var, stratum) %>%
    transmute(feature = var, stratum,
              n_mes, n_non,
              median_mes  = round(median_mes, 3),
              median_non  = round(median_non, 3),
              cliff_delta = round(cliff_delta, 3),
              p     = signif(p, 3),
              p.adj = signif(p.adj, 3),
              signif) %>%
    kable()
}
```

Groups detected in the data:

```{r group-table}
cat('Reference:  ', grp_non, "\n")
cat('Comparison: ', grp_mes, "\n\n")
print(table(df_group$cancer_group))
```

# 1. Overview — Cliff's delta across all features

A single ranking of effect sizes (overall, not yet stratified). Positive
= higher in mesenchymal. Dotted lines mark the negligible/small
thresholds — note how many features are statistically significant yet
land in the negligible band, a consequence of the large cohort (n =
1942).

```{r overview-ranking, fig.height=6}
grp_tests %>%
  filter(stratum == 'overall', !is.na(cliff_delta)) %>%
  mutate(mag = cut(abs(cliff_delta), c(-Inf, 0.147, 0.33, 0.474, Inf),
                   labels = c('negligible', 'small', 'medium', 'large')),
         feature = fct_reorder(var, abs(cliff_delta))) %>%
  ggplot(aes(x = cliff_delta, y = feature, fill = mag)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey40') +
  geom_vline(xintercept = c(-0.147, 0.147), linetype = 'dotted', color = 'grey60') +
  geom_vline(xintercept = c(-0.33, 0.33),   linetype = 'dotted', color = 'grey60') +
  scale_fill_manual(values = c(negligible = 'grey80', small = '#fdcc8a',
                               medium = '#fc8d59', large = '#d7301f'), drop = FALSE) +
  labs(title = "Cliff's delta: Mesenchymal vs Non-Mesenchymal (overall)",
       subtitle = sprintf('Positive = higher in %s (n=%d vs %d).',
                          grp_mes, sum(df_group$cancer_group == grp_mes),
                          sum(df_group$cancer_group == grp_non)),
       x = "Cliff's delta', y = NULL, fill = 'Effect size") +
  theme_bw()
```

## 1b. TERT-available subset (n = 816) — including TERT_FPKM

Same ranking restricted to the n = 816 samples with TERT data, adding
`TERT_FPKM` as a feature. BH correction is re-applied across all 14
features in this subset.

```{r overview-ranking-tert, fig.height=6}
grp_tests_tert <- map_dfr(tert_numeric_vars, function(v) {
  bind_rows(
    one_test(df_tert, v, 'overall'),
    one_test(filter(df_tert, alt_status == 'ALT-low'),  v, 'ALT-low'),
    one_test(filter(df_tert, alt_status == 'ALT-high'), v, 'ALT-high')
  )
}) %>%
  group_by(stratum) %>%
  mutate(p.adj = p.adjust(p, method = 'BH')) %>%
  ungroup() %>%
  mutate(
    stratum = factor(stratum, levels = strata_levels),
    signif  = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                  labels = c('***', '**', '*', 'ns'))
  )

grp_tests_tert %>%
  filter(stratum == 'overall', !is.na(cliff_delta)) %>%
  mutate(mag = cut(abs(cliff_delta), c(-Inf, 0.147, 0.33, 0.474, Inf),
                   labels = c('negligible', 'small', 'medium', 'large')),
         feature = fct_reorder(var, abs(cliff_delta))) %>%
  ggplot(aes(x = cliff_delta, y = feature, fill = mag)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey40') +
  geom_vline(xintercept = c(-0.147, 0.147), linetype = 'dotted', color = 'grey60') +
  geom_vline(xintercept = c(-0.33, 0.33),   linetype = 'dotted', color = 'grey60') +
  scale_fill_manual(values = c(negligible = 'grey80', small = '#fdcc8a',
                               medium = '#fc8d59', large = '#d7301f'), drop = FALSE) +
  labs(title = "Cliff's delta: Mesenchymal vs Non-Mesenchymal — TERT subset (n = 816)",
       subtitle = sprintf('Positive = higher in %s (n=%d vs %d). Includes TERT_FPKM.',
                          grp_mes, sum(df_tert$cancer_group == grp_mes),
                          sum(df_tert$cancer_group == grp_non)),
       x = "Cliff's delta', y = NULL, fill = 'Effect size") +
  theme_bw()
```

# 2. ALT-status composition by group

```{r alt-composition}
alt_table <- df_group %>%
  count(cancer_group, alt_status) %>%
  group_by(cancer_group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

alt_table %>%
  ggplot(aes(x = cancer_group, y = prop, fill = alt_status)) +
  geom_col(position = 'stack', alpha = 0.85) +
  geom_text(aes(label = paste0(n, "\n(', scales::percent(prop, 1), ')")),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c('ALT-low' = 'grey70', 'ALT-high' = 'firebrick')) +
  labs(title = 'ALT status composition by cancer group',
       x = NULL, y = 'Proportion', fill = NULL) +
  theme_bw()
```

```{r alt-prevalence}
set.seed(21)
alt_high_mes <- as.numeric(df_group$alt_status[df_group$cancer_group == grp_mes] == 'ALT-high')
alt_high_non <- as.numeric(df_group$alt_status[df_group$cancer_group == grp_non] == 'ALT-high')

boot_diff <- replicate(2000, {
  s_mes <- sample(alt_high_mes, length(alt_high_mes), replace = TRUE)
  s_non <- sample(alt_high_non, length(alt_high_non), replace = TRUE)
  mean(s_mes) - mean(s_non)
})
boot_diff <- boot_diff[is.finite(boot_diff)]

p_mes_obs <- mean(alt_high_mes); p_non_obs <- mean(alt_high_non)

cat("ALT-high prevalence by cancer group\n")
cat(sprintf("  %s: %.1f%% (%d/%d)\n", grp_mes, 100 * p_mes_obs, sum(alt_high_mes), length(alt_high_mes)))
cat(sprintf("  %s: %.1f%% (%d/%d)\n", grp_non, 100 * p_non_obs, sum(alt_high_non), length(alt_high_non)))
cat(sprintf("\nDifference (mes - non):\n"))
cat(sprintf("  observed: %+.3f\n", p_mes_obs - p_non_obs))
cat(sprintf("  95%% stratified bootstrap CI: [%+.3f, %+.3f]\n",
            quantile(boot_diff, 0.025), quantile(boot_diff, 0.975)))
```

# 3. Continuous variables by group, stratified by ALT status

The tTF/bTF pattern applied to every continuous variable: an **overall**
view (top, groups compared directly) and an **ALT-stratified** view
(bottom, faceted by ALT status) to test whether the group effect
survives holding ALT fixed. Plots carry only the significance star
(`***` \< 0.001, `**` \< 0.01, `*` \< 0.05, BH-adjusted); exact numbers
are in the table.

```{r continuous-table}
fmt_grp(filter(grp_tests, var %in% continuous_vars))
```

```{r alt-strat-fn}
# Stars only (BH q), placed at the top of each panel — table holds the numbers.
alt_strat_plot <- function(v, ylab = v) {
  dd    <- df_group %>% mutate(value = .data[[v]])
  ann_o <- grp_tests %>% filter(var == v, stratum == 'overall')
  ann_a <- grp_tests %>% filter(var == v, stratum %in% c('ALT-low', 'ALT-high')) %>%
    mutate(alt_status = factor(as.character(stratum), levels = levels(dd$alt_status)))

  p_overall <- ggplot(dd, aes(x = cancer_group, y = value, fill = cancer_group)) +
    geom_violin(alpha = 0.4, color = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.15, size = 0.9) +
    geom_text(data = ann_o, aes(x = 1.5, y = Inf, label = signif),
              inherit.aes = FALSE, vjust = 1.5, size = 4.5) +
    scale_fill_manual(values = grp_colors) +
    labs(title = sprintf('%s by cancer group', v), x = NULL, y = ylab) +
    theme_bw() + theme(legend.position = 'none')

  p_alt <- ggplot(dd, aes(x = cancer_group, y = value, fill = cancer_group)) +
    geom_violin(alpha = 0.4, color = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.15, size = 0.8) +
    geom_text(data = ann_a, aes(x = 1.5, y = Inf, label = signif),
              inherit.aes = FALSE, vjust = 1.5, size = 4.5) +
    facet_wrap(~ alt_status) +
    scale_fill_manual(values = grp_colors) +
    labs(title = sprintf('%s by group, stratified by ALT status', v), x = NULL, y = ylab) +
    theme_bw() + theme(legend.position = 'none', strip.text = element_text(face = 'bold'))

  p_overall / p_alt
}
```

```{r continuous-plots, results='asis', fig.height=7}
for (v in continuous_vars) {
  cat("\n\n## ', v, '\n\n")
  print(alt_strat_plot(v))
}
```

The confounder is ALT composition: - Telomere insertion is essentially
an ALT phenomenon: in ALT-low \~75–94% of samples are exactly 0; in
ALT-high the median jumps to \~0.07–0.10.\
\
Mesenchymal is hugely ALT-enriched: 37.5% ALT-high (81/216) vs only 4.9%
(84/1726) in non-mesenchymal.

So the overall "mesenchymal has higher insertion" is almost entirely
"mesenchymal contains far more ALT-high samples, and ALT-high samples
insert."

Once you hold ALT fixed, the marginal effect disappears: - In ALT-low,
mesenchymal actually inserts less (94% zeros vs 75%) → significant
negative δ.

In ALT-high, no detectable group difference (and only 81 vs 84 samples →
low power, which is why p=0.10 despite δ=0.147).

Interpretation: the group effect on telomere_insertion does not survive
ALT stratification — it's a byproduct of ALT prevalence differing
between groups.

# 4. Singleton motifs — all 9 together

The 9 `*_singleton_dist` motifs share a common scale, so they are shown
jointly in two stacked 3×3 grids — the tTF/bTF top/bottom layout: an
overall grid (group comparison) and an ALT-faceted grid below. Stars
only on the plots; numbers in the tables.

```{r singleton-long}
sing_long <- df_group %>%
  select(cancer_group, alt_status, all_of(singleton_vars)) %>%
  pivot_longer(all_of(singleton_vars), names_to = 'motif', values_to = 'value') %>%
  mutate(motif = factor(sub('_singleton_dist$', '', motif)))
```

## Overall — 3×3 grid by cancer group

```{r singleton-overall-table}
fmt_grp(filter(grp_tests, var %in% singleton_vars, stratum == 'overall'), drop_suffix = TRUE)
```

```{r singleton-overall, fig.height=8, fig.width=10}
sing_overall_annot <- grp_tests %>%
  filter(var %in% singleton_vars, stratum == 'overall') %>%
  mutate(motif = factor(sub('_singleton_dist$', '', var)))

ggplot(sing_long, aes(x = cancer_group, y = value, fill = cancer_group)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 0.5) +
  geom_text(data = sing_overall_annot, aes(x = 1.5, y = Inf, label = signif),
            inherit.aes = FALSE, vjust = 1.5, size = 4) +
  facet_wrap(~ motif, ncol = 3, scales = 'free_y') +
  scale_fill_manual(values = grp_colors) +
  labs(title = 'Singleton motifs by cancer group (overall)',
       subtitle = sprintf('%s in red.', grp_mes),
       x = NULL, y = 'singleton_dist') +
  theme_bw() + theme(legend.position = 'none', strip.text = element_text(face = 'bold'))
```

## Stratified by ALT status — 3×3 grid

```{r singleton-alt-table}
fmt_grp(filter(grp_tests, var %in% singleton_vars, stratum %in% c('ALT-low', 'ALT-high')),
        drop_suffix = TRUE)
```

```{r singleton-alt, fig.height=9, fig.width=12}
sing_alt_annot <- grp_tests %>%
  filter(var %in% singleton_vars, stratum %in% c('ALT-low', 'ALT-high')) %>%
  mutate(motif = factor(sub('_singleton_dist$', '', var)),
         alt_status = factor(as.character(stratum), levels = levels(sing_long$alt_status)))

ggplot(sing_long, aes(x = alt_status, y = value, fill = cancer_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(width = 0.8)) +
  geom_text(data = sing_alt_annot, aes(x = alt_status, y = Inf, label = signif),
            inherit.aes = FALSE, vjust = 1.5, size = 3.5) +
  facet_wrap(~ motif, ncol = 3, scales = 'free_y') +
  scale_fill_manual(values = grp_colors) +
  labs(title = 'Singleton motifs by group, stratified by ALT status',
       subtitle = sprintf('%s in red. x = ALT status; colour = cancer group. Star compares groups within each stratum.', grp_mes),
       x = NULL, y = 'singleton_dist', fill = NULL) +
  theme_bw() + theme(legend.position = 'top', strip.text = element_text(face = 'bold'))
```

# 5. TERT expression — TERT-available subset

TERT expression (FPKM) is available for only a subset of samples: after
applying the same filters as the main analysis (`tf_primary_rate < 10`,
`drop_na` on every column including `TERT_FPKM`), **n = 816** samples
remain (vs n = 1942 in sections 1–4). All tests and plots in this
section therefore use this restricted subset. Everything else — test
choice, effect-size metric, BH correction within each stratum — is
identical to the rest of the notebook.

```{r tert-subset}
cat(sprintf(
  "TERT subset: n = %d  (%s: %d, %s: %d)\n",
  nrow(df_tert),
  grp_mes, sum(df_tert$cancer_group == grp_mes),
  grp_non, sum(df_tert$cancer_group == grp_non)
))
```

```{r tert-tests}
tert_tests <- bind_rows(
  one_test(df_tert,                                    'TERT_FPKM', 'overall'),
  one_test(filter(df_tert, alt_status == 'ALT-low'),   'TERT_FPKM', 'ALT-low'),
  one_test(filter(df_tert, alt_status == 'ALT-high'),  'TERT_FPKM', 'ALT-high')
) %>%
  mutate(
    p.adj   = p.adjust(p, method = 'BH'),
    stratum = factor(stratum, levels = strata_levels),
    signif  = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                  labels = c('***', '**', '*', 'ns'))
  )
```

## Table

```{r tert-table}
tert_tests %>%
  transmute(
    stratum,
    n_mes, n_non,
    median_mes  = round(median_mes,  3),
    median_non  = round(median_non,  3),
    cliff_delta = round(cliff_delta, 3),
    p     = signif(p, 3),
    p.adj = signif(p.adj, 3),
    signif
  ) %>%
  kable(caption = "TERT_FPKM (log1p): Mesenchymal vs Non-Mesenchymal — Wilcoxon + Cliff's delta")
```

## Plots

```{r tert-overall-plot}
ann_o_tert <- tert_tests %>% filter(stratum == 'overall')

ggplot(df_tert, aes(x = cancer_group, y = TERT_FPKM, fill = cancer_group)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.15, size = 0.9) +
  geom_text(
    data = ann_o_tert,
    aes(x = 1.5, y = Inf, label = signif),
    inherit.aes = FALSE, vjust = 1.5, size = 4.5
  ) +
  scale_fill_manual(values = grp_colors) +
  labs(
    title   = 'TERT expression (log1p FPKM) by cancer group',
    subtitle = sprintf('TERT subset: n = %d  (%s n=%d, %s n=%d)',
                       nrow(df_tert),
                       grp_mes, sum(df_tert$cancer_group == grp_mes),
                       grp_non, sum(df_tert$cancer_group == grp_non)),
    x = NULL, y = 'TERT_FPKM (log1p)'
  ) +
  theme_bw() + theme(legend.position = 'none')
```

```{r tert-alt-plot}
ann_a_tert <- tert_tests %>%
  filter(stratum %in% c('ALT-low', 'ALT-high')) %>%
  mutate(alt_status = factor(as.character(stratum), levels = levels(df_tert$alt_status)))

ggplot(df_tert, aes(x = cancer_group, y = TERT_FPKM, fill = cancer_group)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.15, size = 0.8) +
  geom_text(
    data = ann_a_tert,
    aes(x = 1.5, y = Inf, label = signif),
    inherit.aes = FALSE, vjust = 1.5, size = 4.5
  ) +
  facet_wrap(~ alt_status) +
  scale_fill_manual(values = grp_colors) +
  labs(
    title    = 'TERT expression by group, stratified by ALT status',
    subtitle = sprintf('TERT subset only (n = %d). Star = BH-adjusted Wilcoxon within stratum.', nrow(df_tert)),
    x = NULL, y = 'TERT_FPKM (log1p)'
  ) +
  theme_bw() +
  theme(legend.position = 'none', strip.text = element_text(face = 'bold'))
```

# 6. Correlation structure differences

```{r correlation}
cor_mes <- df_group %>% filter(cancer_group == grp_mes) %>%
  select(all_of(numeric_vars)) %>% cor(use = 'pairwise.complete.obs')
cor_non <- df_group %>% filter(cancer_group == grp_non) %>%
  select(all_of(numeric_vars)) %>% cor(use = 'pairwise.complete.obs')
cor_diff <- cor_mes - cor_non

cor_diff_long <- cor_diff %>%
  as.data.frame() %>% rownames_to_column('feat1') %>%
  pivot_longer(-feat1, names_to = 'feat2', values_to = 'diff') %>%
  filter(feat1 < feat2) %>% arrange(desc(abs(diff)))

cat("Top pairwise correlation differences (mes - non):\n")
print(head(cor_diff_long, 15))

cor_diff %>%
  as.data.frame() %>% rownames_to_column('feat1') %>%
  pivot_longer(-feat1, names_to = 'feat2', values_to = 'diff') %>%
  ggplot(aes(x = feat1, y = feat2, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = 'steelblue', mid = 'white', high = 'firebrick',
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = 'Correlation structure difference',
       subtitle = sprintf('cor(%s) - cor(%s). Red = stronger in mesenchymal.', grp_mes, grp_non),
       x = NULL, y = NULL, fill = 'Δ cor') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))
```

## 6b. TERT-available subset (n = 816) — including TERT_FPKM

```{r correlation-tert}
cor_tert_mes <- df_tert %>% filter(cancer_group == grp_mes) %>%
  select(all_of(tert_numeric_vars)) %>% cor(use = 'pairwise.complete.obs')
cor_tert_non <- df_tert %>% filter(cancer_group == grp_non) %>%
  select(all_of(tert_numeric_vars)) %>% cor(use = 'pairwise.complete.obs')
cor_tert_diff <- cor_tert_mes - cor_tert_non

cor_tert_diff_long <- cor_tert_diff %>%
  as.data.frame() %>% rownames_to_column('feat1') %>%
  pivot_longer(-feat1, names_to = 'feat2', values_to = 'diff') %>%
  filter(feat1 < feat2) %>% arrange(desc(abs(diff)))

cat("Top pairwise correlation differences — TERT subset (mes - non):\n")
print(head(cor_tert_diff_long, 15))

cor_tert_diff %>%
  as.data.frame() %>% rownames_to_column('feat1') %>%
  pivot_longer(-feat1, names_to = 'feat2', values_to = 'diff') %>%
  ggplot(aes(x = feat1, y = feat2, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = 'steelblue', mid = 'white', high = 'firebrick',
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = 'Correlation structure difference — TERT subset (n = 816)',
       subtitle = sprintf('cor(%s) - cor(%s). Includes TERT_FPKM. Red = stronger in mesenchymal.', grp_mes, grp_non),
       x = NULL, y = NULL, fill = 'Δ cor') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))
```

------------------------------------------------------------------------

# 7. Metastatic dataset — Mesenchymal vs Non-Mesenchymal

This section repeats the mesenchymal group comparison on the
**metastatic cohort** (`metastatic_red_edit_mesenchymal`). Key
differences from the primary analysis:

-   Grouping variable is `tumor_origin` (Mesenchymal / Mix /
    Non_mesenchymal). `Mix` patients are excluded to keep the binary
    contrast.
-   The dataset has multiple rows per patient (one per driver gene). It
    is de-duplicated to **one row per patient** before analysis; TF
    rates and singleton distances are patient-level constants.
-   `alt_status`, `TERT_FPKM`, and `telomere_insertion_rate` are absent
    — no ALT stratification and no TERT section.
-   Column mapping: `tumor_tf_rate` → tTF, `blood_tf_rate` → bTF.

```{r met-setup}
met_raw <- readxl::read_xlsx(here('data', 'processed', 'metastatic_red_edit_mesenchymal.xlsx'))

# One row per patient; exclude Mix
met_data <- met_raw %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  filter(tumor_origin %in% c('Mesenchymal', 'Non_mesenchymal')) %>%
  mutate(
    tumor_tf_rate  = log1p(tumor_tf_rate),
    blood_tf_rate  = log1p(blood_tf_rate),
    cancer_group   = factor(tumor_origin, levels = c('Non_mesenchymal', 'Mesenchymal'))
  )

met_mes <- 'Mesenchymal'
met_non <- 'Non_mesenchymal'

met_colors      <- setNames(c('grey60', 'firebrick'),  c(met_non, met_mes))

met_singleton_vars  <- grep('_singleton_dist$', names(met_data), value = TRUE)
met_continuous_vars <- c('tumor_tf_rate', 'blood_tf_rate')
met_numeric_vars    <- c(met_continuous_vars, met_singleton_vars)

cat(sprintf("Metastatic cohort (binary): n = %d  (%s: %d, %s: %d)\n",
            nrow(met_data),
            met_mes, sum(met_data$cancer_group == met_mes),
            met_non, sum(met_data$cancer_group == met_non)))
```

```{r met-tests}
one_test_met <- function(df, v) {
  x <- df[[v]][df$cancer_group == met_mes]
  y <- df[[v]][df$cancer_group == met_non]
  delta <- tryCatch(suppressWarnings(effsize::cliff.delta(x, y)$estimate),
                    error = function(e) NA_real_)
  p <- tryCatch(suppressWarnings(wilcox.test(x, y)$p.value),
                error = function(e) NA_real_)
  tibble(var = v,
         n_mes = length(x), n_non = length(y),
         median_mes = median(x, na.rm = TRUE), median_non = median(y, na.rm = TRUE),
         cliff_delta = delta, p = p)
}

met_tests <- map_dfr(met_numeric_vars, one_test_met, df = met_data) %>%
  mutate(
    p.adj  = p.adjust(p, method = 'BH'),
    signif = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                 labels = c('***', '**', '*', 'ns'))
  )
```

## 7.1 Overview — Cliff's delta across all features

```{r met-overview, fig.height=5}
met_tests %>%
  filter(!is.na(cliff_delta)) %>%
  mutate(mag = cut(abs(cliff_delta), c(-Inf, 0.147, 0.33, 0.474, Inf),
                   labels = c('negligible', 'small', 'medium', 'large')),
         feature = fct_reorder(var, abs(cliff_delta))) %>%
  ggplot(aes(x = cliff_delta, y = feature, fill = mag)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey40') +
  geom_vline(xintercept = c(-0.147, 0.147), linetype = 'dotted', color = 'grey60') +
  geom_vline(xintercept = c(-0.33, 0.33),   linetype = 'dotted', color = 'grey60') +
  scale_fill_manual(values = c(negligible = 'grey80', small = '#fdcc8a',
                               medium = '#fc8d59', large = '#d7301f'), drop = FALSE) +
  labs(title = "Cliff's delta: Mesenchymal vs Non-Mesenchymal — metastatic cohort",
       subtitle = sprintf('Positive = higher in Mesenchymal (n=%d vs %d). BH-adjusted p.',
                          sum(met_data$cancer_group == met_mes),
                          sum(met_data$cancer_group == met_non)),
       x = "Cliff's delta', y = NULL, fill = 'Effect size") +
  theme_bw()
```

## 7.2 tTF and bTF by group

```{r met-continuous-table}
met_tests %>%
  filter(var %in% met_continuous_vars) %>%
  transmute(feature = var, n_mes, n_non,
            median_mes  = round(median_mes,  3),
            median_non  = round(median_non,  3),
            cliff_delta = round(cliff_delta, 3),
            p     = signif(p, 3),
            p.adj = signif(p.adj, 3),
            signif) %>%
  kable(caption = 'tTF and bTF (log1p): Mesenchymal vs Non-Mesenchymal — metastatic cohort')
```

```{r met-continuous-plots, fig.height=4}
for (v in met_continuous_vars) {
  ann <- met_tests %>% filter(var == v)
  p <- ggplot(met_data, aes(x = cancer_group, y = .data[[v]], fill = cancer_group)) +
    geom_violin(alpha = 0.4, color = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.15, size = 0.9) +
    geom_text(data = ann, aes(x = 1.5, y = Inf, label = signif),
              inherit.aes = FALSE, vjust = 1.5, size = 4.5) +
    scale_fill_manual(values = met_colors) +
    labs(title = sprintf('%s by cancer group — metastatic', v), x = NULL, y = v) +
    theme_bw() + theme(legend.position = 'none')
  print(p)
}
```

## 7.3 Tumor origin composition

```{r met-origin-table}
met_raw %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  count(tumor_origin, primary_tumor_type) %>%
  arrange(tumor_origin, desc(n)) %>%
  kable(caption = 'Tumor types by tumor_origin (all patients, including Mix)')
```

## 7.4 Singleton motifs by group

```{r met-singleton-table}
met_tests %>%
  filter(var %in% met_singleton_vars) %>%
  mutate(var = sub('_singleton_dist$', '', var)) %>%
  transmute(feature = var, n_mes, n_non,
            median_mes  = round(median_mes,  3),
            median_non  = round(median_non,  3),
            cliff_delta = round(cliff_delta, 3),
            p     = signif(p, 3),
            p.adj = signif(p.adj, 3),
            signif) %>%
  kable(caption = 'Singleton motifs: Mesenchymal vs Non-Mesenchymal — metastatic cohort')
```

```{r met-singleton-plot, fig.height=8, fig.width=10}
met_sing_long <- met_data %>%
  select(cancer_group, all_of(met_singleton_vars)) %>%
  pivot_longer(all_of(met_singleton_vars), names_to = 'motif', values_to = 'value') %>%
  mutate(motif = factor(sub('_singleton_dist$', '', motif)))

met_sing_annot <- met_tests %>%
  filter(var %in% met_singleton_vars) %>%
  mutate(motif = factor(sub('_singleton_dist$', '', var)))

ggplot(met_sing_long, aes(x = cancer_group, y = value, fill = cancer_group)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 0.5) +
  geom_text(data = met_sing_annot, aes(x = 1.5, y = Inf, label = signif),
            inherit.aes = FALSE, vjust = 1.5, size = 4) +
  facet_wrap(~ motif, ncol = 3, scales = 'free_y') +
  scale_fill_manual(values = met_colors) +
  labs(title = 'Singleton motifs by cancer group — metastatic cohort',
       subtitle = 'Mesenchymal in red. BH-adjusted Wilcoxon.',
       x = NULL, y = 'singleton_dist') +
  theme_bw() + theme(legend.position = 'none', strip.text = element_text(face = 'bold'))
```

## 7.5 Correlation structure differences — metastatic

```{r met-correlation}
cor_met_mes <- met_data %>% filter(cancer_group == met_mes) %>%
  select(all_of(met_numeric_vars)) %>% cor(use = 'pairwise.complete.obs')
cor_met_non <- met_data %>% filter(cancer_group == met_non) %>%
  select(all_of(met_numeric_vars)) %>% cor(use = 'pairwise.complete.obs')
cor_met_diff <- cor_met_mes - cor_met_non

cor_met_diff %>%
  as.data.frame() %>% rownames_to_column('feat1') %>%
  pivot_longer(-feat1, names_to = 'feat2', values_to = 'diff') %>%
  ggplot(aes(x = feat1, y = feat2, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = 'steelblue', mid = 'white', high = 'firebrick',
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = 'Correlation structure difference — metastatic cohort',
       subtitle = 'cor(Mesenchymal) - cor(Non_mesenchymal). Red = stronger in mesenchymal.',
       x = NULL, y = NULL, fill = 'Δ cor') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))
```
