---
title: "Mesenchymal vs Non-Mesenchymal group comparison"
author: "Andrea Polazzi"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 9
    fig-height: 7
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

# Setup

```{r setup}
library(tidyverse)
library(here)
library(patchwork)
library(effsize)   # Cliff's delta
library(knitr)     # kable tables

data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
clean_data <- data %>%
  select(-TERT_FPKM) %>%    # remove this line to include TERT but lose sample size (n=1942 vs n=816)
  drop_na() %>%
  filter(tf_primary_rate < 10)

cancer_group_vector <- clean_data$cancer_group

clean_data <- clean_data %>%
  mutate(
    tf_primary_rate         = log1p(tf_primary_rate),
    tf_blood_rate           = log1p(tf_blood_rate),
    telomere_insertion_rate = log1p(telomere_insertion_rate),
    alt_status = factor(alt_status, levels = c("ALT-low", "ALT-high"))
  ) %>%
  select(-donor_id, -cancer_type, -Specimen.Type.Summary, -cancer_group)
```

```{r build-frame}
# ---- Build the working frame -----------------------------
df_group <- clean_data %>%
  mutate(cancer_group = factor(cancer_group_vector))

# Extract group names ONCE from the data, don't hardcode
grp_levels <- levels(df_group$cancer_group)
stopifnot(length(grp_levels) == 2)  # this analysis assumes 2 groups

# Identify which is which by name pattern (robust to '-' vs '_')
grp_mes <- grp_levels[grepl("(?i)^mesenchymal", grp_levels)]
grp_non <- setdiff(grp_levels, grp_mes)
stopifnot(length(grp_mes) == 1, length(grp_non) == 1)

# Reorder so non-mesenchymal is the reference
df_group$cancer_group <- factor(df_group$cancer_group, levels = c(grp_non, grp_mes))

grp_colors      <- setNames(c("grey60", "firebrick"),  c(grp_non, grp_mes))
grp_colors_dark <- setNames(c("grey40", "firebrick4"), c(grp_non, grp_mes))

numeric_vars    <- df_group %>% select(-cancer_group, -alt_status) %>% names()
singleton_vars  <- grep("_singleton_dist$", numeric_vars, value = TRUE)
continuous_vars <- setdiff(numeric_vars, singleton_vars)

strata_levels <- c("overall", "ALT-low", "ALT-high")
```

```{r grp-tests}
# Master results table — every numeric feature × {overall, ALT-low, ALT-high}.
# This is the single source of truth: plots show only the significance star,
# the exact numbers (Cliff's delta, p, BH-adjusted p) live in the tables below
# so they can be cited verbatim in the text.
#
#  * Test:      Wilcoxon rank-sum (Mann-Whitney), mesenchymal vs non-mesenchymal.
#  * Effect:    Cliff's delta, computed mes vs non — its SIGN is the direction
#               (delta > 0 => higher in mesenchymal) and its magnitude the size.
#  * Multiple testing: Benjamini-Hochberg, applied WITHIN each view (stratum)
#               across all 13 features tested in that view.
one_test <- function(df, v, stratum) {
  x <- df[[v]][df$cancer_group == grp_mes]
  y <- df[[v]][df$cancer_group == grp_non]
  delta <- tryCatch(suppressWarnings(effsize::cliff.delta(x, y)$estimate),
                    error = function(e) NA_real_)
  p <- tryCatch(suppressWarnings(wilcox.test(x, y)$p.value),
                error = function(e) NA_real_)
  tibble(var = v, stratum = stratum,
         n_mes = length(x), n_non = length(y),
         median_mes = median(x, na.rm = TRUE), median_non = median(y, na.rm = TRUE),
         cliff_delta = delta, p = p)
}

grp_tests <- map_dfr(numeric_vars, function(v) {
  bind_rows(
    one_test(df_group, v, "overall"),
    one_test(filter(df_group, alt_status == "ALT-low"),  v, "ALT-low"),
    one_test(filter(df_group, alt_status == "ALT-high"), v, "ALT-high")
  )
}) %>%
  group_by(stratum) %>%
  mutate(p.adj = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  mutate(
    stratum = factor(stratum, levels = strata_levels),
    signif  = cut(p.adj, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                  labels = c("***", "**", "*", "ns"))
  )

# Thesis-ready table: rounded effect size, p, adjusted p, star.
fmt_grp <- function(tbl, drop_suffix = FALSE) {
  if (drop_suffix) tbl <- tbl %>% mutate(var = sub("_singleton_dist$", "", var))
  tbl %>%
    arrange(var, stratum) %>%
    transmute(feature = var, stratum,
              n_mes, n_non,
              median_mes  = round(median_mes, 3),
              median_non  = round(median_non, 3),
              cliff_delta = round(cliff_delta, 3),
              p     = signif(p, 3),
              p.adj = signif(p.adj, 3),
              signif) %>%
    kable()
}
```

Groups detected in the data:

```{r group-table}
cat("Reference:  ", grp_non, "\n")
cat("Comparison: ", grp_mes, "\n\n")
print(table(df_group$cancer_group))
```

# 1. Overview — Cliff's delta across all features

A single ranking of effect sizes (overall, not yet stratified). Positive
= higher in mesenchymal. Dotted lines mark the negligible/small
thresholds — note how many features are statistically significant yet
land in the negligible band, a consequence of the large cohort (n =
1942).

```{r overview-ranking, fig.height=6}
grp_tests %>%
  filter(stratum == "overall", !is.na(cliff_delta)) %>%
  mutate(mag = cut(abs(cliff_delta), c(-Inf, 0.147, 0.33, 0.474, Inf),
                   labels = c("negligible", "small", "medium", "large")),
         feature = fct_reorder(var, abs(cliff_delta))) %>%
  ggplot(aes(x = cliff_delta, y = feature, fill = mag)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.147, 0.147), linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = c(-0.33, 0.33),   linetype = "dotted", color = "grey60") +
  scale_fill_manual(values = c(negligible = "grey80", small = "#fdcc8a",
                               medium = "#fc8d59", large = "#d7301f"), drop = FALSE) +
  labs(title = "Cliff's delta: Mesenchymal vs Non-Mesenchymal (overall)",
       subtitle = sprintf("Positive = higher in %s (n=%d vs %d).",
                          grp_mes, sum(df_group$cancer_group == grp_mes),
                          sum(df_group$cancer_group == grp_non)),
       x = "Cliff's delta", y = NULL, fill = "Effect size") +
  theme_bw()
```

# 2. ALT-status composition by group

```{r alt-composition}
alt_table <- df_group %>%
  count(cancer_group, alt_status) %>%
  group_by(cancer_group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

alt_table %>%
  ggplot(aes(x = cancer_group, y = prop, fill = alt_status)) +
  geom_col(position = "stack", alpha = 0.85) +
  geom_text(aes(label = paste0(n, "\n(", scales::percent(prop, 1), ")")),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c("ALT-low" = "grey70", "ALT-high" = "firebrick")) +
  labs(title = "ALT status composition by cancer group",
       x = NULL, y = "Proportion", fill = NULL) +
  theme_bw()
```

```{r alt-prevalence}
set.seed(21)
alt_high_mes <- as.numeric(df_group$alt_status[df_group$cancer_group == grp_mes] == "ALT-high")
alt_high_non <- as.numeric(df_group$alt_status[df_group$cancer_group == grp_non] == "ALT-high")

boot_diff <- replicate(2000, {
  s_mes <- sample(alt_high_mes, length(alt_high_mes), replace = TRUE)
  s_non <- sample(alt_high_non, length(alt_high_non), replace = TRUE)
  mean(s_mes) - mean(s_non)
})
boot_diff <- boot_diff[is.finite(boot_diff)]

p_mes_obs <- mean(alt_high_mes); p_non_obs <- mean(alt_high_non)

cat("ALT-high prevalence by cancer group\n")
cat(sprintf("  %s: %.1f%% (%d/%d)\n", grp_mes, 100 * p_mes_obs, sum(alt_high_mes), length(alt_high_mes)))
cat(sprintf("  %s: %.1f%% (%d/%d)\n", grp_non, 100 * p_non_obs, sum(alt_high_non), length(alt_high_non)))
cat(sprintf("\nDifference (mes - non):\n"))
cat(sprintf("  observed: %+.3f\n", p_mes_obs - p_non_obs))
cat(sprintf("  95%% stratified bootstrap CI: [%+.3f, %+.3f]\n",
            quantile(boot_diff, 0.025), quantile(boot_diff, 0.975)))
```

# 3. Continuous variables by group, stratified by ALT status

The tTF/bTF pattern applied to every continuous variable: an **overall**
view (top, groups compared directly) and an **ALT-stratified** view
(bottom, faceted by ALT status) to test whether the group effect
survives holding ALT fixed. Plots carry only the significance star
(`***` \< 0.001, `**` \< 0.01, `*` \< 0.05, BH-adjusted); exact numbers
are in the table.

```{r continuous-table}
fmt_grp(filter(grp_tests, var %in% continuous_vars))
```

```{r alt-strat-fn}
# Stars only (BH q), placed at the top of each panel — table holds the numbers.
alt_strat_plot <- function(v, ylab = v) {
  dd    <- df_group %>% mutate(value = .data[[v]])
  ann_o <- grp_tests %>% filter(var == v, stratum == "overall")
  ann_a <- grp_tests %>% filter(var == v, stratum %in% c("ALT-low", "ALT-high")) %>%
    mutate(alt_status = factor(as.character(stratum), levels = levels(dd$alt_status)))

  p_overall <- ggplot(dd, aes(x = cancer_group, y = value, fill = cancer_group)) +
    geom_violin(alpha = 0.4, color = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.15, size = 0.9) +
    geom_text(data = ann_o, aes(x = 1.5, y = Inf, label = signif),
              inherit.aes = FALSE, vjust = 1.5, size = 4.5) +
    scale_fill_manual(values = grp_colors) +
    labs(title = sprintf("%s by cancer group", v), x = NULL, y = ylab) +
    theme_bw() + theme(legend.position = "none")

  p_alt <- ggplot(dd, aes(x = cancer_group, y = value, fill = cancer_group)) +
    geom_violin(alpha = 0.4, color = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.15, size = 0.8) +
    geom_text(data = ann_a, aes(x = 1.5, y = Inf, label = signif),
              inherit.aes = FALSE, vjust = 1.5, size = 4.5) +
    facet_wrap(~ alt_status) +
    scale_fill_manual(values = grp_colors) +
    labs(title = sprintf("%s by group, stratified by ALT status", v), x = NULL, y = ylab) +
    theme_bw() + theme(legend.position = "none", strip.text = element_text(face = "bold"))

  p_overall / p_alt
}
```

```{r continuous-plots, results='asis', fig.height=7}
for (v in continuous_vars) {
  cat("\n\n## ", v, "\n\n")
  print(alt_strat_plot(v))
}
```

The confounder is ALT composition: - Telomere insertion is essentially
an ALT phenomenon: in ALT-low \~75–94% of samples are exactly 0; in
ALT-high the median jumps to \~0.07–0.10. \
\
Mesenchymal is hugely ALT-enriched: 37.5% ALT-high (81/216) vs only 4.9%
(84/1726) in non-mesenchymal.

So the overall "mesenchymal has higher insertion" is almost entirely
"mesenchymal contains far more ALT-high samples, and ALT-high samples
insert."

Once you hold ALT fixed, the marginal effect disappears: - In ALT-low,
mesenchymal actually inserts less (94% zeros vs 75%) → significant
negative δ.

In ALT-high, no detectable group difference (and only 81 vs 84 samples →
low power, which is why p=0.10 despite δ=0.147).

Interpretation: the group effect on telomere_insertion does not survive
ALT stratification — it's a byproduct of ALT prevalence differing
between groups.

# 4. Singleton motifs — all 9 together

The 9 `*_singleton_dist` motifs share a common scale, so they are shown
jointly in two stacked 3×3 grids — the tTF/bTF top/bottom layout: an
overall grid (group comparison) and an ALT-faceted grid below. Stars
only on the plots; numbers in the tables.

```{r singleton-long}
sing_long <- df_group %>%
  select(cancer_group, alt_status, all_of(singleton_vars)) %>%
  pivot_longer(all_of(singleton_vars), names_to = "motif", values_to = "value") %>%
  mutate(motif = factor(sub("_singleton_dist$", "", motif)))
```

## Overall — 3×3 grid by cancer group

```{r singleton-overall-table}
fmt_grp(filter(grp_tests, var %in% singleton_vars, stratum == "overall"), drop_suffix = TRUE)
```

```{r singleton-overall, fig.height=8, fig.width=10}
sing_overall_annot <- grp_tests %>%
  filter(var %in% singleton_vars, stratum == "overall") %>%
  mutate(motif = factor(sub("_singleton_dist$", "", var)))

ggplot(sing_long, aes(x = cancer_group, y = value, fill = cancer_group)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 0.5) +
  geom_text(data = sing_overall_annot, aes(x = 1.5, y = Inf, label = signif),
            inherit.aes = FALSE, vjust = 1.5, size = 4) +
  facet_wrap(~ motif, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = grp_colors) +
  labs(title = "Singleton motifs by cancer group (overall)",
       subtitle = sprintf("%s in red.", grp_mes),
       x = NULL, y = "singleton_dist") +
  theme_bw() + theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

## Stratified by ALT status — 3×3 grid

```{r singleton-alt-table}
fmt_grp(filter(grp_tests, var %in% singleton_vars, stratum %in% c("ALT-low", "ALT-high")),
        drop_suffix = TRUE)
```

```{r singleton-alt, fig.height=9, fig.width=12}
sing_alt_annot <- grp_tests %>%
  filter(var %in% singleton_vars, stratum %in% c("ALT-low", "ALT-high")) %>%
  mutate(motif = factor(sub("_singleton_dist$", "", var)),
         alt_status = factor(as.character(stratum), levels = levels(sing_long$alt_status)))

ggplot(sing_long, aes(x = alt_status, y = value, fill = cancer_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(width = 0.8)) +
  geom_text(data = sing_alt_annot, aes(x = alt_status, y = Inf, label = signif),
            inherit.aes = FALSE, vjust = 1.5, size = 3.5) +
  facet_wrap(~ motif, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = grp_colors) +
  labs(title = "Singleton motifs by group, stratified by ALT status",
       subtitle = sprintf("%s in red. x = ALT status; colour = cancer group. Star compares groups within each stratum.", grp_mes),
       x = NULL, y = "singleton_dist", fill = NULL) +
  theme_bw() + theme(legend.position = "top", strip.text = element_text(face = "bold"))
```

# 5. Correlation structure differences

```{r correlation}
cor_mes <- df_group %>% filter(cancer_group == grp_mes) %>%
  select(all_of(numeric_vars)) %>% cor(use = "pairwise.complete.obs")
cor_non <- df_group %>% filter(cancer_group == grp_non) %>%
  select(all_of(numeric_vars)) %>% cor(use = "pairwise.complete.obs")
cor_diff <- cor_mes - cor_non

cor_diff_long <- cor_diff %>%
  as.data.frame() %>% rownames_to_column("feat1") %>%
  pivot_longer(-feat1, names_to = "feat2", values_to = "diff") %>%
  filter(feat1 < feat2) %>% arrange(desc(abs(diff)))

cat("Top pairwise correlation differences (mes - non):\n")
print(head(cor_diff_long, 15))

cor_diff %>%
  as.data.frame() %>% rownames_to_column("feat1") %>%
  pivot_longer(-feat1, names_to = "feat2", values_to = "diff") %>%
  ggplot(aes(x = feat1, y = feat2, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation structure difference",
       subtitle = sprintf("cor(%s) - cor(%s). Red = stronger in mesenchymal.", grp_mes, grp_non),
       x = NULL, y = NULL, fill = "Δ cor") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))
```
