---
title: "TF-blood_vs_primary"
format: pdf
date: today
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Setup and data loading

```{r setup}
library(tidyverse)
library(here)
library(corrplot)
data <- readxl::read_xlsx(here('data', 'processed', 'PCAWG_primary.xlsx'))
clean_data <- data
summary(clean_data)
```

# Distributions

## Distributions - summary

```{r summary-distributions}
summary(clean_data$tf_blood_rate)
summary(clean_data$tf_primary_rate)

zero_blood <- sum(clean_data$tf_blood_rate == 0)/nrow(clean_data) * 100
zero_primary <- sum(clean_data$tf_primary_rate == 0)/nrow(clean_data) * 100

cat("Percentuale di zeri in tf_blood_rate: ", zero_blood, "%\n")

cat("Percentuale di zeri in tf_primary_rate: ", zero_primary, "%\n")
```

## Distributions - visual

```{r visual-distributions}
clean_data %>% 
  ggplot(aes(x = tf_blood_rate)) +
  geom_histogram(binwidth = 0.01, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of tf_blood_rate", x = "tf_blood_rate", y = "Frequency")

clean_data %>% 
  ggplot(aes(x = tf_primary_rate)) +
  geom_histogram(binwidth = 0.1, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of tf_primary_rate", x = "tf_primary_rate", y = "Frequency")

clean_data %>% 
  ggplot(aes(x = "", y = tf_blood_rate)) +
  geom_boxplot(fill = "blue", color = "black") +
  labs(title = "Boxplot of tf_blood_rate", y = "tf_blood_rate")

clean_data %>% 
  ggplot(aes(x = "", y = tf_primary_rate)) +
  geom_boxplot(fill = "green", color = "black") +
  labs(title = "Boxplot of tf_primary_rate", y = "tf_primary_rate")
```

# Scatter plots

## Scatterplot raw

```{r scatterplot-raw}
clean_data %>% 
  ggplot(aes(tf_blood_rate, tf_primary_rate)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  labs(title = "Scatterplot tf_blood_rate vs tf_primary_rate", x = "tf_blood_rate", y = "tf_primary_rate")
```

## Scatterplot log1p transformed

```{r scatterplot-log1p}
clean_data %>% 
  ggplot(aes(log1p(tf_blood_rate), log1p(tf_primary_rate), color = alt_status)) +
  geom_point(alpha = 0.5) +
  labs(title = "Scatterplot log1p(tf_blood_rate) vs log1p(tf_primary_rate)", x = "log1p(tf_blood_rate)", y = "log1p(tf_primary_rate)")
```

# Distinction zero vs positive

```{r zero-vs-positive}
clean_data <- clean_data %>% 
  mutate(blood_pos = ifelse(tf_blood_rate > 0, 1, 0),
         primary_pos = ifelse(tf_primary_rate > 0, 1, 0))

table(clean_data$blood_pos, clean_data$primary_pos, dnn = c('blood_pos', 'primary_pos'))
chisq.test(table(clean_data$blood_pos, clean_data$primary_pos))

ggplot(clean_data, aes(x = as.factor(blood_pos), fill = as.factor(primary_pos))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of signal presence between blood and primary", x = "blood_pos", y = "Proporzione") +
  scale_fill_manual(values = c("red", "green"), labels = c("0", "1"))
```

# Discordant samples: primary TF = 0, blood TF \> 0

## Definition and ALT status enrichment

```{r discordant-alt}
# Samples with no detectable primary TF but positive blood TF
discordant <- clean_data %>%
  mutate(group = case_when(
    primary_pos == 0 & blood_pos == 1 ~ "primary=0, blood>0",
    primary_pos == 1 & blood_pos == 1 ~ "both>0",
    primary_pos == 1 & blood_pos == 0 ~ "primary>0, blood=0",
    TRUE ~ "both=0"
  ))

cat("Group counts:\n")
print(table(discordant$group))

# Focus on discordant group vs the rest (excluding both=0 as uninformative)
disc_subset <- discordant %>%
  filter(group %in% c("primary=0, blood>0", "both>0", "primary>0, blood=0")) %>%
  mutate(is_discordant = group == "primary=0, blood>0")

cat("\nALT status in discordant group (primary=0, blood>0):\n")
print(table(discordant$alt_status[discordant$group == "primary=0, blood>0"]))

cat("\nALT status in non-discordant groups:\n")
print(table(discordant$alt_status[discordant$group != "primary=0, blood>0"]))

# Fisher's exact test: is ALT-low enriched in discordant samples?
ct <- table(disc_subset$is_discordant, disc_subset$alt_status)
rownames(ct) <- c("other", "discordant")
print(ct)
fisher_res <- fisher.test(ct)
cat("\nFisher's exact test (discordant vs other groups, ALT-low enrichment):\n")
print(fisher_res)
```

## ALT status proportion plot

```{r discordant-alt-plot}
# Bar chart: ALT status proportions by group
discordant %>%
  filter(group != "both=0") %>%
  ggplot(aes(x = group, fill = alt_status)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("ALT-high" = "#E74C3C", "ALT-low" = "#3498DB")) +
  labs(
    title = "ALT status proportion by TF signal group",
    x = NULL, y = "Proportion", fill = "ALT status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## Discordant rate by cancer type

```{r discordant-cancer-pct}
disc_pct <- discordant %>%
  group_by(cancer_type) %>%
  summarise(
    n_total  = n(),
    n_disc   = sum(group == "primary=0, blood>0"),
    pct_disc = n_disc / n_total * 100,
    .groups  = "drop"
  ) %>%
  filter(n_total >= 10) %>%
  arrange(desc(pct_disc))

print(disc_pct)

ggplot(disc_pct, aes(x = pct_disc, y = reorder(cancer_type, pct_disc))) +
  geom_segment(aes(x = 0, xend = pct_disc,
                   y = reorder(cancer_type, pct_disc),
                   yend = reorder(cancer_type, pct_disc)),
               color = "grey70") +
  geom_point(size = 3, color = "#2C3E50") +
  labs(
    title = "Discordant rate by cancer type (primary=0, blood>0)",
    x = "% discordant", y = NULL
  ) +
  theme_minimal()
```

# Spearman correlation

```{r cor-spearman}
positive_data <- clean_data %>% 
  filter(blood_pos == 1 & primary_pos == 1)

cor.test(positive_data$tf_blood_rate[positive_data$blood_pos == 1],
         positive_data$tf_primary_rate[positive_data$primary_pos == 1],
         method = 'spearman')
```

# Relationship visualization

```{r relationship-visualization}
ggplot(positive_data, aes(x = log1p(tf_blood_rate), y = log1p(tf_primary_rate))) +
  geom_point(alpha = 0.5, color = "darkred") +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "tf_blood_rate vs tf_primary_rate", x = "log1p(tf_blood_rate)", y = "log1p(tf_primary_rate)")
```
