---
title: "02_duplicate_columns"
author: "Andrea Polazzi"
date: "`r Sys.Date()`"
output: html_document
---

# Duplicate-column audit

The source dataset contains apparently repeated fields for cancer annotations,
fusion rate, and singleton-order measurements. This notebook determines which
pairs are true duplicates and which represent discordant or mislabeled data.

Load the unmodified source so the duplicate fields can be compared directly.
```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(here)
data <- read_xlsx(here('data', 'raw', 'drivers.xlsx'))
```

First test whether each candidate pair contains equivalent values.

```{r check-duplicates}
dup1 <- c('cancer_type...2', 'cancer_type_code...3', 'Mixed_def...4', 'Tumor_Telomere fusion rate (Gt-end/coverage*purity)...5', 'TTGGGG_singletons_norm_by_all_reads_ORDER...42', 'CATGGG_singletons_norm_by_all_reads_ORDER...50')
dup2 <- c('cancer_type...28', 'cancer_type_code...29', 'Mixed_def...30', 'Tumor_Telomere fusion rate (Gt-end/coverage*purity)...32', 'TTGGGG_singletons_norm_by_all_reads_ORDER...62', 'CATGGG_singletons_norm_by_all_reads_ORDER...70')

res <- mapply(function(col1, col2) {
  isTRUE(all.equal(data[[col1]], data[[col2]], tolerance = 1e-8))
}, dup1, dup2)

res
```

`TRUE` indicates equivalent content within tolerance; discordant pairs are
inspected individually below.

## Cancer type
```{r cancer-type}
unique(data[data$cancer_type...2!=data$cancer_type...28, c('Patient_ID_IF_mod', 'cancer_type...2', 'cancer_type...28', 'primaryTumorLocation')])
```

Here we see the patients that have a discording cancer type description. The cancer types seem somewhat related in the two columns, but there are some strange exceptions (e.g. Patient_3822). **Should discuss merging**

## Cancer type code
```{r cancer-type-code}
unique(data[data$cancer_type_code...3!=data$cancer_type_code...29, c('Patient_ID_IF_mod', 'cancer_type_code...3', 'cancer_type_code...29', 'primaryTumorLocation')])
```

Accordant to the previous comparison (**same patients**).

## Mixed def
```{r mixed-def}
unique(data[data$Mixed_def...4!=data$Mixed_def...30, c('Patient_ID_IF_mod', 'Mixed_def...4', 'Mixed_def...30')])
```

Again, accordant to previous comparisons but there are **more** and **different entries** in disagreement.

## TF rate
```{r tf-rate}
unique(data[
  abs(data$`Tumor_Telomere fusion rate (Gt-end/coverage*purity)...5` - data$`Tumor_Telomere fusion rate (Gt-end/coverage*purity)...32`) > 1e-8, c('Patient_ID_IF_mod', 'Tumor_Telomere fusion rate (Gt-end/coverage*purity)...5', 'Tumor_Telomere fusion rate (Gt-end/coverage*purity)...32')
])
```

No entry is returned, as expected: this is just a **duplicated column**. (CAVEAT: due to floating point differences, we cannot compare with != or abs() > 0. We must allow some microscopic tolerance)

## TTGGGG singletons norm. by all reads ORDER
```{r TTGGGG all-reads}
unique(data[data$TTGGGG_singletons_norm_by_all_reads_ORDER...42!=data$TTGGGG_singletons_norm_by_all_reads_ORDER...62, c('Patient_ID_IF_mod', 'TTGGGG_singletons_norm_by_all_reads_ORDER...42', 'TTGGGG_singletons_norm_by_all_reads_ORDER...62')])
```

Here the behavior is a bit odd: columns have the same name, and should represent something unambiguous and identical (rank by TTGGGG reads normalized over total reads), but **values differ in every patient**. The point here is that the column in the original dataset probably has an error in this column name: it should be 'TTGGGG_singletons_norm_by_**intratel**_ORDER', in agreement with the previous column that reports the singletons norm. by intratel reads. **This column should be renamed**

## CATGGG singletons norm. by all reads ORDER
```{r CATGGG all-reads}
unique(data[data$CATGGG_singletons_norm_by_all_reads_ORDER...50!=data$CATGGG_singletons_norm_by_all_reads_ORDER...70, c('Patient_ID_IF_mod', 'CATGGG_singletons_norm_by_all_reads_ORDER...50', 'CATGGG_singletons_norm_by_all_reads_ORDER...70')])
```

Here we have to be careful: although the columns have the same name and the same values for all patients, there is an **error in the dataset** again. The error is the same as before (first is correct, second should be intratel), but there was no detectable difference in the content because CATGGG reads are 0 in both cases and in every patient. **This column should be renamed**.


# Standardized column names

Apply the shared cleaning function after documenting which duplicates can be
removed and which mislabeled columns require distinct names.
```{r column-rename}
source(here('scripts', 'clean_columns.R'))
tel_data <- clean_drivers_data(data)
colnames(tel_data)
```
