---
title: "03_further_EDA"
author: "Andrea Polazzi"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

Let's do another EDA, but this time let's see if there are any differences correlating **telomere length** to variables such as **TF rates**, **TVR norm. by all reads** or **TVR norm. by intratel. reads**.

First of all, we must load the dataset and the required packages:
```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(here)
data <- read_xlsx(here('data', 'raw', 'drivers.xlsx'))
```

And rename/remove duplicate columns, along with awkwardly named ones:

```{r renaming}
source(here('scripts', 'clean_columns.R'))
clean_data <- clean_drivers_data(data)
```

Let's create the subset that we need (the column `tel_reads/total reads` can be considered as a representative of telomere length):

```{r subset}
tel_data <- unique(
  clean_data[, c(
    'patient_id',
    'tumor_tf_rate',
    'tel_reads/total reads',
    'tel_reads/total reads_ORDER',
    'tvr_all',
    'tvr_intratel'
  )]
)
summary(tel_data)
```

# Correlation between Telomere Length and all other features
## Telomere Length vs TVR norm. by total reads
We choose Spearman as our correlation method since it deals with monotonic, non-parametric relationships. Pearson's correlation has some assumptions that do not hold true in most biological data: normality and absence of outliers
```{r correlation-allreads, warning=FALSE, message=FALSE}
ggplot(tel_data, aes(`tel_reads/total reads`, tvr_all)) +
  geom_point(alpha=0.2) +
  geom_smooth(method = 'lm', se = TRUE)

cor.test(tel_data$`tel_reads/total reads`, tel_data$tvr_all, method = 'spearman')
```
We can see that **telomere length shows much more correlation with TVR norm. by total reads than TF rates does** (ρ = 0.703 vs ρ = 0.093). This strong positive correlation makes sense: samples with longer chromosomes will have an higher proportion of TVRs over the total reads.

## Telomere Length vs TVR norm. by intratel reads
```{r correlation-intratel, warning=FALSE, message=FALSE}
ggplot(tel_data, aes(`tel_reads/total reads`, tvr_intratel)) +
  geom_point(alpha=0.2) +
  geom_smooth(method = 'lm', se = TRUE)

cor.test(tel_data$`tel_reads/total reads`, tel_data$tvr_intratel, method = 'spearman')
```
Here we see a **weak, negative relatioship** (ρ = -0.074), **almost flat**, suggesting that **TVR density inside telomeres is largely independent of telomere length**.

## Telomere Length vs log(1+TF rate)

```{r correlation-TF-rate, warning=FALSE, message=FALSE}
ggplot(tel_data, aes(`tel_reads/total reads`, log1p(tumor_tf_rate))) +
  geom_point(alpha=0.2) +
  geom_smooth(method = 'lm', se = TRUE)

cor.test(tel_data$`tel_reads/total reads`, log1p(tel_data$tumor_tf_rate), method = 'spearman')
```

As regards the TF rate (log transformed) we see that telomere length is **positively yet weakly correlated** with it.

```{r}
model <- lm(log1p(tumor_tf_rate)~`tel_reads/total reads`, data=tel_data)
summary(model)
```

