---
title: "4_metastatic_dataset"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

```{r setup, warning=FALSE}
library(tidyverse, quietly = T)
library(readxl)
library(here)
source(here('scripts', 'clean_columns.R'))
source(here('scripts', 'useful_plots.R'))
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
data <- read_xlsx(here('data', 'raw', 'drivers.xlsx'))
metastatic_mesenchymal <- read_xlsx(here('data', 'processed', 'metastatic_red_edit_mesenchymal.xlsx'))
clean_data <- clean_drivers_data(data)

clean_data <- clean_data %>% 
  distinct(patient_id, .keep_all = T) %>% 
  select(tumor_tf_rate, `Intratel_reads/total_reads`, `tel_reads/total reads`, TCAGGG_singletons_norm_by_all_reads,
        TGAGGG_singletons_norm_by_all_reads, TTGGGG_singletons_norm_by_all_reads, TTCGGG_singletons_norm_by_all_reads,
        TTTGGG_singletons_norm_by_all_reads, ATAGGG_singletons_norm_by_all_reads, CTAGGG_singletons_norm_by_all_reads,
        GTAGGG_singletons_norm_by_all_reads, TAAGGG_singletons_norm_by_all_reads, tvr_all) %>% 
  rename_with(~gsub('_singletons_norm_by_all_reads', '', .)) %>% 
  rename(tel_insertion = `Intratel_reads/total_reads`, tel_content = `tel_reads/total reads`)

summary(clean_data)
```

**Importantly** we don't have info on TERT expression in this dataset. Moreover, TVR motifs are expressed in counts normalized by total reads.

# PCA using metastatic mesenchymal groups

Here we repeat the PCA/biplot workflow used in `02_ALT_status_full.Rmd`, but using the processed metastatic dataset with mesenchymal grouping. The PCA is run on one row per patient, with samples coloured by `tumor_origin`.

```{r metastatic-mesenchymal-pca-data}
metastatic_features <- c(
  "tumor_tf_rate",
  "blood_tf_rate",
  "intrachrom_reads_total_reads",
  grep("_singleton_dist$", names(metastatic_mesenchymal), value = TRUE)
)

metastatic_pca_data <- metastatic_mesenchymal %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  mutate(tumor_origin = factor(tumor_origin)) %>%
  select(tumor_origin, all_of(metastatic_features)) %>%
  drop_na()

summary(metastatic_pca_data)
```

## PCA

```{r pca-metastatic-mesenchymal}
res_met <- pca_plot(
  metastatic_pca_data,
  coloring = metastatic_pca_data$tumor_origin,
  title = paste0("PCA - metastatic tumor origin (n = ", nrow(metastatic_pca_data), ")")
)

res_met$plot
summary(res_met$pca)
res_met$pca$rotation
ggsave(here('outputs', 'PCA_metastatic_tumor_origin.pdf'), res_met$plot, width = 7, height = 6)
```

## Biplot

```{r biplot-metastatic-mesenchymal}
comp_met <- ifelse(
  grepl("singleton_dist", metastatic_features),
  sub("_singleton_dist", "", metastatic_features),
  metastatic_features
)

bip_met <- pca_biplot(
  metastatic_pca_data,
  coloring = metastatic_pca_data$tumor_origin,
  scale_arrows = 10,
  compnames = comp_met,
  title = "PCA biplot - metastatic tumor origin"
)

bip_met$plot
ggsave(here('outputs', 'PCA_biplot_metastatic_tumor_origin.pdf'), bip_met$plot, width = 7, height = 6)
```

# Include also TERT mutation and cancer type info 

```{r filter}
data_filt <- data %>% 
  clean_drivers_data() %>% 
  select(patient_id, tumor_tf_rate, `Intratel_reads/total_reads`, `tel_reads/total reads`, TCAGGG_singletons_norm_by_all_reads,
        TGAGGG_singletons_norm_by_all_reads, TTGGGG_singletons_norm_by_all_reads, TTCGGG_singletons_norm_by_all_reads,
        TTTGGG_singletons_norm_by_all_reads, ATAGGG_singletons_norm_by_all_reads, CTAGGG_singletons_norm_by_all_reads,
        GTAGGG_singletons_norm_by_all_reads, TAAGGG_singletons_norm_by_all_reads, tvr_all, mixed_def, gene) %>%
  group_by(patient_id) %>%
  mutate(TERT_mut = as.factor(any(gene == "TERT", na.rm = TRUE))) %>%
  ungroup() %>%
  distinct(patient_id, .keep_all = T) %>%
  rename_with(~gsub('_singletons_norm_by_all_reads', '', .)) %>% 
  rename(tel_insertion = `Intratel_reads/total_reads`, tel_content = `tel_reads/total reads`) %>% 
  filter(tumor_tf_rate < 10, tel_content < 0.002)

summary(data_filt)
```

## Removing TVR motifs 

TVR motifs may act as confounders, masking the shared structure that could be explained by TF rate, tel. insertion and tel. content.
Removing them could help us see if they are "masking" the telomere signal.

```{r pca-biplot-noTVRs, echo=FALSE}
tel_data <- data_filt %>%
  select(tumor_tf_rate, tel_content, tel_insertion, TERT_mut, mixed_def) %>% 
  mutate(across(1:3, log1p))

X3 <- scale(tel_data[, 1:3])
pca3 <- prcomp(X3, center = TRUE, scale. = TRUE)
var_explained3 <- summary(pca3)$importance[2,]

# top N cancer types — stesso parametro di prima per coerenza
top_n_types <- 12
top_types <- data_filt %>%
  count(mixed_def) %>%
  slice_max(n, n = top_n_types) %>%
  pull(mixed_def)

scores3 <- data.frame(
  PC1 = pca3$x[, 1],
  PC2 = pca3$x[, 2],
  TERT_mut   = tel_data$TERT_mut,
  cancer_type_plot = factor(
    ifelse(tel_data$mixed_def %in% top_types, tel_data$mixed_def, "Other"),
    levels = c(sort(top_types), "Other")
  )
)

loadings3 <- data.frame(
  PC1 = pca3$rotation[, 1] * 40,
  PC2 = pca3$rotation[, 2] * 40,
  comp_names = rownames(pca3$rotation)
)

type_colors <- setNames(
  c(scales::hue_pal()(top_n_types), "grey80"),
  c(sort(top_types), "Other")
)

ggplot(scores3, aes(PC1, PC2)) +
  geom_point(
    data = filter(scores3, cancer_type_plot == "Other"),
    aes(color = cancer_type_plot),
    shape = 16, size = 1.5, alpha = 0.3
  ) +
  geom_point(
    data = filter(scores3, cancer_type_plot != "Other"),
    aes(color = cancer_type_plot, shape = TERT_mut),
    size = 2, alpha = 0.7
  ) +
  geom_segment(
    data = loadings3,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.15, "cm")),
    color = "grey30", linewidth = 0.4,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = loadings3,
    aes(x = PC1, y = PC2, label = comp_names),
    inherit.aes = FALSE,
    size = 3, color = "grey20",
    box.padding = 0.5
  ) +
  scale_color_manual(values = type_colors) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
  theme_bw() +
  labs(
    title = "PCA biplot on Hartwig dataset - no TVRs",
    x = paste0("PC1 (", round(var_explained3[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(var_explained3[2] * 100, 1), "%)"),
    color = "Cancer type",
    shape = "TERT mut"
  )
```

`tel_content` and `tel_insertion` confirm themselves as **strongly correlated features**, but it is **not the case for** `tumor_tf_rate` as well. It is an **independent "fusion activity" axis**, where orthogonality is not influenced by TVR motifs or scaling issues, but is intrinsic.

```{r correlation}
cor(clean_data$tumor_tf_rate, clean_data$tel_content, method = "spearman")
cor(clean_data$tumor_tf_rate, clean_data$tel_insertion, method = "spearman")
```

Spearman correlations between tumor fusion rate and telomere content (ρ ≈ 0.10) and telomere insertion rate (ρ ≈ 0.22) are low, indicating **weak monotonic association**.

```{r scatter-plots, echo=FALSE}
clean_data %>% 
  filter(tumor_tf_rate < 10, tel_content < 0.002) %>% 
  ggplot(aes(tel_content, tumor_tf_rate)) +
  geom_point(alpha = 0.5)+
  theme_bw()

clean_data %>% 
  filter(tumor_tf_rate < 10) %>% 
  ggplot(aes(tel_insertion, tumor_tf_rate)) +
  geom_point(alpha = 0.5)+
  theme_bw()
```

**Scatter plots obtained after removing strong TF rate and tel_insertion outliers**. They show **no clear relationship** between fusion rate and telomere content or insertion rate, with **most samples clustered near low fusion values** and a **few outliers** driving the range.

## Remove outliers and colour by cancer type + TERT mutation

We now try to discriminate, using all features, among cancer types and patients with registered TERT mutations.

```{r pca-cancer-TERT, echo=FALSE}
pca_vars <- data_filt %>%
  select(where(is.numeric))

pca <- prcomp(pca_vars, center = TRUE, scale. = TRUE)

var_explained <- summary(pca)$importance[2, ]

# Parametro facilmente modificabile
top_n_types <- 12

top_types <- data_filt %>%
  count(mixed_def) %>%
  slice_max(n, n = top_n_types) %>%
  pull(mixed_def)

# Aggiungi la colonna cancer_type_plot agli scores
scores <- as.data.frame(pca$x) %>%
  mutate(
    cancer_type_plot = ifelse(data_filt$mixed_def %in% top_types,
                              data_filt$mixed_def, "Other"),
    cancer_type_plot = factor(cancer_type_plot,
                              levels = c(sort(top_types), "Other")),
    TERT_mut = data_filt$TERT_mut
  )

# Palette: colori distinti per i top types, grigio chiaro per Other
type_colors <- setNames(
  c(scales::hue_pal()(top_n_types), "grey70"),
  c(sort(top_types), "Other")
)

# Loadings scalati — secondari, quindi frecce grigie e sottili
loadings <- as.data.frame(pca$rotation) %>%
  rownames_to_column("feature") %>%
  mutate(
    PC1 = PC1 * 60,
    PC2 = PC2 * 60
  )

ggplot(scores, aes(PC1, PC2)) +
  # Prima "Other" sotto, poi i top types sopra
  geom_point(
    data = filter(scores, cancer_type_plot == "Other"),
    aes(color = cancer_type_plot),
    shape = 16, size = 1.5, alpha = 0.3
  ) +
  geom_point(
    data = filter(scores, cancer_type_plot != "Other"),
    aes(color = cancer_type_plot, shape = TERT_mut),
    size = 2, alpha = 0.7
  ) +
  # Frecce secondarie: grigio, sottili
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.15, "cm")),
    color = "grey50", linewidth = 0.4,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = loadings,
    aes(PC1, PC2, label = feature),
    inherit.aes = FALSE,
    size = 2.5, color = "grey30",
    max.overlaps = 20
  ) +
  scale_color_manual(values = type_colors) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
  theme_bw() +
  coord_equal() +
  labs(
    title = "PCA biplot - Hartwig dataset",
    x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"),
    color = "Cancer type",
    shape = "TERT mut"
  )
```


# Trying with UMAP

PCA finds the main linear directions of variation and is interpretable through feature loadings, whereas **UMAP preserves local neighborhood structure** and is better suited to **reveal non-linear clustering patterns**.

```{r umap}
library(umap)

# Stessa matrice numerica usata per PCA
umap_vars <- data_filt %>%
  select(where(is.numeric))

# Parametri UMAP — questi influenzano molto il risultato, documentali sempre
umap_config <- umap.defaults
umap_config$n_neighbors <- 15   # vicinato locale: valori bassi = cluster più tight
umap_config$min_dist    <- 0.1  # compattezza interna dei cluster
umap_config$random_state <- 21  # riproducibilità

umap_res <- umap(scale(umap_vars), config = umap_config)

scores_umap <- as.data.frame(umap_res$layout) %>%
  rename(UMAP1 = V1, UMAP2 = V2) %>%
  mutate(
    cancer_type_plot = ifelse(data_filt$mixed_def %in% top_types,
                              data_filt$mixed_def, "Other"),
    cancer_type_plot = factor(cancer_type_plot,
                              levels = c(sort(top_types), "Other")),
    TERT_mut = data_filt$TERT_mut
  )

ggplot(scores_umap, aes(UMAP1, UMAP2)) +
  geom_point(
    data = filter(scores_umap, cancer_type_plot == "Other"),
    aes(color = cancer_type_plot),
    shape = 16, size = 1.5, alpha = 0.3
  ) +
  geom_point(
    data = filter(scores_umap, cancer_type_plot != "Other"),
    aes(color = cancer_type_plot, shape = TERT_mut),
    size = 2, alpha = 0.7
  ) +
  scale_color_manual(values = type_colors) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17)) +
  theme_bw() +
  labs(
    title = "UMAP - Hartwig dataset",
    x = "UMAP1",
    y = "UMAP2",
    color = "Cancer type",
    shape = "TERT mut"
  )

ggplot(scores_umap, aes(UMAP1, UMAP2)) +
  geom_point(
    data = filter(scores_umap, TERT_mut == FALSE),
    color = "grey80", size = 1.2, alpha = 0.3, shape = 16
  ) +
  geom_point(
    data = filter(scores_umap, TERT_mut == TRUE),
    color = "#E41A1C", size = 2, alpha = 0.8, shape = 17
  ) +
  theme_bw() +
  labs(
    title = "UMAP - Hartwig dataset, TERT mutation highlighted",
    x = "UMAP1", y = "UMAP2"
  )
```
