---
title: "4_metastatic_dataset"
author: 'Andrea Polazzi'
date: '`r Sys.Date()`'
output: html_document
---

# Setup

This notebook prepares one-row-per-patient metastatic features and examines
whether mesenchymal versus non-mesenchymal tumor origin structures the PCA space.

```{r setup, warning=FALSE}
library(tidyverse, quietly = TRUE)
library(readxl)
library(here)
source(here('scripts', 'clean_columns.R'))
source(here('scripts', 'useful_plots.R'))
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
data <- read_xlsx(here('data', 'raw', 'drivers.xlsx'))
metastatic_mesenchymal <- read_xlsx(here('data', 'processed', 'metastatic_red_edit_mesenchymal.xlsx'))
clean_data <- clean_drivers_data(data)

clean_data <- clean_data %>% 
  distinct(patient_id, .keep_all = TRUE) %>%
  select(tumor_tf_rate, `Intratel_reads/total_reads`, `tel_reads/total reads`, TCAGGG_singletons_norm_by_all_reads,
        TGAGGG_singletons_norm_by_all_reads, TTGGGG_singletons_norm_by_all_reads, TTCGGG_singletons_norm_by_all_reads,
        TTTGGG_singletons_norm_by_all_reads, ATAGGG_singletons_norm_by_all_reads, CTAGGG_singletons_norm_by_all_reads,
        GTAGGG_singletons_norm_by_all_reads, TAAGGG_singletons_norm_by_all_reads, tvr_all) %>% 
  rename_with(~gsub('_singletons_norm_by_all_reads', '', .)) %>% 
  rename(tel_insertion = `Intratel_reads/total_reads`, tel_content = `tel_reads/total reads`)

summary(clean_data)
```

TERT expression is unavailable in this cohort. The initial source-table summary
also uses TVR counts normalized by total reads, whereas the PCA below uses the
processed singleton-distance features shared with later metastatic analyses.

# PCA using metastatic mesenchymal groups

Here we repeat the PCA/biplot workflow used in `02_ALT_status_full.Rmd`, but using the processed metastatic dataset with mesenchymal grouping. The PCA is run on one row per patient, with samples coloured by `tumor_origin`.

```{r metastatic-mesenchymal-pca-data}
metastatic_features <- c(
  'tumor_tf_rate',
  'blood_tf_rate',
  grep('_singleton_dist$', names(metastatic_mesenchymal), value = TRUE)
)

metastatic_pca_data <- metastatic_mesenchymal %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  mutate(tumor_origin = factor(tumor_origin)) %>%
  select(tumor_origin, all_of(metastatic_features)) %>%
  drop_na()

summary(metastatic_pca_data)
```

## PCA

```{r pca-metastatic-mesenchymal}
res_met <- pca_plot(
  metastatic_pca_data,
  coloring = metastatic_pca_data$tumor_origin,
  title = paste0('PCA - metastatic tumor origin (n = ', nrow(metastatic_pca_data), ')')
)

res_met$plot
summary(res_met$pca)
res_met$pca$rotation
ggsave(here('outputs', 'PCA_metastatic_tumor_origin.pdf'), res_met$plot, width = 7, height = 6)
```

## Biplot

```{r biplot-metastatic-mesenchymal}
comp_met <- ifelse(
  grepl('singleton_dist', metastatic_features),
  sub('_singleton_dist', '', metastatic_features),
  metastatic_features
)

bip_met <- pca_biplot(
  metastatic_pca_data,
  coloring = metastatic_pca_data$tumor_origin,
  scale_arrows = 15,
  compnames = comp_met,
  title = 'PCA biplot - metastatic tumor origin'
)

bip_met$plot
ggsave(here('outputs', 'PCA_biplot_metastatic_tumor_origin.pdf'), bip_met$plot, width = 7, height = 6)
```
