---
title: "6_TVRs_comparison"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

# Setup

```{r setup}
library(tidyverse)
library(here)
library(readxl)
library(ggpubr)
library(rstatix)
library(knitr)

met_intrachrom <- read_xlsx(here('data', 'processed', 'metastatic_red_edit_singleton_dist.xlsx'))
met_mesenchymal <- read_xlsx(here('data', 'processed', 'metastatic_red_edit_mesenchymal.xlsx'))
primary <- read_xlsx(here('data', 'processed', 'PCAWG_primary+genes.xlsx'))
```

# TVR singleton distance by primary tumor group (primary)

```{r tvr-singleton-dist-primary, fig.width = 14, fig.height = 11}
data_prim <- primary %>%
  select(donor_id, ends_with("_singleton_dist"), cancer_type, tf_primary_rate, tf_blood_rate) %>%
  filter(complete.cases(.)) %>%
  mutate(
    tumor_group = case_when(
      str_detect(cancer_type, regex("melanoma", ignore_case = TRUE)) ~ "Melanoma",
      str_detect(cancer_type, regex("soft.?tissue|bone|sarc", ignore_case = TRUE)) ~ "Sarcoma",
      str_detect(cancer_type, regex("adenoca|SCC|HCC|RCC|TCC|DCIS|lobular|biliary|eso|cervix|bladder|colon|rect|uterus|breast|lung|panc|prost|stomach|thyroid|ovary|head|kidney|liver", ignore_case = TRUE)) ~ "Carcinoma",
      TRUE ~ "Other"
    ),
    tumor_group = factor(tumor_group, levels = c("Carcinoma", "Melanoma", "Sarcoma", "Other"))
  )

motif_long_prim <- data_prim %>%
  pivot_longer(
    cols = ends_with("_singleton_dist"),
    names_to = "motif",
    values_to = "singleton_dist"
  ) %>%
  mutate(motif = str_remove(motif, "_singleton_dist"))

group_labels_prim <- data_prim %>%
  count(tumor_group) %>%
  mutate(label = paste0(tumor_group, " (n=", n, ")")) %>%
  { setNames(.$label, .$tumor_group) }

kw_tests_prim <- motif_long_prim %>%
  group_by(motif) %>%
  kruskal_test(singleton_dist ~ tumor_group) %>%
  adjust_pvalue(method = "BH") %>%
  mutate(label = paste0("KW p=", signif(p.adj, 2)))

kw_sig_motifs_prim <- kw_tests_prim %>%
  filter(p.adj < 0.05) %>%
  pull(motif)

dunn_sig_prim <- motif_long_prim %>%
  filter(motif %in% kw_sig_motifs_prim) %>%
  group_by(motif) %>%
  dunn_test(singleton_dist ~ tumor_group, p.adjust.method = "BH") %>%
  filter(p.adj.signif != "ns") %>%
  add_xy_position(x = "tumor_group", scales = "free", step.increase = 0.12)

ggplot(motif_long_prim, aes(tumor_group, singleton_dist, fill = tumor_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.65, width = 0.65) +
  geom_jitter(
    aes(color = tumor_group),
    width = 0.18,
    alpha = 0.25,
    size = 0.7,
    show.legend = FALSE
  ) +
  stat_summary(
    fun = median,
    geom = "point",
    shape = 23,
    size = 2.2,
    fill = "white",
    color = "black"
  ) +
  stat_pvalue_manual(
    dunn_sig_prim,
    label = "p.adj.signif",
    tip.length = 0.01,
    inherit.aes = FALSE,
    hide.ns = TRUE
  ) +
  geom_text(
    data = kw_tests_prim,
    aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.4,
    size = 2.6,
    fontface = "italic",
    color = "grey30"
  ) +
  facet_wrap(~ motif, scales = "free_y", ncol = 3) +
  scale_fill_brewer(palette = "Set2", labels = group_labels_prim) +
  scale_color_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Singleton distance",
    fill = "Tumor group",
    title = "TVR singleton distance distributions by tumor group (primary)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

# TVR singleton distance by mesenchymal vs non-mesenchymal origin (primary)

```{r tvr-mes-primary, fig.width = 14, fig.height = 11}
data_mes_prim <- primary %>%
  select(donor_id, ends_with("_singleton_dist"), cancer_group, tf_primary_rate, tf_blood_rate) %>%
  filter(if_any(ends_with("_singleton_dist"), ~ !is.na(.))) %>%
  mutate(cancer_group = factor(cancer_group,
                               levels = c("Non_Mesenchymal_origin", "Mesenchymal_origin")))

motif_long_mes_prim <- data_mes_prim %>%
  pivot_longer(
    cols = ends_with("_singleton_dist"),
    names_to = "motif",
    values_to = "singleton_dist"
  ) %>%
  mutate(motif = str_remove(motif, "_singleton_dist"))

group_labels_mes_prim <- data_mes_prim %>%
  count(cancer_group) %>%
  mutate(label = paste0(cancer_group, " (n=", n, ")")) %>%
  { setNames(.$label, .$cancer_group) }

kw_tests_mes_prim <- motif_long_mes_prim %>%
  group_by(motif) %>%
  kruskal_test(singleton_dist ~ cancer_group) %>%
  adjust_pvalue(method = "BH") %>%
  mutate(label = paste0("KW p=", signif(p.adj, 2)))

kw_sig_motifs_mes_prim <- kw_tests_mes_prim %>%
  filter(p.adj < 0.05) %>%
  pull(motif)

dunn_sig_mes_prim <- motif_long_mes_prim %>%
  filter(motif %in% kw_sig_motifs_mes_prim) %>%
  group_by(motif) %>%
  dunn_test(singleton_dist ~ cancer_group, p.adjust.method = "BH") %>%
  filter(p.adj.signif != "ns") %>%
  add_xy_position(x = "cancer_group", scales = "free", step.increase = 0.12)

mes_colors      <- c("Non_Mesenchymal_origin" = "grey60",  "Mesenchymal_origin" = "firebrick")
mes_colors_dark <- c("Non_Mesenchymal_origin" = "grey40",  "Mesenchymal_origin" = "firebrick4")

ggplot(motif_long_mes_prim, aes(cancer_group, singleton_dist, fill = cancer_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.65, width = 0.65) +
  geom_jitter(
    aes(color = cancer_group),
    width = 0.18, alpha = 0.25, size = 0.7,
    show.legend = FALSE
  ) +
  stat_summary(
    fun = median, geom = "point",
    shape = 23, size = 2.2, fill = "white", color = "black"
  ) +
  stat_pvalue_manual(
    dunn_sig_mes_prim,
    label = "p.adj.signif",
    tip.length = 0.01,
    inherit.aes = FALSE,
    hide.ns = TRUE
  ) +
  geom_text(
    data = kw_tests_mes_prim,
    aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05, vjust = 1.4,
    size = 2.6, fontface = "italic", color = "grey30"
  ) +
  facet_wrap(~ motif, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = mes_colors, labels = group_labels_mes_prim) +
  scale_color_manual(values = mes_colors_dark) +
  labs(
    x = NULL,
    y = "Singleton distance",
    fill = "Cancer group",
    title = "TVR singleton distance: Mesenchymal vs Non-Mesenchymal origin (primary)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

# TVR singleton distance by primary tumor group (metastatic)

```{r tvr-singleton-dist, fig.width = 14, fig.height = 11}
data <- met_intrachrom %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  mutate(
    primary_tumor_group = case_when(
      # Trust primary_tumor_type when available
      !is.na(primary_tumor_type) & primary_tumor_type == "Carcinoma" ~ "Carcinoma",
      !is.na(primary_tumor_type) & primary_tumor_type == "Melanoma"  ~ "Melanoma",
      !is.na(primary_tumor_type) & str_detect(primary_tumor_type, regex("sarcoma|gastrointestinal stromal", ignore_case = TRUE)) ~ "Sarcoma",
      !is.na(primary_tumor_type) ~ "Other",
      # Fallback to cancer_type when primary_tumor_type is NA — only unambiguous keywords
      str_detect(cancer_type, regex("melanoma",  ignore_case = TRUE)) ~ "Melanoma",
      str_detect(cancer_type, regex("sarcoma|gastrointestinal stromal", ignore_case = TRUE)) ~ "Sarcoma",
      str_detect(cancer_type, regex("carcinoma", ignore_case = TRUE)) ~ "Carcinoma",
      TRUE ~ "Other"
    ),
    primary_tumor_group = factor(primary_tumor_group, levels = c("Carcinoma", "Melanoma", "Sarcoma", "Other"))
  ) %>%
  select(patient_id, ends_with("_singleton_dist"), primary_tumor_group, tumor_tf_rate, blood_tf_rate) %>%
  filter(if_any(ends_with("_singleton_dist"), ~ !is.na(.)))

motif_long <- data %>%
  pivot_longer(
    cols = ends_with("_singleton_dist"),
    names_to = "motif",
    values_to = "singleton_dist"
  ) %>%
  mutate(motif = str_remove(motif, "_singleton_dist"))

# Per-group sample sizes for the legend (one row per patient).
group_labels <- data %>%
  count(primary_tumor_group) %>%
  mutate(label = paste0(primary_tumor_group, " (n=", n, ")")) %>%
  { setNames(.$label, .$primary_tumor_group) }

# Omnibus Kruskal-Wallis per motif (BH-adjusted across motifs). Used both as the
# gate for the post-hoc and for the per-facet label.
kw_tests <- motif_long %>%
  group_by(motif) %>%
  kruskal_test(singleton_dist ~ primary_tumor_group) %>%
  adjust_pvalue(method = "BH") %>%
  mutate(label = paste0("KW p=", signif(p.adj, 2)))

# Protected post-hoc: only run pairwise tests for motifs where the omnibus is
# significant.
kw_sig_motifs <- kw_tests %>%
  filter(p.adj < 0.05) %>%
  pull(motif)

# Dunn's post-hoc (BH-adjusted) identifies WHICH groups differ within each
# significant motif. Pooled ranking keeps all pairwise comparisons on a common
# rank scale. Brackets are drawn only for significant pairs.
dunn_sig <- motif_long %>%
  filter(motif %in% kw_sig_motifs) %>%
  group_by(motif) %>%
  dunn_test(singleton_dist ~ primary_tumor_group, p.adjust.method = "BH") %>%
  filter(p.adj.signif != "ns") %>%
  add_xy_position(x = "primary_tumor_group", scales = "free", step.increase = 0.12)

ggplot(motif_long, aes(primary_tumor_group, singleton_dist, fill = primary_tumor_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.65, width = 0.65) +
  geom_jitter(
    aes(color = primary_tumor_group),
    width = 0.18,
    alpha = 0.25,
    size = 0.7,
    show.legend = FALSE
  ) +
  stat_summary(
    fun = median,
    geom = "point",
    shape = 23,
    size = 2.2,
    fill = "white",
    color = "black"
  ) +
  # Dunn's post-hoc brackets (BH-adjusted), only for significant comparisons
  stat_pvalue_manual(
    dunn_sig,
    label = "p.adj.signif",
    tip.length = 0.01,
    inherit.aes = FALSE,
    hide.ns = TRUE
  ) +
  # Omnibus Kruskal-Wallis p (BH-adjusted), one small label per facet top-right
  geom_text(
    data = kw_tests,
    aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.4,
    size = 2.6,
    fontface = "italic",
    color = "grey30"
  ) +
  facet_wrap(~ motif, scales = "free_y", ncol = 3) +
  scale_fill_brewer(palette = "Set2", labels = group_labels) +
  scale_color_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Singleton distance",
    fill = "Primary tumor group",
    title = "TVR singleton distance distributions by primary tumor group"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

# TVR singleton distance by mesenchymal vs non-mesenchymal origin (metastatic)

```{r tvr-mes-metastatic, fig.width = 14, fig.height = 11}
data_mes_met <- met_mesenchymal %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  filter(tumor_origin %in% c("Mesenchymal", "Non_mesenchymal")) %>%
  mutate(tumor_origin = factor(tumor_origin,
                               levels = c("Non_mesenchymal", "Mesenchymal"))) %>%
  select(patient_id, ends_with("_singleton_dist"), tumor_origin, tumor_tf_rate, blood_tf_rate) %>%
  filter(if_any(ends_with("_singleton_dist"), ~ !is.na(.)))

motif_long_mes_met <- data_mes_met %>%
  pivot_longer(
    cols = ends_with("_singleton_dist"),
    names_to = "motif",
    values_to = "singleton_dist"
  ) %>%
  mutate(motif = str_remove(motif, "_singleton_dist"))

group_labels_mes_met <- data_mes_met %>%
  count(tumor_origin) %>%
  mutate(label = paste0(tumor_origin, " (n=", n, ")")) %>%
  { setNames(.$label, .$tumor_origin) }

kw_tests_mes_met <- motif_long_mes_met %>%
  group_by(motif) %>%
  kruskal_test(singleton_dist ~ tumor_origin) %>%
  adjust_pvalue(method = "BH") %>%
  mutate(label = paste0("KW p=", signif(p.adj, 2)))

kw_sig_motifs_mes_met <- kw_tests_mes_met %>%
  filter(p.adj < 0.05) %>%
  pull(motif)

dunn_sig_mes_met <- motif_long_mes_met %>%
  filter(motif %in% kw_sig_motifs_mes_met) %>%
  group_by(motif) %>%
  dunn_test(singleton_dist ~ tumor_origin, p.adjust.method = "BH") %>%
  filter(p.adj.signif != "ns") %>%
  add_xy_position(x = "tumor_origin", scales = "free", step.increase = 0.12)

mes_colors      <- c("Non_mesenchymal" = "grey60", "Mesenchymal" = "firebrick")
mes_colors_dark <- c("Non_mesenchymal" = "grey40", "Mesenchymal" = "firebrick4")

ggplot(motif_long_mes_met, aes(tumor_origin, singleton_dist, fill = tumor_origin)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.65, width = 0.65) +
  geom_jitter(
    aes(color = tumor_origin),
    width = 0.18, alpha = 0.25, size = 0.7,
    show.legend = FALSE
  ) +
  stat_summary(
    fun = median, geom = "point",
    shape = 23, size = 2.2, fill = "white", color = "black"
  ) +
  stat_pvalue_manual(
    dunn_sig_mes_met,
    label = "p.adj.signif",
    tip.length = 0.01,
    inherit.aes = FALSE,
    hide.ns = TRUE
  ) +
  geom_text(
    data = kw_tests_mes_met,
    aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05, vjust = 1.4,
    size = 2.6, fontface = "italic", color = "grey30"
  ) +
  facet_wrap(~ motif, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = mes_colors, labels = group_labels_mes_met) +
  scale_color_manual(values = mes_colors_dark) +
  labs(
    x = NULL,
    y = "Singleton distance",
    fill = "Tumor origin",
    title = "TVR singleton distance: Mesenchymal vs Non-Mesenchymal origin (metastatic)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

# Singleton-distribution differences in top-tTF samples — Primary vs Metastatic

```{r setup-topTF}
# --- Load data -------------------------------------------------------------
# primary already loaded in setup chunk above

# Metastatic file is long-format (one row per driver gene); collapse to one
# row per patient before any per-sample analysis.
metastatic <- readxl::read_xlsx(
  here("data", "processed", "metastatic_red_edit_singleton_dist.xlsx")
) |>
  distinct(patient_id, .keep_all = TRUE)

# --- Shared motif set ------------------------------------------------------
# The 9 TVR motifs are stored under identical "<MOTIF>_singleton_dist" names in
# both datasets, so the comparison is strictly parallel.
motifs <- c("ATAGGG", "CTAGGG", "GTAGGG", "TAAGGG",
            "TCAGGG", "TGAGGG", "TTCGGG", "TTGGGG", "TTTGGG")
sd_col <- function(m) paste0(m, "_singleton_dist")

group_colors <- c("rest" = "#4E79A7", "top" = "#E15759")
```

```{r helpers}
# Label samples as "top" vs "rest" by an upper-tail TF-rate quantile.
# Ties at the threshold fall into "rest" (strict > cutoff for "top").
assign_top <- function(df, tf_var, top_frac) {
  cutoff <- quantile(df[[tf_var]], probs = 1 - top_frac, na.rm = TRUE)
  df |> mutate(tf_group = factor(if_else(.data[[tf_var]] > cutoff, "top", "rest"),
                                 levels = c("rest", "top")))
}

# Per-motif Wilcoxon rank-sum (top vs rest) on the raw singleton_dist value.
# NAs are dropped per motif. BH/FDR correction is applied across the 9 motifs
# within this single dataset x threshold block.
test_motifs <- function(df, top_frac) {
  map_dfr(motifs, function(m) {
    col <- sd_col(m)
    sub <- df |> filter(!is.na(.data[[col]])) |> select(value = all_of(col), tf_group)
    top  <- sub$value[sub$tf_group == "top"]
    rest <- sub$value[sub$tf_group == "rest"]
    wt <- wilcox.test(top, rest)
    tibble(
      motif       = m,
      n_top       = length(top),
      n_rest      = length(rest),
      median_top  = median(top),
      median_rest = median(rest),
      median_diff = median(top) - median(rest),
      W           = unname(wt$statistic),
      p           = wt$p.value
    )
  }) |>
    mutate(p.adj = p.adjust(p, method = "BH"),
           signif = cut(p.adj,
                        breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                        labels = c("***", "**", "*", "ns"))) |>
    relocate(p.adj, signif, .after = p)
}

fmt_table <- function(tbl) {
  tbl |>
    mutate(across(c(median_top, median_rest, median_diff), \(x) round(x, 3)),
           p     = signif(p, 3),
           p.adj = signif(p.adj, 3)) |>
    kable()
}

# Long-format frame for faceted boxplots.
boxplot_df <- function(df) {
  df |>
    select(tf_group, all_of(sd_col(motifs))) |>
    pivot_longer(-tf_group, names_to = "motif", values_to = "value") |>
    mutate(motif = str_remove(motif, "_singleton_dist")) |>
    filter(!is.na(value))
}

plot_motifs <- function(df, res, title) {
  sig_df <- res |> select(motif, label = signif)

  ggplot(boxplot_df(df), aes(tf_group, value, fill = tf_group)) +
    geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3, width = 0.6) +
    geom_text(
      data = sig_df,
      aes(x = 1.5, y = Inf, label = label),
      inherit.aes = FALSE,
      vjust = 1.5,
      size = 4.5
    ) +
    facet_wrap(~ motif, scales = "free_y", nrow = 3) +
    scale_fill_manual(values = group_colors) +
    labs(x = NULL, y = "singleton_dist", title = title) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none")
}
```

This analysis runs the **same comparison independently** in the primary (PCAWG) and metastatic cohorts: for each of the 9 telomere variant repeat (TVR) motifs, does the per-sample `singleton_dist` value differ between the highest-TF-rate samples and the rest? The two datasets are reported side by side so any shared pattern can be judged by eye — there is **no automated cross-dataset test**.

-   **Test:** Wilcoxon rank-sum (Mann–Whitney), `top` vs `rest`, on the raw `singleton_dist` value.
-   **Top group:** highest 5% and highest 10% of tumour TF rate (tTF: `tf_primary_rate` in primary, `tumor_tf_rate` in metastatic), reported separately.
-   **Multiple testing:** Benjamini–Hochberg FDR across the 9 motifs, applied within each dataset × threshold block. Each dataset is corrected on its own.
-   **Direction:** `median_diff = median(top) − median(rest)`.

```{r assign-groups}
prim_5 <- assign_top(primary,    "tf_primary_rate", 0.05)
prim_10 <- assign_top(primary,    "tf_primary_rate", 0.10)
meta_5 <- assign_top(metastatic, "tumor_tf_rate",   0.05)
meta_10 <- assign_top(metastatic, "tumor_tf_rate",   0.10)
```

## Primary (PCAWG)

### Top 5% TF rate

```{r prim-5-tbl}
res_prim_5 <- test_motifs(prim_5, 0.05)
fmt_table(res_prim_5)
```

```{r prim-5-plot, fig.width = 10, fig.height = 8}
plot_motifs(prim_5, res_prim_5, "Primary — singleton_dist, top 5% tTF vs rest")
```

### Top 10% TF rate

```{r prim-10-tbl}
res_prim_10 <- test_motifs(prim_10, 0.10)
fmt_table(res_prim_10)
```

```{r prim-10-plot, fig.width = 10, fig.height = 8}
plot_motifs(prim_10, res_prim_10, "Primary — singleton_dist, top 10% tTF vs rest")
```

## Metastatic

### Top 5% TF rate

```{r meta-5-tbl}
res_meta_5 <- test_motifs(meta_5, 0.05)
fmt_table(res_meta_5)
```

```{r meta-5-plot, fig.width = 10, fig.height = 8}
plot_motifs(meta_5, res_meta_5, "Metastatic — singleton_dist, top 5% tTF vs rest")
```

### Top 10% TF rate

```{r meta-10-tbl}
res_meta_10 <- test_motifs(meta_10, 0.10)
fmt_table(res_meta_10)
```

```{r meta-10-plot, fig.width = 10, fig.height = 8}
plot_motifs(meta_10, res_meta_10, "Metastatic — singleton_dist, top 10% tTF vs rest")
```

# Singleton-distribution differences in top-bTF samples — Primary vs Metastatic

Same analysis as above but stratifying by **blood TF rate** (bTF: `tf_blood_rate` in primary, `blood_tf_rate` in metastatic).

-   **Test:** Wilcoxon rank-sum (Mann–Whitney), `top` vs `rest`, on the raw `singleton_dist` value.
-   **Top group:** highest 5% and highest 10% of blood TF rate (bTF: `tf_blood_rate` in primary, `blood_tf_rate` in metastatic), reported separately.
-   **Multiple testing:** Benjamini–Hochberg FDR across the 9 motifs, applied within each dataset × threshold block. Each dataset is corrected on its own.
-   **Direction:** `median_diff = median(top) − median(rest)`.

```{r assign-groups-bTF}
prim_b5  <- assign_top(primary,    "tf_blood_rate",  0.05)
prim_b10 <- assign_top(primary,    "tf_blood_rate",  0.10)
meta_b5  <- assign_top(metastatic, "blood_tf_rate",  0.05)
meta_b10 <- assign_top(metastatic, "blood_tf_rate",  0.10)
```

## Primary (PCAWG)

### Top 5% bTF rate

```{r prim-b5-tbl}
res_prim_b5 <- test_motifs(prim_b5, 0.05)
fmt_table(res_prim_b5)
```

```{r prim-b5-plot, fig.width = 10, fig.height = 8}
plot_motifs(prim_b5, res_prim_b5, "Primary — singleton_dist, top 5% bTF vs rest")
```

### Top 10% bTF rate

```{r prim-b10-tbl}
res_prim_b10 <- test_motifs(prim_b10, 0.10)
fmt_table(res_prim_b10)
```

```{r prim-b10-plot, fig.width = 10, fig.height = 8}
plot_motifs(prim_b10, res_prim_b10, "Primary — singleton_dist, top 10% bTF vs rest")
```

## Metastatic

### Top 5% bTF rate

```{r meta-b5-tbl}
res_meta_b5 <- test_motifs(meta_b5, 0.05)
fmt_table(res_meta_b5)
```

```{r meta-b5-plot, fig.width = 10, fig.height = 8}
plot_motifs(meta_b5, res_meta_b5, "Metastatic — singleton_dist, top 5% bTF vs rest")
```

### Top 10% bTF rate

```{r meta-b10-tbl}
res_meta_b10 <- test_motifs(meta_b10, 0.10)
fmt_table(res_meta_b10)
```

```{r meta-b10-plot, fig.width = 10, fig.height = 8}
plot_motifs(meta_b10, res_meta_b10, "Metastatic — singleton_dist, top 10% bTF vs rest")
```

# Final comparison — primary vs metastatic

`diff` = `median(top) − median(rest)`; `p.adj` are BH-corrected within each cohort × stratification block.

## tTF

### Top 5%

```{r final-tTF-5}
left_join(
  res_prim_5 |> select(motif, prim_diff = median_diff, prim_p.adj = p.adj, prim_signif = signif),
  res_meta_5 |> select(motif, meta_diff = median_diff, meta_p.adj = p.adj, meta_signif = signif),
  by = "motif"
) |>
  mutate(across(c(prim_diff, meta_diff), \(x) round(x, 3)),
         across(c(prim_p.adj, meta_p.adj), \(x) signif(x, 3))) |>
  kable()
```

### Top 10%

```{r final-tTF-10}
left_join(
  res_prim_10 |> select(motif, prim_diff = median_diff, prim_p.adj = p.adj, prim_signif = signif),
  res_meta_10 |> select(motif, meta_diff = median_diff, meta_p.adj = p.adj, meta_signif = signif),
  by = "motif"
) |>
  mutate(across(c(prim_diff, meta_diff), \(x) round(x, 3)),
         across(c(prim_p.adj, meta_p.adj), \(x) signif(x, 3))) |>
  kable()
```

## bTF

### Top 5%

```{r final-bTF-5}
left_join(
  res_prim_b5 |> select(motif, prim_diff = median_diff, prim_p.adj = p.adj, prim_signif = signif),
  res_meta_b5 |> select(motif, meta_diff = median_diff, meta_p.adj = p.adj, meta_signif = signif),
  by = "motif"
) |>
  mutate(across(c(prim_diff, meta_diff), \(x) round(x, 3)),
         across(c(prim_p.adj, meta_p.adj), \(x) signif(x, 3))) |>
  kable()
```

### Top 10%

```{r final-bTF-10}
left_join(
  res_prim_b10 |> select(motif, prim_diff = median_diff, prim_p.adj = p.adj, prim_signif = signif),
  res_meta_b10 |> select(motif, meta_diff = median_diff, meta_p.adj = p.adj, meta_signif = signif),
  by = "motif"
) |>
  mutate(across(c(prim_diff, meta_diff), \(x) round(x, 3)),
         across(c(prim_p.adj, meta_p.adj), \(x) signif(x, 3))) |>
  kable()
```

# TTTGGG low values by cancer type (one-vs-rest)

```{r tttggg-tests}
tttggg_min_n <- 10

tttggg_data <- bind_rows(
  primary %>%
    select(sample_id = donor_id, cancer_type, TTTGGG_singleton_dist) %>%
    mutate(cohort = "Primary"),
  met_intrachrom %>%
    distinct(patient_id, .keep_all = TRUE) %>%
    select(sample_id = patient_id, cancer_type, TTTGGG_singleton_dist) %>%
    mutate(cohort = "Metastatic")
) %>%
  filter(!is.na(TTTGGG_singleton_dist), !is.na(cancer_type)) %>%
  group_by(cohort, cancer_type) %>%
  mutate(n_cancer_type = n()) %>%
  ungroup() %>%
  filter(n_cancer_type >= tttggg_min_n) %>%
  mutate(cohort = factor(cohort, levels = c("Primary", "Metastatic")))

# One-vs-rest Wilcoxon rank-sum, one-sided lower tail: is this cancer type
# shifted to LOWER TTTGGG_singleton_dist than all other cancer types in the same
# cohort? Continuous test on the raw log2 ratio — no dichotomization, so the
# magnitude of the shift drives the result. Effect size is the rank-biserial
# correlation (positive = lower than the rest of the cohort). BH correction is
# applied separately within each cohort.
tttggg_tests <- tttggg_data %>%
  group_by(cohort) %>%
  group_modify(~ {
    df <- .x
    map_dfr(unique(df$cancer_type), function(ct) {
      a <- df$TTTGGG_singleton_dist[df$cancer_type == ct]
      b <- df$TTTGGG_singleton_dist[df$cancer_type != ct]
      wt <- wilcox.test(a, b, alternative = "less")
      tibble(
        cancer_type   = ct,
        n             = length(a),
        median        = median(a),
        rest_median   = median(b),
        median_diff   = median(a) - median(b),
        rank_biserial = 1 - 2 * unname(wt$statistic) / (length(a) * length(b)),
        p_low         = wt$p.value
      )
    })
  }) %>%
  ungroup() %>%
  group_by(cohort) %>%
  mutate(
    p_low_adj = p.adjust(p_low, method = "BH"),
    low_signal = if_else(p_low_adj < 0.05, "lower than rest", "not significant")
  ) %>%
  ungroup() %>%
  arrange(cohort, median)

tttggg_plot_data <- tttggg_data %>%
  left_join(
    tttggg_tests %>% select(cohort, cancer_type, n, median, low_signal),
    by = c("cohort", "cancer_type")
  ) %>%
  mutate(
    cancer_type_ordered = fct_reorder(
      paste0(cohort, "___", cancer_type, "\n", "n=", n),
      median,
      .fun = first
    )
  )

tam_cols <- c('TERT_amp', 'TERTp_mut', 'TERT_mult', 'TERTp_sv')

tttggg_tam_data <- tttggg_plot_data %>%
  filter(cohort == "Primary") %>%
  left_join(
    primary %>% select(donor_id, all_of(tam_cols)),
    by = c("sample_id" = "donor_id")
  ) %>%
  pivot_longer(
    cols      = all_of(tam_cols),
    names_to  = "tam_type",
    values_to = "tam_present"
  ) %>%
  filter(tam_present == TRUE)
```

`TTTGGG_singleton_dist` is a log2 tumor/control ratio normalized by telomere content. The analysis keeps cancer types with at least `r tttggg_min_n` samples in the corresponding cohort. For each cancer type, a one-sided Wilcoxon rank-sum test asks whether its `TTTGGG_singleton_dist` values are **shifted lower** than all other cancer types in the same cohort, using the raw continuous values (no threshold). The rank-biserial correlation reports the effect size (positive = lower than the rest), and Benjamini-Hochberg correction is applied separately within each cohort. The dashed line at 0 marks no tumor/control difference and is a visual reference only.

```{r tttggg-cancer-type-rank-plot, fig.width = 18, fig.height = 10}
ggplot(tttggg_plot_data, aes(cancer_type_ordered, TTTGGG_singleton_dist)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(
    aes(color = low_signal),
    position = position_jitter(width = 0.22, height = 0),
    alpha = 0.35,
    size = 0.75,
    show.legend = FALSE
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.55,
    fatten = 0,
    color = "black"
  ) +
  geom_point(
    data     = tttggg_tam_data,
    aes(shape = tam_type, fill = tam_type),
    color    = "black",
    size     = 2.4,
    alpha    = 0.7,
    position = position_jitter(width = 0.22, height = 0, seed = 1)
  ) +
  facet_wrap(~ cohort, scales = "free_x", ncol = 1) +
  scale_color_manual(values = c("lower than rest" = "firebrick", "not significant" = "grey65")) +
  scale_shape_manual(
    name   = "TAM (primary)",
    values = c(TERT_amp = 21, TERTp_mut = 24, TERT_mult = 22, TERTp_sv = 23)
  ) +
  scale_fill_manual(
    name   = "TAM (primary)",
    values = c(TERT_amp = "#E69F00", TERTp_mut = "#56B4E9", TERT_mult = "#CC79A7", TERTp_sv = "#009E73")
  ) +
  scale_x_discrete(labels = \(x) str_remove(x, "^.*___")) +
  labs(
    x = NULL,
    y = "TTTGGG singleton_dist log2(tumor/control)",
    title = "TTTGGG by cancer type — ordered by median, lower-tail test highlighted",
    subtitle = paste0(
      "Cancer types with n >= ", tttggg_min_n,
      "; red = significantly lower than rest of cohort (one-vs-rest Wilcoxon, BH < 0.05)"
    )
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 55, hjust = 1, size = 8),
    strip.text = element_text(face = "bold")
  )
```

```{r tttggg-effect-size-plot, fig.width = 15, fig.height = 9}
ggplot(
  tttggg_tests,
  aes(
    x = median_diff,
    y = fct_reorder(paste0(cancer_type, " (n=", n, ")"), median_diff, .desc = TRUE)
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey35") +
  geom_col(aes(fill = low_signal), width = 0.72) +
  facet_wrap(~ cohort, scales = "free_y", ncol = 2) +
  scale_fill_manual(
    values = c(
      "lower than rest" = "firebrick",
      "not significant" = "grey75"
    )
  ) +
  labs(
    x = "Median TTTGGG singleton_dist relative to rest of cohort (log2)",
    y = NULL,
    fill = "Signal",
    title = "How much lower is each cancer type's TTTGGG vs the rest of its cohort",
    subtitle = paste0(
      "Cancer types with n >= ", tttggg_min_n,
      "; negative = lower than cohort rest. Red = one-vs-rest Wilcoxon BH < 0.05"
    )
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r tttggg-test-table}
tttggg_tests %>%
  select(
    cohort, cancer_type, n, median, rest_median, median_diff,
    rank_biserial, p_low, p_low_adj, low_signal
  ) %>%
  mutate(
    across(c(median, rest_median, median_diff, rank_biserial), \(x) round(x, 3)),
    p_low = signif(p_low, 3),
    p_low_adj = signif(p_low_adj, 3)
  ) %>%
  arrange(cohort, p_low_adj) %>%
  kable(caption = "TTTGGG one-vs-rest lower-tail Wilcoxon by cancer type (primary and metastatic)")
```
