---
title: "Singleton-distribution differences in top-TF samples — Primary vs Metastatic"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    df-print: kable
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
library(tidyverse)
library(here)
library(ggpubr)
library(knitr)

# --- Load data -------------------------------------------------------------
primary <- readxl::read_xlsx(here("data", "processed", "PCAWG_primary+genes.xlsx"))

# Metastatic file is long-format (one row per driver gene); collapse to one
# row per patient before any per-sample analysis.
metastatic <- readxl::read_xlsx(
  here("data", "processed", "metastatic_red_edit_singleton_dist.xlsx")
) |>
  distinct(patient_id, .keep_all = TRUE)

# --- Shared motif set ------------------------------------------------------
# The 9 TVR motifs are stored under identical "<MOTIF>_singleton_dist" names in
# both datasets, so the comparison is strictly parallel.
motifs <- c("ATAGGG", "CTAGGG", "GTAGGG", "TAAGGG",
            "TCAGGG", "TGAGGG", "TTCGGG", "TTGGGG", "TTTGGG")
sd_col <- function(m) paste0(m, "_singleton_dist")

group_colors <- c("rest" = "#4E79A7", "top" = "#E15759")
```

```{r helpers}
# Label samples as "top" vs "rest" by an upper-tail TF-rate quantile.
# Ties at the threshold fall into "rest" (strict > cutoff for "top").
assign_top <- function(df, tf_var, top_frac) {
  cutoff <- quantile(df[[tf_var]], probs = 1 - top_frac, na.rm = TRUE)
  df |> mutate(tf_group = factor(if_else(.data[[tf_var]] > cutoff, "top", "rest"),
                                 levels = c("rest", "top")))
}

# Per-motif Wilcoxon rank-sum (top vs rest) on the raw singleton_dist value.
# NAs are dropped per motif. BH/FDR correction is applied across the 9 motifs
# within this single dataset x threshold block.
test_motifs <- function(df, top_frac) {
  map_dfr(motifs, function(m) {
    col <- sd_col(m)
    sub <- df |> filter(!is.na(.data[[col]])) |> select(value = all_of(col), tf_group)
    top  <- sub$value[sub$tf_group == "top"]
    rest <- sub$value[sub$tf_group == "rest"]
    wt <- wilcox.test(top, rest)
    tibble(
      motif       = m,
      n_top       = length(top),
      n_rest      = length(rest),
      median_top  = median(top),
      median_rest = median(rest),
      median_diff = median(top) - median(rest),
      W           = unname(wt$statistic),
      p           = wt$p.value
    )
  }) |>
    mutate(p.adj = p.adjust(p, method = "BH"),
           signif = cut(p.adj,
                        breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                        labels = c("***", "**", "*", "ns"))) |>
    relocate(p.adj, signif, .after = p)
}

fmt_table <- function(tbl) {
  tbl |>
    mutate(across(c(median_top, median_rest, median_diff), \(x) round(x, 3)),
           p     = signif(p, 3),
           p.adj = signif(p.adj, 3)) |>
    kable()
}

# Long-format frame for faceted boxplots.
boxplot_df <- function(df) {
  df |>
    select(tf_group, all_of(sd_col(motifs))) |>
    pivot_longer(-tf_group, names_to = "motif", values_to = "value") |>
    mutate(motif = str_remove(motif, "_singleton_dist")) |>
    filter(!is.na(value))
}

plot_motifs <- function(df, title) {
  ggplot(boxplot_df(df), aes(tf_group, value, fill = tf_group)) +
    geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3, width = 0.6) +
    stat_compare_means(method = "wilcox.test", label = "p.signif",
                       label.x = 1.5, size = 3.5, hide.ns = FALSE) +
    facet_wrap(~ motif, scales = "free_y", nrow = 3) +
    scale_fill_manual(values = group_colors) +
    labs(x = NULL, y = "singleton_dist", title = title) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none")
}
```

This analysis runs the **same comparison independently** in the primary
(PCAWG) and metastatic cohorts: for each of the 9 telomere variant
repeat (TVR) motifs, does the per-sample `singleton_dist` value differ
between the highest-TF-rate samples and the rest? The two datasets are
reported side by side so any shared pattern can be judged by eye — there
is **no automated cross-dataset test**.

-   **Test:** Wilcoxon rank-sum (Mann–Whitney), `top` vs `rest`, on the
    raw `singleton_dist` value.
-   **Top group:** highest 5% and highest 10% of tumour TF rate
    (`tf_primary_rate` in primary, `tumor_tf_rate` in metastatic),
    reported separately.
-   **Multiple testing:** Benjamini–Hochberg FDR across the 9 motifs,
    applied within each dataset × threshold block. Each dataset is
    corrected on its own.
-   **Direction:** `median_diff = median(top) − median(rest)`.

```{r assign-groups}
prim_5 <- assign_top(primary,    "tf_primary_rate", 0.05)
prim_10 <- assign_top(primary,    "tf_primary_rate", 0.10)
meta_5 <- assign_top(metastatic, "tumor_tf_rate",   0.05)
meta_10 <- assign_top(metastatic, "tumor_tf_rate",   0.10)
```

# Primary (PCAWG)

## Top 5% TF rate

```{r prim-5-tbl}
res_prim_5 <- test_motifs(prim_5, 0.05)
fmt_table(res_prim_5)
```

```{r prim-5-plot, fig.width = 10, fig.height = 8}
plot_motifs(prim_5, "Primary — singleton_dist, top 5% TF vs rest")
```

## Top 10% TF rate

```{r prim-10-tbl}
res_prim_10 <- test_motifs(prim_10, 0.10)
fmt_table(res_prim_10)
```

```{r prim-10-plot, fig.width = 10, fig.height = 8}
plot_motifs(prim_10, "Primary — singleton_dist, top 10% TF vs rest")
```

# Metastatic

## Top 5% TF rate

```{r meta-5-tbl}
res_meta_5 <- test_motifs(meta_5, 0.05)
fmt_table(res_meta_5)
```

```{r meta-5-plot, fig.width = 10, fig.height = 8}
plot_motifs(meta_5, "Metastatic — singleton_dist, top 5% TF vs rest")
```

## Top 10% TF rate

```{r meta-10-tbl}
res_meta_10 <- test_motifs(meta_10, 0.10)
fmt_table(res_meta_10)
```

```{r meta-10-plot, fig.width = 10, fig.height = 8}
plot_motifs(meta_10, "Metastatic — singleton_dist, top 10% TF vs rest")
```

# Side-by-side comparison

Primary and metastatic results aligned per motif, for visual comparison
only. `diff` columns are `median(top) − median(rest)`; `p.adj` are the
BH-corrected p-values within each cohort.

## Top 5%

```{r compare-5}
left_join(
  res_prim_5 |> select(motif, prim_diff = median_diff, prim_p.adj = p.adj, prim_signif = signif),
  res_meta_5 |> select(motif, meta_diff = median_diff, meta_p.adj = p.adj, meta_signif = signif),
  by = "motif"
) |>
  mutate(across(c(prim_diff, meta_diff), \(x) round(x, 3)),
         across(c(prim_p.adj, meta_p.adj), \(x) signif(x, 3))) |>
  kable()
```

## Top 10%

```{r compare-10}
left_join(
  res_prim_10 |> select(motif, prim_diff = median_diff, prim_p.adj = p.adj, prim_signif = signif),
  res_meta_10 |> select(motif, meta_diff = median_diff, meta_p.adj = p.adj, meta_signif = signif),
  by = "motif"
) |>
  mutate(across(c(prim_diff, meta_diff), \(x) round(x, 3)),
         across(c(prim_p.adj, meta_p.adj), \(x) signif(x, 3))) |>
  kable()
```

# Group sizes

Realized `top` / `rest` counts per dataset and threshold (per-motif test
counts may be lower where `singleton_dist` is missing).

```{r group-sizes}
tibble(
  dataset   = c("primary", "primary", "metastatic", "metastatic"),
  threshold = c("top 5%", "top 10%", "top 5%", "top 10%"),
  n_top  = c(sum(prim_5$tf_group == "top"),  sum(prim_10$tf_group == "top"),
             sum(meta_5$tf_group == "top"),  sum(meta_10$tf_group == "top")),
  n_rest = c(sum(prim_5$tf_group == "rest"), sum(prim_10$tf_group == "rest"),
             sum(meta_5$tf_group == "rest"), sum(meta_10$tf_group == "rest"))
) |> kable()
```
